---
title: "Recap and getting data into R"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
```

## Where you are now

The previous six chapters introduced R from the ground up. You have not yet analyzed any real data, and that is intentional: most of the difficulty new users encounter with R is not statistical but environmental. People struggle not because regression is hard but because they cannot tell the difference between an R script and an R project, or because they do not know that an R session has been silently inheriting state from yesterday. By the end of Section 1, those obstacles are behind you.

This chapter has three purposes. The first is to recap the six earlier chapters in compact form. The second is to introduce one practical step that Section 1 has not yet covered: how to load data from a file into R. The third is to point you toward what comes next. The questions at the end consolidate concepts across all six earlier chapters and the new material on data import.

## What you have learned

**Chapter 1** introduced R and the case for working with code. R is a free, open-source programming language designed for statistics. The reasons to use it rather than SPSS or Excel are reproducibility (the script is a complete record of what you did), error detection (every step is inspectable), and scaling (operations defined once can be applied to many variables). R is "object-based" in the sense that almost everything you work with is a named object, created by you, that can be passed into functions to produce new objects.

**Chapter 2** distinguished R the engine from RStudio the dashboard. R performs the computations; RStudio provides the working environment around it. You install both, and the standard environment for new users is RStudio Desktop (Open Source Edition). Alternatives exist (Positron, VS Code with the R extension, Jupyter), but RStudio remains the easiest entry point.

**Chapter 3** mapped out the relevant kinds of files and one in-memory state. The **environment** holds your objects while R is running and disappears when R closes. An **R script** (`.R`) is a saved record of code. An **RData file** (`.RData`) is a saved snapshot of objects. An **R project** (`.Rproj`) is a folder formally designated as a project, which sets the working directory and groups related files. **R Markdown** (`.Rmd`) files mix prose and code to produce finished documents. The script is the authoritative record; everything else is for convenience or for reporting.

**Chapter 4** examined R sessions and how they relate to projects. A session is one running instance of R; opening a project starts a session, closing or restarting the session destroys the in-memory state. The recommended habit is to restart the session frequently and re-run the script from the top, so that no object exists in the environment for reasons other than your script. The corresponding RStudio settings ("Restore .RData into workspace at startup" and "Save workspace to .RData on exit") should be turned off, because they silently preserve state across sessions in a way that obscures bugs.

**Chapter 5** introduced the main object types: single values (numbers, strings, logicals), vectors (ordered sequences of values of the same type), and data frames (rows are observations, columns are variables, each column a vector). It introduced the indexing rules, including 1-based indexing, the `$` for column extraction, and the `[rows, columns]` two-dimensional bracket form. It also introduced functions: pieces of R that take objects as input and return objects as output, with a small toolkit (`class`, `length`, `str`, `head`, `summary`, `nrow`, `ncol`, `colnames`) that you will use to inspect almost any unfamiliar object.

**Chapter 6** covered the basic operations of arithmetic and summary statistics, gave a first illustration of writing a custom function (with `if` and `else`), and worked through a substantive example in which a survey dataset with many distorted variables was corrected three ways: manually, with a function applied repeatedly, and with a function applied to all variables in one line. The third approach is the one that pays off in practice. The chapter ended with packages: how to install them, how to load them, how to find documentation, with `psych::describe()` and `psych::describeBy()` as concrete examples, and the `::` notation as an alternative to loading.

## What this means for the rest of the book

You now have everything you need to start using R for actual data analysis. The remaining sections of the book will introduce specific kinds of analysis and operations relevant to the [**4 other pillars**]{style="color: red;"} ([**data visualization**]{style="color: orange;"}, [**statistical modeling**]{style="color: orange;"}, [**statistical inference**]{style="color: orange;"}, [**causal inference**]{style="color: orange;"}) in combination with R coding (i.e., the [**5th pillar**]{style="color: red;"}), and the corresponding R workflow for each. Where Section 1 was about setting up the environment, the remaining sections will be about doing things with it.

Two pieces of advice for the transition.

The first is that you should expect to consult the help system constantly. Even researchers with decades of R experience read help pages routinely. The skill that distinguishes someone fluent in R is not memorizing every function but knowing how to find the right one.

The second is that the workflow you have learned (project, restart, script, save explicitly when needed) is more important than any specific analysis you will run later. Statistical operations come and go; methods evolve; specific R functions are sometimes superseded by better ones. The discipline of organizing a project, restarting cleanly, and writing a script that reproduces an analysis from start to finish will serve you regardless of which analyses you eventually run.

## A final missing piece: getting data into R

There is one practical step we have not yet covered. You know how to create objects from scratch, but in actual research projects your data does not arrive in your R script as hand-typed vectors; it arrives as a file on your computer, exported from a survey platform, an experimental software, a spreadsheet a collaborator sent you, or a database. Before you can analyze it, you have to load it into R.

This section covers the most common cases: comma-separated values (CSV) files, Excel files, and the point-and-click import feature in RStudio. We then return to the same comparison the rest of Section 1 has been building toward: code-based imports versus click-based imports, and why the code-based approach is the one to make your default.

For all the examples below, we assume you are working inside an R project (Chapter 4), so that the working directory is set to the project folder and you can refer to data files by paths relative to that folder. A common convention is to place data files in a subfolder called `data` inside the project, so that paths look like `"data/survey_2024.csv"`.

### Reading CSV files

A CSV file is a plain-text file in which rows of a table are stored as lines, and columns are separated by commas. CSV is the most common format for sharing tabular data because it is human-readable, software-agnostic, and supported by every statistical tool.

Base R has the function `read.csv()` for this:

```{r}
#| eval: false
survey <- read.csv("data/survey_2024.csv")
```

After this line, `survey` is a data frame with the contents of the file. You can inspect it with `head(survey)`, `str(survey)`, or `summary(survey)` as introduced in Chapter 5.

Three practical points about CSV files.

The first point is the **separator**. The "C" in CSV stands for "comma," but in many European locales (including Austria, Germany, France) the convention is to use a semicolon instead, because the comma is already used as the decimal separator. If you open a CSV with `read.csv()` and the entire file ends up in a single column, the separator is probably a semicolon. Use `read.csv2()` instead, which assumes semicolon separators and comma decimals:

```{r}
#| eval: false
survey <- read.csv2("data/survey_2024.csv")
```

Or pass the separator explicitly to `read.csv()`:

```{r}
#| eval: false
survey <- read.csv("data/survey_2024.csv", sep = ";", dec = ",")
```

The second point is the **header row**. By default, `read.csv()` assumes the first line of the file contains the column names. If your file has no header row, set `header = FALSE`.

The third point is **alternatives in packages**. The `readr` package (part of the tidyverse) provides `read_csv()`, which is faster on large files and gives clearer messages about how each column was interpreted. It returns a slightly different object (a "tibble," which is a data frame with a few cosmetic differences). For Section 1 the base R `read.csv()` is sufficient.

### Reading Excel files

Excel files have the extension `.xlsx` (modern Excel) or `.xls` (older Excel). Base R does not include a function to read them, so you need a package. The standard choice is `readxl`:

```{r}
#| eval: false
install.packages("readxl")  # only once, per computer
library(readxl)

survey <- read_excel("data/survey_2024.xlsx")
```

`read_excel()` handles both `.xlsx` and `.xls`. By default it reads the first sheet of the workbook. To read a specific sheet:

```{r}
#| eval: false
survey <- read_excel("data/survey_2024.xlsx", sheet = "responses")
```

The argument `sheet` accepts either the sheet name as a string or its position as a number (`sheet = 2` for the second sheet).

Two notes worth flagging.

First, Excel files often contain things that do not translate well to a data frame: merged cells, formatting, formulas, multiple tables on one sheet, headers spanning several rows. If your Excel file is structured cleanly (one table per sheet, one header row, one variable per column), `read_excel()` works without further argument. If it is not, you may need to use `range = "A3:F50"` or similar to specify which cells to read. When you have control over the file, exporting it as a CSV first usually saves time.

Second, `readxl` reads from Excel files without needing Excel installed. The package parses the file format directly. This is a real advantage on machines that do not have Microsoft Office.

### Point-and-click import in RStudio

RStudio includes a graphical import feature that handles both CSV and Excel files. To use it, click in the **Files** panel on the file you want to import, or go to **File > Import Dataset** in the menu and choose "From Text (base)," "From Text (readr)," or "From Excel."

A dialog opens. It shows a preview of the data, lets you adjust options (delimiter, decimal separator, sheet for Excel files, header row, column types), and previews the result.

This is genuinely useful for one important reason: at the bottom of the dialog there is a **Code Preview** box that shows the R code corresponding to your current settings. Once you have the preview looking right, you can copy that code and paste it into your script.

If you simply click "Import," the code runs once and the data appears, but the code is not added to your script. You will not be able to reproduce the import without going through the dialog again. The good practice is to use the dialog as a way to *generate* the code, copy it into your script, and then close the dialog.

### Why coding the import is better

The reason to put the import in your script is the same reason to put any other step in your script. We have made this argument repeatedly in Section 1; in the context of data import it has four specific consequences.

The first is **reproducibility**. The import is the very first step of any analysis. If the import is in your script, then re-running the script from a clean session loads the data automatically; you do not need to remember which menu options you chose. If the import is not in your script, then your analysis is not reproducible: a colleague running your script will get an error on the very first line, because their environment does not contain the data.

The second is **transparency**. Reading `read.csv("data/survey_2024.csv")` tells the reader exactly which file is being loaded, with which separator, with which decimal convention. A click-through history tells nobody anything.

The third is **portability**. A scripted import using a relative path (`"data/survey_2024.csv"`, not `"C:/Users/yourname/Downloads/.../survey_2024.csv"`) works on any computer that has the project folder. You can email the project folder to a collaborator, or push it to a repository, and the same script will run.

The fourth is **error correction**. If the file's structure changes (a new column is added, a separator changes, the file is renamed), you adjust one line of code. With a click-through workflow, you redo the entire dialog walkthrough, hoping to remember every setting you previously chose.

Use the import dialog as a *helper* that produces the code, then keep the code. The dialog is a good way to discover the right arguments for an unfamiliar file format; it is not a substitute for the script.

### A practical pattern

A typical first script in a project loads the data, inspects it, and assigns it to a clean object. For example:

```{r}
#| eval: false
library(readxl)

# Load the raw data
survey <- read_excel("data/survey_2024.xlsx", sheet = "responses")

# Inspect
str(survey)
head(survey)
summary(survey)
```

You now have everything you need to begin analyzing real data, and you have a script that any future reader (yourself, a collaborator, a reviewer) can run from start to finish.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  answer = "R is the engine that performs computations; RStudio is a dashboard providing a working environment around it.",
  "RStudio includes its own copy of R, so installing R separately is unnecessary.",
  "R and RStudio are the same program, just with different names.",
  "RStudio is a paid add-on that performs the statistics, while R only provides the user interface."
)

opts_q2 <- c(
  "An R project, because it stores objects on disk.",
  answer = "An R script, because it is the complete written record of every step of the analysis.",
  "An RData file, because it preserves the entire environment.",
  "The Environment panel in RStudio."
)

opts_q3 <- c(
  "It saves the environment to disk in case of a crash.",
  "It reinstalls all loaded packages from CRAN.",
  answer = "It clears the environment and unloads packages, ensuring that running the script from the top depends only on the script itself, not on accumulated session state.",
  "It is a slow operation that should only be performed at the end of the day."
)

opts_q4 <- c(
  "Once: the package is permanently active on your computer.",
  answer = "Every R session in which you want to use the package's functions.",
  "Never: `install.packages()` already loads the package.",
  "Only when the package's authors release a new version."
)

opts_q5 <- c(
  "`survey[age]`",
  answer = "`survey$age`",
  "`survey.age`",
  "`survey->age`"
)

opts_q6 <- c(
  "Position 0: R indexes from zero.",
  answer = "Position 1: R indexes from one.",
  "Position -1: R indexes from the end backwards.",
  "It depends on whether the vector is numeric or character."
)

opts_q7 <- c(
  "The mean of the entire vector regardless of NA.",
  answer = "`NA`, because at least one value is missing; you would need `na.rm = TRUE` to compute the mean of the non-missing values.",
  "An error message that stops execution.",
  "Zero, by convention."
)

opts_q8 <- c(
  "Faster computation.",
  "Automatic generation of APA-formatted output.",
  answer = "It scales to large numbers of variables, makes the operation defined in one place easy to correct, and keeps the intent of the code transparent.",
  "It produces graphical menus for the analysis."
)

opts_q9 <- c(
  "`load.csv()`",
  "`import.csv()`",
  answer = "`read.csv()`",
  "`open.csv()`"
)

opts_q10 <- c(
  "Use `read.csv()` with no extra arguments and trust R to detect the format.",
  answer = "Use `read.csv2()`, or pass `sep = \";\"` and `dec = \",\"` to `read.csv()`.",
  "Convert the file to JSON before importing.",
  "Open the file in Excel and re-save it; nothing else can be done."
)

opts_q11 <- c(
  "It is faster than coding the import.",
  "It is the only way to read Excel files in R.",
  answer = "Clicking 'Import' loads the data once, but the code is not added to your script, so the import is not reproducible. The good practice is to use the dialog to generate code, then copy that code into your script.",
  "The dialog cannot read CSV files, only Excel files."
)
```

**1. Which best describes the relationship between R and RStudio?**

`r longmcq(opts_q1)`

**2. Which file type is the authoritative record of an analysis?**

`r longmcq(opts_q2)`

**3. Why is restarting the R session a recommended habit?**

`r longmcq(opts_q3)`

**4. How often do you need to call `library(packagename)`?**

`r longmcq(opts_q4)`

**5. Which expression extracts the column `age` from a data frame called `survey`?**

`r longmcq(opts_q5)`

**6. At what position is the first element of a vector in R?**

`r longmcq(opts_q6)`

**7. If `x` is a numeric vector containing one `NA`, what does `mean(x)` return?**

`r longmcq(opts_q7)`

**8. What is the main practical benefit of writing a custom function and applying it to many variables at once, rather than performing the operation manually on each variable?**

`r longmcq(opts_q8)`

**9. Which base R function reads a comma-separated values (CSV) file into a data frame?**

`r longmcq(opts_q9)`

**10. You receive a CSV file from a European colleague. When you open it with `read.csv()`, the entire file ends up in a single column. What is the most likely cause and fix?**

`r longmcq(opts_q10)`

**11. What is the recommended way to use RStudio's "Import Dataset" dialog?**

`r longmcq(opts_q11)`
