---
title: "From the mean to the line: the simple linear model"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
theme_set(theme_minimal(base_size = 12))
data <- read.csv("data/vgames_study.csv")
```

## Overview

Chapter 14 fitted two models to a small dataset: the **grand-mean model**, which makes the same prediction for every observation, and the **group-means model**, which makes a different prediction for each experimental condition. The two-group model fit better. The reason: it used more information about each participant.

This chapter takes the next step. What if the information we use to make a participant's prediction is not their group, but the value of a *continuous* variable measured on them? For example: what if we predict each participant's aggression score from how many hours per week they play video games?

A model that uses one continuous predictor needs a new mathematical form. We will replace the flat horizontal line (the grand mean) with a sloped line, and we will introduce the two numbers that define such a line. This chapter is conceptual: by the end you will understand what a simple linear model is, what its parameters mean, what counts as a prediction, and what criterion R uses to choose "the best" line. Chapter 16 will then walk through how to fit and interpret the model in R using `lm()`.

## A new question, a new picture

The motivating question for this chapter is one many researchers have asked: **do people who play more hours of video games per week tend to report higher aggression?**

We have two variables in `vgames_study`:

- **The predictor** (which we will call $X$ from here on): `hours_weekly`, how many hours of gaming per week.
- **The outcome** (which we will call $Y$): `aggression`, the post-experiment aggression score on a 1-10 scale.

The natural picture for one continuous predictor and one continuous outcome is a **scatterplot**, which you met in Chapter 10. Each participant is one point, positioned by their $X$ value horizontally and their $Y$ value vertically.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 6.5
#| fig.height: 4
ggplot(data, aes(x = hours_weekly, y = aggression)) +
  geom_point(alpha = 0.55, color = "#1F2937") +
  labs(x = "Weekly playing hours",
       y = "Aggression (1-10)")
```

There is a vague upward trend: participants who play more hours tend to score slightly higher on aggression, on average. But the trend is not tight; there is substantial scatter. Plenty of heavy players have low aggression scores; plenty of light players have high ones. The relationship is real but weak, and the eye alone is not good at quantifying it.

In Chapter 14, our model for `aggression` was a single number (the grand mean, around 4.89). On this scatterplot, that model is a flat horizontal line:

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 6.5
#| fig.height: 4
grand_mean <- mean(data$aggression)
ggplot(data, aes(x = hours_weekly, y = aggression)) +
  geom_point(alpha = 0.55, color = "#1F2937") +
  geom_hline(yintercept = grand_mean, color = "#EF4444", linewidth = 0.8) +
  annotate("text", x = max(data$hours_weekly), y = grand_mean + 0.2,
           label = paste0("Grand mean = ", round(grand_mean, 2)),
           hjust = 1, color = "#EF4444", fontface = "bold") +
  labs(x = "Weekly playing hours",
       y = "Aggression (1-10)")
```

This is the mean-only model viewed in two dimensions. The line is flat because the prediction does not depend on `hours_weekly`: every participant is predicted to have aggression 4.89, regardless of how many hours they play. The residual of each participant is the vertical distance from their point to this red line, which is exactly the residual we computed in Chapter 14.

What if we let the prediction depend on `hours_weekly`?

## The line as a two-number model

The simplest way to let the prediction vary with $X$ is to tilt the line. Instead of a flat horizontal line at one fixed value, we draw a line that has both a height (where it sits on the y-axis) and a tilt (how much it rises or falls per unit of $X$).

This is the equation of a line, which you may recognize from high school mathematics:

$$
Y = b_0 + b_1 X.
$$

Two numbers define the line:

- **The intercept**, $b_0$, is the value of $Y$ when $X$ is zero. It is the height at which the line meets the y-axis.
- **The slope**, $b_1$, is the amount $Y$ changes for each one-unit increase in $X$. It is the line's tilt.

For real data, the line will not pass exactly through every point; the points scatter around it. The full **simple linear model** acknowledges this by adding a residual term, one residual per observation:

$$
Y_i = b_0 + b_1 X_i + \varepsilon_i.
$$

The subscript $i$ indexes each participant. $Y_i$ is participant $i$'s observed outcome; $X_i$ is participant $i$'s predictor value; $b_0 + b_1 X_i$ is the model's prediction for that participant (sometimes written $\hat{Y}_i$); and $\varepsilon_i$ is participant $i$'s residual, the vertical distance between the point and the line.

When $b_1$ is zero, the line is flat and the model reduces to the mean-only model from Chapter 14: a single value ($b_0$) predicted for everyone, regardless of $X$. The simple linear model contains the mean-only model as a special case.

## What the slope means

The slope $b_1$ is the most important number in the model. It captures both the **direction** and the **size** of the relationship between $X$ and $Y$.

- **Positive slope** ($b_1 > 0$): as $X$ increases, $Y$ increases. The line goes up to the right. In our example: more weekly hours of gaming, higher predicted aggression.
- **Negative slope** ($b_1 < 0$): as $X$ increases, $Y$ decreases. The line goes down to the right.
- **Slope of zero** ($b_1 = 0$): no relationship; the line is flat (this is the mean-only model).

The size of the slope tells you how strong the predicted change in $Y$ is for a one-unit increase in $X$. For our running example, the best-fitting line has $b_1 \approx 0.08$. The interpretation in plain English:

> Each additional hour of weekly gaming is associated with an increase of 0.08 points in aggression.

A few practical points about this interpretation.

The slope is **per one-unit change in $X$**. If you want to know what the model predicts for a *ten*-hour difference, multiply: $0.08 \times 10 = 0.8$ points. The model says ten more hours of gaming is associated with about eight tenths of a point higher aggression.

The slope is an **average over the dataset**. It is what the model predicts; it does not mean every participant who plays one more hour scores 0.08 more on aggression. Individual participants vary widely around the line.

The slope is **a description, not a cause**. Notice the wording: "associated with." We have a statistical relationship between two variables in a sample. We have not (yet) said that gaming *causes* aggression. Causal inference is a separate question, with its own logic, and we will return to it in later sections of the book.

## What the intercept means

The intercept $b_0$ is the predicted value of $Y$ when $X = 0$. In our example, the best-fitting line has $b_0 \approx 4.20$, so a participant who plays zero hours per week is predicted to have an aggression score of 4.20.

The intercept is sometimes a substantively meaningful number, and sometimes it is not. In our case, $X = 0$ (zero hours of gaming) is a value participants could plausibly take, so the intercept has a real interpretation: it is the model's prediction for non-gamers. In other studies, $X = 0$ is outside the range of the data (for example, a study measuring height in adults: $X = 0$ would mean a height of zero, which is not a meaningful baseline). In those cases the intercept is still mathematically necessary (the line has to start somewhere) but should not be interpreted as a real prediction. We will see techniques for making the intercept meaningful (such as centering predictors) in later chapters.

## A prediction for any value of $X$

Once you have $b_0$ and $b_1$, the line gives a prediction for any value of $X$ you care to plug in. You simply evaluate the equation:

$$
\hat{Y} = b_0 + b_1 X.
$$

In our example, $\hat{Y} = 4.20 + 0.08 X$:

- For $X = 0$ hours: $\hat{Y} = 4.20 + 0.08 \times 0 = 4.20$.
- For $X = 5$ hours: $\hat{Y} = 4.20 + 0.08 \times 5 = 4.60$.
- For $X = 10$ hours: $\hat{Y} = 4.20 + 0.08 \times 10 = 5.00$.
- For $X = 20$ hours: $\hat{Y} = 4.20 + 0.08 \times 20 = 5.80$.

Every point along the line is a prediction. Unlike the mean-only model (one prediction for everyone), the simple linear model produces a *different* prediction for each value of the predictor.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 6.5
#| fig.height: 4
fit <- lm(aggression ~ hours_weekly, data = data)
b0 <- coef(fit)[1]
b1 <- coef(fit)[2]
data_pred <- data.frame(hours_weekly = seq(min(data$hours_weekly), max(data$hours_weekly), length.out = 200))
data_pred$pred <- b0 + b1 * data_pred$hours_weekly
ggplot(data, aes(x = hours_weekly, y = aggression)) +
  geom_point(alpha = 0.55, color = "#1F2937") +
  geom_line(data = data_pred, aes(x = hours_weekly, y = pred),
            color = "#3B82F6", linewidth = 0.9) +
  annotate("text", x = max(data$hours_weekly) - 1, y = max(data_pred$pred) - 0.3,
           label = paste0("Aggression = ", round(b0, 2), " + ", round(b1, 2), " * Hours"),
           hjust = 1, vjust = 1, color = "#3B82F6", fontface = "bold") +
  labs(x = "Weekly playing hours",
       y = "Aggression (1-10)")
```

The blue line is the simple linear model's prediction. It is not flat: it tilts upward gently. Two participants with different `hours_weekly` values receive different predictions, instead of the same one as under the mean-only model.

One caution. Plugging values of $X$ outside the range of the observed data into the equation is called **extrapolation**, and you should treat any such prediction with caution. Our data go up to about 30 hours of gaming per week. The model will dutifully produce a prediction for $X = 100$, but we have no evidence that the linear relationship continues that far. A line that fits well in the observed range can curve in ways we cannot see beyond it.

## Residuals from the line

For each participant, the residual is exactly the same idea as in Chapter 14: observed minus predicted.

$$
\varepsilon_i = Y_i - \hat{Y}_i = Y_i - (b_0 + b_1 X_i).
$$

Geometrically, the residual is the vertical distance from each point to the line. Below, the residuals are drawn as grey segments connecting each observation to its prediction on the line:

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 6.5
#| fig.height: 4
data$pred <- b0 + b1 * data$hours_weekly
ggplot(data, aes(x = hours_weekly, y = aggression)) +
  geom_segment(aes(xend = hours_weekly, y = pred, yend = aggression),
               color = "#9CA3AF", linewidth = 0.4, alpha = 0.7) +
  geom_point(alpha = 0.6, color = "#1F2937") +
  geom_line(data = data_pred, aes(x = hours_weekly, y = pred),
            color = "#3B82F6", linewidth = 0.9) +
  labs(x = "Weekly playing hours",
       y = "Aggression (1-10)")
```

Participants above the line have positive residuals (they scored higher than the model predicted); participants below the line have negative residuals. Compare this picture to the analogous picture in Chapter 14 (with the grey segments running to the flat grand-mean line). The line tilts in a direction that, on average, makes the grey segments shorter than they were under the mean-only model. That improvement in fit will be quantified by R-squared in the next chapter.

## How is "the best" line chosen?

There are infinitely many candidate lines you could draw through a scatterplot. Some are clearly too steep, some too flat, some pass too low, some too high. We need a precise rule for choosing one of them as "the" line of the model.

The rule is the same one introduced in Chapter 14:

> Choose the line that minimizes the **sum of squared residuals**.

This estimation method is called **Ordinary Least Squares**, abbreviated **OLS**. "Least squares" because it chooses the intercept and slope that make the squared distances from the points to the line as small as possible, in total. "Ordinary" because there are more elaborate versions for special situations (we will not need them in this book), and this is the standard one.

Three candidate lines through the same data illustrate the idea. The first is too flat (it does not tilt enough to capture the upward trend). The second is too steep. The third is the OLS line, where the sum of squared vertical distances from the points to the line is as small as it can be.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 8
#| fig.height: 3.3
x_range <- range(data$hours_weekly)
lines_df <- data.frame(
  hours_weekly = rep(seq(x_range[1], x_range[2], length.out = 100), 3),
  line = rep(c("(a) too flat", "(b) too steep", "(c) OLS line"), each = 100)
)
lines_df$pred <- with(lines_df, {
  ifelse(line == "(a) too flat",  4.7 + 0.01 * hours_weekly,
  ifelse(line == "(b) too steep", 3.2 + 0.20 * hours_weekly,
                                  b0 + b1 * hours_weekly))
})
ggplot(data, aes(x = hours_weekly, y = aggression)) +
  geom_point(alpha = 0.4, color = "#1F2937", size = 1.1) +
  geom_line(data = lines_df, aes(x = hours_weekly, y = pred),
            color = "#3B82F6", linewidth = 0.8) +
  facet_wrap(~ line) +
  labs(x = "Weekly playing hours",
       y = "Aggression (1-10)") +
  theme(strip.text = element_text(face = "bold"))
```

The OLS line is the unique line that makes the sum of squared vertical distances from the points to the line as small as possible. R does this calculation for you automatically: the function `lm()` (introduced in the next chapter) takes a dataset and a formula and returns the OLS intercept and slope. You do not need to know the closed-form formulas for $b_0$ and $b_1$ to use the model; you need to know what they mean and what criterion produced them.

A worthwhile fact about OLS. The mean-only model from Chapter 14 is itself an OLS estimate: the single number that minimizes the sum of squared residuals from a constant prediction is exactly the mean (this is a theorem from a first calculus course). So the framework in this chapter is not a different framework. It is the same one, applied to a slightly more flexible class of predictions.

## Looking ahead

We now have a richer model than the mean: the simple linear model, with two parameters ($b_0$ and $b_1$) and the OLS criterion for choosing them. We have spent this chapter on what the model is, how to read its parameters, and how to think about its predictions and residuals. We have not yet computed any of these quantities in R. That is the topic of the next chapter.

Chapter 16 will:

- introduce the `lm()` function and the formula syntax `Y ~ X`,
- show how to read the `summary()` output (focusing on the coefficients, which are $b_0$ and $b_1$),
- introduce **R-squared** as the formal measure of how much better the line is than the mean,
- show that **correlation** (which you may have heard of) is a special case of simple regression,
- and walk through a small worked example so you can read regression output yourself.

By the end of Chapter 16 you will be fitting and interpreting your first regression models.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "It is the same as the mean-only model from Chapter 14.",
  "It uses a different prediction for each observation, ignoring any predictor.",
  answer = "It uses a continuous predictor $X$ to make a different prediction for each observation; the predictions form a line with two parameters: an intercept ($b_0$) and a slope ($b_1$).",
  "It guarantees a perfect fit to the data."
)

opts_q2 <- c(
  "The predicted value of $X$ when $Y = 0$.",
  answer = "The predicted value of $Y$ when $X = 0$; the height at which the line crosses the y-axis.",
  "The average of all $Y$ values.",
  "The slope of the line."
)

opts_q3 <- c(
  "How much $X$ changes for a one-unit increase in $Y$.",
  answer = "How much $Y$ changes for a one-unit increase in $X$. A positive slope means $Y$ increases as $X$ increases; a negative slope means $Y$ decreases as $X$ increases.",
  "The correlation between $X$ and $Y$.",
  "The intercept of the line."
)

opts_q4 <- c(
  "It produces an exact, deterministic, fully accurate prediction.",
  answer = "It produces a prediction $\\hat{Y} = b_0 + b_1 X$; the actual observed $Y_i$ for that participant will typically differ from the prediction by some residual $\\varepsilon_i = Y_i - \\hat{Y}_i$.",
  "It produces the same prediction for every participant.",
  "It does not produce a prediction; only the mean-only model does."
)

opts_q5 <- c(
  "Pick the line that passes through the most data points.",
  "Pick the line that has the steepest slope.",
  answer = "Pick the intercept and slope that minimize the sum of squared residuals (the sum of squared vertical distances from the observations to the line). This is called Ordinary Least Squares (OLS).",
  "Pick the line that has slope = 0."
)

opts_q6 <- c(
  "Slope of zero means the line is vertical.",
  answer = "Slope of zero means the line is flat, the model predicts the same value (the intercept) for every observation, and the simple linear model reduces to the mean-only model from Chapter 14.",
  "Slope of zero is impossible in OLS regression.",
  "Slope of zero means $Y$ is always zero."
)

opts_q7 <- c(
  "It is the safest place to make predictions because it lies outside the data.",
  answer = "It is making predictions for values of the predictor outside the range of the observed data, and such predictions should be treated with caution because we have no evidence the linear relationship continues that far.",
  "It is the same thing as interpolation.",
  "It always produces correct predictions."
)

opts_q8 <- c(
  "Yes: a positive slope and a strong relationship together imply a causal effect.",
  answer = "No: a slope describes a statistical association in a sample. Concluding that $X$ causes $Y$ requires a different argument (causal inference) that the simple linear model itself does not provide.",
  "Yes: the OLS estimator was designed to recover causal effects.",
  "It depends only on the sample size."
)
```

**1. What is a simple linear model?**

`r longmcq(opts_q1)`

**2. What does the intercept ($b_0$) of a simple linear model represent?**

`r longmcq(opts_q2)`

**3. What does the slope ($b_1$) of a simple linear model represent?**

`r longmcq(opts_q3)`

**4. What happens when you plug a value of $X$ into the model equation $\\hat{Y} = b_0 + b_1 X$?**

`r longmcq(opts_q4)`

**5. How does Ordinary Least Squares (OLS) choose the best-fitting line?**

`r longmcq(opts_q5)`

**6. The mean-only model is a special case of the simple linear model. Which value of the slope makes them equivalent?**

`r longmcq(opts_q6)`

**7. What is "extrapolation," and why is it a concern in simple regression?**

`r longmcq(opts_q7)`

**8. Suppose your simple linear model has a slope of $b_1 = 0.5$, meaning each unit increase in $X$ is associated with a 0.5-unit increase in $Y$. Can you conclude from this slope alone that $X$ causes $Y$?**

`r longmcq(opts_q8)`

## References

Navarro, D. J. (2018). *Learning statistics with R: A tutorial for psychology students and other beginners* (Version 0.6), Chapter 15. <https://learningstatisticswithr.com>
