---
title: "Decisions, errors, power, and the SESOI"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
library(pwr)
library(TOSTER)
theme_set(theme_minimal(base_size = 12))
set.seed(2025)
data <- read.csv("data/vgames_study.csv")
```

## Where we are

Chapters 21-23 built the machinery of statistical inference: sampling distributions, standard errors, p-values, test statistics, reference distributions, degrees of freedom. By the end of Chapter 23 you could read every column of `summary(lm(...))` and explain what R was computing when it printed a p-value.

This chapter steps back and looks at the *decision-theoretic* side of the Neyman-Pearson framework. We have been treating the procedure ("reject if p < α") as if it were a mechanical step. Now we ask: what happens when the procedure is wrong? What kinds of errors are there, what are their probabilities, and how should those probabilities trade off against each other? And, just as importantly, what does it actually mean to "have 80% power" for a study? The conventional answer is incomplete in a way that has practical consequences.

By the end of the chapter you will know:

- the two ways a decision procedure can be wrong (Type I and Type II errors), and what controls them,
- what **statistical power** is, and why it is always defined *relative to an effect size*,
- how α, β, n, and effect size lock together in a four-quantity system: pick any three and the fourth is determined,
- what the **Smallest Effect Size of Interest** (the SESOI) is, and why it is the anchor that gives "80% power" a meaning,
- a brief look at **equivalence testing**, which lets you make positive claims about there being no meaningful effect, and
- how to run a power analysis in R with the `pwr` package.

Lakens (2022, Chapter 2) is the canonical reference for this material; we cite him repeatedly throughout.

## The Neyman-Pearson framework, revisited

Recall the structure of the decision procedure from Chapter 22:

1. Before collecting data, specify two hypotheses: a null (H₀) and an alternative (H₁).
2. Choose an α level: the maximum acceptable rate of falsely rejecting H₀ when it is in fact true.
3. Decide on a sample size that gives adequate **power** to detect an effect you care about.
4. Collect the data, run the test, and reject H₀ if p < α; otherwise do not reject.

The justification for treating this as a sensible procedure is that, *across the many studies you run over your career using α = .05 when H₀ is true*, about 5% of them will give a "reject" decision purely by chance. That is the calibrated false-positive rate. The procedure does not promise that any individual decision is correct; it promises that the long-run frequency of certain types of mistake is controlled.

Two kinds of mistake can occur each time you run the procedure.

## The two ways the procedure can be wrong

A 2×2 table summarizes the situation cleanly. There are two states of the world (the null is actually true, or actually false) and two decisions you can make (reject H₀, or do not reject H₀). Two of the four cells are correct decisions; two are errors.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 8
#| fig.height: 3.6
tbl <- expand.grid(
  Truth = factor(c("H0 is true", "H0 is false"), levels = c("H0 is true", "H0 is false")),
  Decision = factor(c("Reject H0", "Do not reject H0"),
                    levels = c("Reject H0", "Do not reject H0"))
)
tbl$Outcome <- c("Type I error\n(false positive)",
                 "Correct non-rejection\n(true negative)",
                 "Correct rejection\n(power = 1 - beta)",
                 "Type II error\n(false negative)")
tbl$Prob <- c("Rate = alpha", "Rate = 1 - alpha",
              "Rate = 1 - beta", "Rate = beta")
tbl$IsError <- c(TRUE, FALSE, FALSE, TRUE)

ggplot(tbl, aes(x = Truth, y = Decision)) +
  geom_tile(aes(fill = IsError), color = "white", linewidth = 1.5) +
  geom_text(aes(label = paste(Outcome, "\n", Prob)), size = 4.2, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("FALSE" = "#10B981", "TRUE" = "#EF4444"),
                    guide = "none") +
  labs(title = "The four possible outcomes of a hypothesis test",
       x = NULL, y = NULL) +
  theme(panel.grid = element_blank(),
        axis.text = element_text(face = "bold", size = 12))
```

**Type I error (α):** the procedure rejects H₀ when H₀ is actually true. A "false positive": we claim there is an effect, but there is not. Across all the studies in your career where H₀ is true, the procedure produces a Type I error in α of them (about 5% if α = .05).

**Type II error (β):** the procedure fails to reject H₀ when H₀ is actually false. A "false negative": there is an effect, but the test does not detect it. Across all the studies where H₀ is false with a particular alternative effect, the procedure misses the effect in β of them.

**Correct rejection (1 − β = power):** the procedure rejects H₀ when H₀ is actually false. This is what we want: we catch the real effect. Power is the long-run rate of correct rejections, given a particular alternative.

**Correct non-rejection (1 − α):** the procedure does not reject H₀ when H₀ is true. Also a desirable outcome; this happens 95% of the time at α = .05 when there is no real effect.

Note carefully: Type I and Type II error rates are *long-run frequencies of the procedure*, not probabilities about any specific study. The procedure produces a Type I error in α of all the studies where the null is actually true. It does not say "this study has a 5% chance of being a false positive"; for any individual study, the truth is either H₀ or H₁, and you either committed an error or you did not. The probability is about the procedure across uses, not about the single result.

## Power, conditional on an effect size

**Statistical power** is the probability of correctly rejecting H₀ when H₀ is actually false: 1 − β. It is the "ability to detect a real effect." Most people who have heard of statistical methods have heard of power, often as the slogan "you should design studies with at least 80% power."

What is often missing from that slogan is a critical condition. Power is not an absolute property of a procedure. Power depends on **how false** H₀ is. Detecting a huge effect requires less power than detecting a tiny effect. A study can have 99% power to detect a giant effect and 5% power to detect a small one. The full definition is:

> Power = the probability of correctly rejecting H₀, **given that the true effect size is some specific value δ.**

When researchers say "my study has 80% power," they always mean *for a specific effect size that they cared about*. Without that anchor, the claim is undefined. We will return to this point in detail when we get to the SESOI.

A picture helps. Imagine the sampling distribution of the test statistic under H₀ (centered at 0) and the sampling distribution under H₁ (centered at the true effect size, in test-statistic units). The decision rule is "reject if the test statistic falls beyond the critical value." The probabilities of Type I error and Type II error correspond to specific areas under those two curves.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 8.5
#| fig.height: 4.4
xs <- seq(-4, 7, length.out = 800)
null_d <- dnorm(xs, mean = 0, sd = 1)
alt_d  <- dnorm(xs, mean = 2.5, sd = 1)
crit <- 1.96
df_plot <- bind_rows(
  data.frame(x = xs, y = null_d, curve = "H0 (true effect = 0)"),
  data.frame(x = xs, y = alt_d,  curve = "H1 (true effect > 0)")
)
df_typeI <- data.frame(x = xs[xs >= crit], y = null_d[xs >= crit])
df_typeII <- data.frame(x = xs[xs < crit], y = alt_d[xs < crit])
df_power  <- data.frame(x = xs[xs >= crit], y = alt_d[xs >= crit])
ggplot(df_plot, aes(x, y, color = curve)) +
  geom_ribbon(data = df_typeII, aes(x = x, ymin = 0, ymax = y),
              fill = "#EF4444", alpha = 0.55, inherit.aes = FALSE) +
  geom_ribbon(data = df_typeI, aes(x = x, ymin = 0, ymax = y),
              fill = "#F59E0B", alpha = 0.7, inherit.aes = FALSE) +
  geom_ribbon(data = df_power, aes(x = x, ymin = 0, ymax = y),
              fill = "#10B981", alpha = 0.4, inherit.aes = FALSE) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = crit, color = "#1F2937", linewidth = 0.5, linetype = "dashed") +
  scale_color_manual(values = c("H0 (true effect = 0)" = "#3B82F6",
                                "H1 (true effect > 0)" = "#7C3AED"),
                     name = NULL) +
  # Labels placed far apart vertically and horizontally to avoid overlap
  annotate("label", x = 1.0, y = 0.13, label = "Type II error\n(beta)",
           color = "#EF4444", fontface = "bold", size = 3.6,
           label.size = 0, fill = "white", alpha = 0.85) +
  annotate("label", x = 2.6, y = 0.035, label = "Type I error\n(alpha)",
           color = "#F59E0B", fontface = "bold", size = 3.6,
           label.size = 0, fill = "white", alpha = 0.85) +
  annotate("label", x = 4.0, y = 0.22, label = "Power\n(1 - beta)",
           color = "#10B981", fontface = "bold", size = 3.6,
           label.size = 0, fill = "white", alpha = 0.85) +
  annotate("text", x = crit, y = 0.43, label = "Decision\nthreshold",
           color = "#1F2937", size = 3.3, hjust = -0.05) +
  labs(title = "Type I error, Type II error, and power, visualized",
       subtitle = "Dashed line = decision threshold; colored areas = long-run rates",
       x = "Test statistic", y = "Density") +
  coord_cartesian(ylim = c(0, 0.48)) +
  theme(legend.position = "top")
```

The yellow region is α: the tail of the H₀ distribution beyond the critical value (false rejections when H₀ is true). The red region is β: the part of the H₁ distribution that falls *short* of the critical value (failures to reject when H₁ is true). The green region is power: the part of the H₁ distribution that falls beyond the critical value (correct rejections).

Notice how the two distributions overlap. The closer they are to each other (the smaller the true effect), the more they overlap, the larger β becomes, and the smaller power becomes. The further apart they are (the larger the true effect), the less they overlap, the smaller β, the larger power.

### The four factors that determine power

For a given test, power depends on exactly four things:

1. **Effect size (δ).** Larger true effects are easier to detect. Power increases with effect size.
2. **Sample size (n).** Larger samples produce smaller standard errors, which make the test statistic larger for any given effect. Power increases with n.
3. **Alpha (α).** A stricter alpha (e.g., α = .01 instead of .05) raises the decision threshold and makes the test harder to reject. Power decreases as α decreases.
4. **Variability in the data (σ).** Less variable data make effects easier to detect, because the same effect produces a larger test statistic relative to the noise. Power increases as σ decreases.

These four factors are linked by a fixed mathematical relationship. **For a given test and σ, you cannot freely choose all of α, β, and n.** Pick any two and the third is determined.

The practical consequence: when you design a study, you commit to three of the four (α, power, effect size, n) and let the math give you the fourth. Most often, you commit to α (usually .05), power (often .80 or .90), and an effect size you care about, and you compute the required n. Software like the `pwr` package in R will do this calculation.

## A four-quantity system: pick three, get the fourth

Here is what the four-quantity system looks like in concrete terms. Suppose we want to detect a between-groups difference of d = 0.5 (Cohen's d, a standardized effect size, with σ implicit). We use a two-tailed independent-samples t-test with α = .05.

For different choices of power, the required sample size changes:

```{r}
pwr.t.test(d = 0.5, sig.level = .05, power = .80, type = "two.sample")$n
pwr.t.test(d = 0.5, sig.level = .05, power = .90, type = "two.sample")$n
pwr.t.test(d = 0.5, sig.level = .05, power = .95, type = "two.sample")$n
```

Each row gives the sample size **per group** needed to achieve the listed power. With 80% power, you need 64 participants per group (128 total). For 95% power, you need 105 per group (210 total). Higher power requires more participants. The math is mechanical.

What if we fix the sample size and vary the effect size we hope to detect?

```{r}
# For n = 64 per group at power = .80, what effect size can we detect?
pwr.t.test(n = 64, sig.level = .05, power = .80, type = "two.sample")$d

# For n = 30 per group at power = .80, what effect size can we detect?
pwr.t.test(n = 30, sig.level = .05, power = .80, type = "two.sample")$d
```

With n = 64 per group, we can detect d = 0.50 with 80% power. With n = 30 per group, only d ≈ 0.74. A smaller sample can only detect larger effects.

A picture makes the trade-off vivid. The power curve below shows, for each sample size per group (x-axis), what power we have to detect a true effect of d = 0.5 with α = .05.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 6.5
#| fig.height: 3.5
ns <- seq(5, 200, by = 5)
pwr_at_d05 <- sapply(ns, function(n)
  pwr.t.test(n = n, d = 0.5, sig.level = .05, type = "two.sample")$power)
df_pwr <- data.frame(n = ns, power = pwr_at_d05)
ggplot(df_pwr, aes(x = n, y = power)) +
  geom_line(color = "#3B82F6", linewidth = 0.9) +
  geom_hline(yintercept = c(0.8, 0.95), linetype = "dashed", color = "#EF4444") +
  annotate("text", x = 195, y = 0.78, label = "80% power", color = "#EF4444", hjust = 1) +
  annotate("text", x = 195, y = 0.93, label = "95% power", color = "#EF4444", hjust = 1) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(title = "Power vs sample size per group, for a true effect d = 0.5",
       x = "Sample size per group",
       y = "Power")
```

The curve rises as n grows. To get from 50% power to 80% power requires roughly a doubling of sample size. Power increases with diminishing returns: getting from 80% to 95% costs more than getting from 50% to 80%.

The same logic applies to other test types: t-tests, F-tests, correlations, chi-squared tests, all have analogous power-vs-n curves, which the `pwr` package can produce.

## What is missing from this picture: the SESOI

Everything in the previous section assumed we knew the effect size we wanted to detect. We picked d = 0.5 essentially arbitrarily. But where does that number come from in a real study?

Three options are common in practice:

1. **A conventional value** (Cohen's small d = 0.2, medium d = 0.5, large d = 0.8). The trouble: these are by-eye conventions that have no necessary connection to your specific research question. There is no reason to think the "smallest effect worth detecting" in your study is d = 0.5.
2. **An effect size from a prior study.** The trouble: prior published effect sizes are typically biased upward because of publication bias, optional stopping, and other forms of selective reporting. Building your power analysis on a biased estimate gives you a biased target.
3. **The smallest effect you would actually care about.** This is the right anchor. And it has a name.

The **Smallest Effect Size of Interest** (the SESOI) is the smallest effect size that you, the researcher, would care about. Below the SESOI, you would not change your behavior, your conclusions, or your recommendations, even if the effect were demonstrably real. The SESOI is a value chosen *in advance*, based on:

- **Practical relevance.** What is the smallest effect that would justify acting on the result? For a clinical intervention, the smallest improvement that would change the recommendation. For a brief priming experiment, the smallest priming effect that would change the theoretical interpretation.
- **Theoretical relevance.** What is the smallest effect that the theory would predict? A theory that predicts effects of d ≥ 0.3 implicitly says the SESOI is 0.3.
- **Decision-relevance.** What is the smallest difference that, if real, would change the choice between two practical alternatives?

The SESOI is not the effect you *expect*. It is the effect you would *care about*. The two are different.

### A concrete example

In the Hannah's Bakery example from Chapter 23, suppose Hannah advertises 100-gram scones. Would a true average of 99.5 grams matter? Probably not; that is a half-gram deviation, well below anyone's threshold of complaint. What about 99 grams? Still probably not. What about 95 grams? Yes; that is a 5% shortfall, which most people would consider a meaningful underweight. We might set the SESOI at 5 grams: anything smaller than 5 g of underweight is not worth caring about, anything larger is.

The corresponding design question becomes: how many scones must we weigh to detect a true 5-gram shortfall with 80% power and α = .05? The math gives a specific number, which we can run with `pwr`:

```{r}
# Suppose population SD of scone weights is 8 g. To detect a difference of 5 g:
pwr.t.test(d = 5/8, sig.level = .05, power = .80, type = "one.sample")$n
```

We would need to weigh about 22 scones to have 80% power to detect a true 5-gram shortfall.

### Why "80% power" is meaningless without the SESOI

Now we can be precise about why "I designed the study for 80% power" is incomplete. Power is conditional on an effect size, so the claim is missing the second half of the sentence. Two researchers can both say "my study has 80% power" and mean completely different things.

- Researcher A: 80% power to detect d = 0.1 (a tiny effect; needs n ≈ 1571 per group).
- Researcher B: 80% power to detect d = 1.0 (a huge effect; needs n ≈ 17 per group).

These are extremely different studies with extremely different implications. The first one is set up to catch nearly any meaningful effect, real or theoretical; the second is set up only to catch effects so large they would be obvious in casual inspection. Without naming the effect size, "80% power" is empty.

The remedy is **the SESOI**. State the smallest effect you care about, derive the sample size you need to detect it with adequate power, and report both quantities. Then the reader can evaluate whether your study was actually designed to test the claim it ends up making.

### How the SESOI changes the interpretation of results

The SESOI changes not only the design of the study but also how the results are read.

**Scenario 1: significant but smaller than the SESOI.** Suppose you designed a study with 80% power to detect a SESOI of 0.2, and you find p = .002 with an observed effect of 0.05. The effect is statistically significant. But the observed effect is smaller than the SESOI. The right interpretation: there may be a real effect, but it is too small to matter for the decision you cared about. Statistical significance does not rescue practical irrelevance.

**Scenario 2: non-significant but larger than the SESOI.** Same design. You find p = .15 with an observed effect of 0.4. The effect is not statistically significant. But the observed effect is larger than the SESOI, and you had adequate power. The right interpretation: your data did not provide strong enough evidence to reject the null, but the observed effect is larger than what you would have called meaningful. This is genuinely inconclusive: you cannot confidently call the effect zero (it might just be that the noise was high) and you cannot confidently confirm the predicted effect.

**Scenario 3: non-significant and smaller than the SESOI.** Same design. You find p = .25 with an observed effect of 0.05. The effect is not statistically significant *and* the observed effect is smaller than the SESOI. With adequate power, this is meaningful evidence that the effect, if it exists, is smaller than what you would have called interesting. We can do better than just "fail to reject"; we can construct a test that directly evaluates whether the effect is below the SESOI. That test is equivalence testing, the topic of the next section.

## Equivalence testing and the TOST procedure

We have stressed several times that "p > .05" does not mean "there is no effect." A non-significant standard hypothesis test only tells you that you cannot reject the null; it never confirms the null. But sometimes you *want* to argue, on positive grounds, that an effect is small enough to be practically negligible. A pharmaceutical company wants to claim their cheaper drug works as well as the expensive one. A replicator wants to claim that an effect originally reported as large is in fact too small to matter. A theorist wants to claim that the predicted moderator does not actually moderate. None of these claims can be supported by a non-significant *p*-value in a standard test. They need a different statistical machinery.

That machinery is **equivalence testing**, and the most common version of it is the **Two One-Sided Tests (TOST)** procedure. Lakens (2017, 2022) is the standard modern reference for the technique in psychology and we draw on those sources here.

### The intuition: flipping the question

In a standard hypothesis test, the null is "the effect is exactly zero" and the alternative is "the effect is not zero." If you reject the null, you conclude there is an effect; if you do not reject, you cannot conclude either way. The asymmetry is built in: you can support "effect exists" but not "effect does not exist."

Equivalence testing flips this asymmetry. It starts with a SESOI: a value $\delta$ such that effects smaller than $\delta$ are considered too small to matter. The new null becomes "the effect is at least as large as the SESOI (in either direction)." The new alternative becomes "the effect is smaller than the SESOI." Rejecting the new null means concluding that the effect is too small to matter, which is a positive claim of practical equivalence.

In symbols, suppose we want to argue that the true difference between two means is within $\pm\delta$. We set up two one-sided tests:

- $H_{0,\text{lower}}$: the true difference is $\leq -\delta$. Reject if the observed difference is significantly *greater than* $-\delta$.
- $H_{0,\text{upper}}$: the true difference is $\geq +\delta$. Reject if the observed difference is significantly *less than* $+\delta$.

If we reject **both** one-sided nulls, then we have evidence that the true difference is both above $-\delta$ AND below $+\delta$, that is, within the equivalence band $(-\delta, +\delta)$. This is the TOST procedure.

### A picture of the equivalence band

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 8
#| fig.height: 3.4
sesoi <- 0.5
df_band <- data.frame(
  scenario = factor(c(
    "Equivalence supported (CI inside band)",
    "Inconclusive (CI overlaps band)",
    "Effect detected (CI outside band)"),
    levels = c("Equivalence supported (CI inside band)",
               "Inconclusive (CI overlaps band)",
               "Effect detected (CI outside band)")
  ),
  est = c(0.05, 0.35, 0.8),
  lo  = c(-0.20, -0.05, 0.45),
  hi  = c(0.30, 0.75, 1.15)
)
ggplot(df_band, aes(x = est, y = scenario)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "#10B981", alpha = 0.15) +
  geom_vline(xintercept = c(-sesoi, sesoi), color = "#10B981", linetype = "dashed", linewidth = 0.6) +
  geom_vline(xintercept = 0, color = "#1F2937", linewidth = 0.3) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.15, color = "#1F2937", linewidth = 0.8) +
  geom_point(size = 3, color = "#1F2937") +
  annotate("text", x = -sesoi, y = 3.45, label = "-SESOI", color = "#10B981",
           fontface = "bold", size = 3.5) +
  annotate("text", x = sesoi, y = 3.45, label = "+SESOI", color = "#10B981",
           fontface = "bold", size = 3.5) +
  coord_cartesian(xlim = c(-1.0, 1.4), clip = "off") +
  labs(title = "The equivalence band, and three possible scenarios",
       subtitle = "Green region = effects smaller than the SESOI; dots = observed effect; bars = 90% CI",
       x = "Observed effect (e.g., difference in means)",
       y = NULL) +
  theme(plot.margin = margin(20, 10, 10, 10))
```

The shaded green band shows the region of practically negligible effects: anything between $-$SESOI and $+$SESOI is considered "equivalent to zero" by our pre-specified standard. The three bars are 90% confidence intervals from three hypothetical studies.

- **Top row.** The estimated effect is 0.05; the CI is $[-0.20, 0.30]$, which sits entirely inside the green band. We have evidence of equivalence: the true effect is plausibly within $\pm$SESOI of zero.
- **Middle row.** The estimated effect is 0.35; the CI is $[-0.05, 0.75]$, which crosses both the band and zero. The result is inconclusive; we have neither evidence of equivalence nor evidence against the null.
- **Bottom row.** The estimated effect is 0.80; the CI is $[0.45, 1.15]$, which sits entirely outside the band on the positive side. The effect is detected and is larger than the SESOI.

This visual rule of thumb is, in fact, exactly equivalent to running the TOST procedure with the corresponding alpha level (with the standard convention of using a 90% CI for $\alpha = .05$, because the two one-sided tests "spend" alpha on each side).

### Running TOST in R with the `TOSTER` package

The `TOSTER` package by Lakens and Caldwell makes this very easy. The main function we will use is `t_TOST()` for two-sample, paired, or one-sample equivalence tests on raw data.

A worked example on our running dataset. In `vgames_study.csv`, `empathy` was measured *before* the experimental manipulation (which game participants were assigned to play). Random assignment to violent vs neutral conditions should produce groups with very similar pre-treatment empathy. We can ask: are the two condition means *equivalent* to each other on empathy, within a tolerance we declare in advance?

Suppose we decide that a SESOI of $\pm 0.5$ points on the 1-7 empathy scale would be the smallest difference we would consider meaningful (a half-point shift is a substantial fraction of a one-unit category).

```{r}
#| message: false
#| warning: false
library(TOSTER)
t_TOST(
  formula = empathy ~ condition,
  data = data,
  hypothesis = "EQU",       # equivalence
  eqb = 0.5,                # SESOI; equivalence bound of +/- 0.5
  eqbound_type = "raw",     # bound is in raw outcome units (not standardized)
  alpha = 0.05
)
```

The output reports:

- The observed group difference (empathy in the neutral group minus empathy in the violent group).
- The two one-sided tests, each with its own t-statistic, df, and p-value.
- A decision: equivalent? rejected? inconclusive?
- A 90% confidence interval on the difference (since TOST at $\alpha = .05$ corresponds to the 90% CI rule).

The standard reporting rule is straightforward: if **both** one-sided p-values are below your $\alpha$, declare equivalence. If neither is, you do not have evidence of equivalence (but you also do not have evidence against equivalence; you just have an inconclusive result). If one is below alpha and the other is not, your CI partially overlaps the bound, and again the result is inconclusive in the strict TOST sense.

A small variant of the function call uses Cohen's *d* as the SESOI metric, which is convenient when the original scale is hard to interpret:

```{r}
#| eval: false
t_TOST(
  formula = empathy ~ condition,
  data = data,
  hypothesis = "EQU",
  eqb = 0.4,                # SESOI = 0.4 standardized units
  eqbound_type = "SMD",     # standardized mean difference
  alpha = 0.05
)
```

### How to choose the SESOI for an equivalence test

The same logic from the earlier section applies, but the choice is now central rather than incidental. Anvari and Lakens (2021) discuss in detail how to justify a SESOI for psychology research; their summary is that the choice should be driven by *what would change a decision*, not by convention. For empathy on a 1-7 scale, a half-point difference is roughly half a response-category unit, which most researchers would treat as a meaningful shift. For aggression on a 1-10 scale, a SESOI of 0.5 to 1.0 is plausible. For a clinical outcome where 1 point on a depression scale corresponds to a measurable improvement in functioning, the SESOI would be even smaller. None of this is arbitrary; the SESOI is a declared, defensible value, chosen before data collection, that the equivalence test then uses as its target.

### Equivalence testing combined with the standard test

A common (and recommended) practice in modern psychology methodology is to run **both** a standard test and an equivalence test on the same data. The two tests together address two questions:

- The standard test asks "is there evidence of any effect?"
- The equivalence test asks "is there evidence that the effect is too small to matter?"

Four combinations are possible:

| Standard test | Equivalence test | Interpretation |
|--------------|-----------------|----------------|
| Significant | Equivalent | Effect detected, but smaller than the SESOI; statistically present but practically negligible. |
| Significant | Not equivalent | Effect detected and large enough to matter (the standard case for a "meaningful" finding). |
| Not significant | Equivalent | No detectable effect, and the data are consistent with effects too small to matter (a positive null result). |
| Not significant | Not equivalent | Inconclusive: we have neither detected an effect nor ruled out a meaningful one. |

Lakens (2017) argues that the bottom-left cell, "non-significant + equivalent," is the appropriate way to communicate a positive null finding, and the bottom-right cell, "not significant + not equivalent," is the result that most often gets mistakenly written up as "no effect" when it is actually inconclusive. The discipline of running both tests together prevents that confusion.

For more on equivalence testing, including extensions to correlations, ANOVA, and meta-analysis, Lakens (2017) is the most accessible primer, and Lakens (2022) Chapter 9 is a detailed treatment.

## Power analysis in practice: the `pwr` package

For the kinds of analyses you will run in psychology research, the `pwr` package in R is a sufficient tool. Common entry points:

```{r}
#| eval: false
library(pwr)

# t-test
pwr.t.test(d = 0.5, sig.level = .05, power = .80, type = "two.sample")

# correlation
pwr.r.test(r = 0.3, sig.level = .05, power = .80)

# one-way ANOVA
pwr.anova.test(k = 3, f = 0.25, sig.level = .05, power = .80)

# chi-squared
pwr.chisq.test(w = 0.3, df = 4, sig.level = .05, power = .80)

# multiple regression
pwr.f2.test(u = 3, f2 = 0.10, sig.level = .05, power = .80)
```

Each of these takes three of the four quantities (effect size, α, power, n) and returns the fourth. The effect size argument is parameterized in the conventional metric for each test (Cohen's d for t-tests, r for correlations, f for ANOVA, w for chi-squared, f² for multiple regression). The principle is identical across tests.

For more elaborate designs (multilevel models, mixed designs, structural equation models) the `simr` package and Monte Carlo simulation are the standard tools. They are beyond the scope of this chapter, but the workflow is the same: specify the design, specify the smallest effect you would care about, simulate many data sets, and count how often the procedure correctly rejects.

## What "serious" frequentist practice looks like

We now have all the pieces to describe the version of Neyman-Pearson hypothesis testing that Lakens, Cohen, and several generations of methodologists have advocated. It is more disciplined than the casual hybrid most researchers use, but it pays off in interpretability.

1. **Pre-specify the test.** Decide before data collection: what hypothesis, what statistic, what α, what test (one-tailed or two), what exclusion rules.
2. **Anchor n to the SESOI.** Choose the smallest effect size that would matter, and compute the required sample size from a power analysis.
3. **Run the test once.** No peeking, no optional stopping, no swapping the analysis after seeing the data.
4. **Report the result.** Statistical significance is one piece. Whether the observed effect is larger or smaller than the SESOI is another. Confidence intervals make both visible.
5. **Accept the long-run error rates.** Your individual decision might be wrong. The procedure is calibrated, not infallible. The calibration is the point.

This is what Neyman-Pearson originally proposed. It requires more commitment up-front than the hybrid most researchers use, but it produces decisions that are honestly calibrated and that connect to real-world consequences.

## Looking ahead

Chapter 25, the last chapter of Section 4, returns to the methodological issues we hinted at in Chapter 22. We have been talking about *one* test, run *once*, on a pre-specified hypothesis. What happens when researchers do not follow that discipline? In particular, what happens when many tests are run, or when the procedure depends on the data, or when only the significant results are reported? Those are the failure modes that drive much of the methodological literature in the last fifteen years: p-hacking, HARKing, the garden of forking paths, multiple-testing corrections, and the broader question of how often "significant" published findings are actually true. Ioannidis's famous PPV result enters there. And the practical discipline that addresses all of this together is **pre-registration**, which we will develop in detail.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "It is the probability of a Type I error occurring at least once across all the experiments a researcher conducts during their career on the same broad topic.",
  "It is the difference between the observed effect size and the true effect size in the population, computed as the bias of the estimator under repeated sampling.",
  "It is the probability of failing to detect a real effect, where the true effect is assumed to equal the conventional medium effect size of d = 0.5 in standard tests.",
  answer = "It is the long-run rate at which the procedure rejects a true null hypothesis; the maximum acceptable false-positive rate set before data are collected, conventionally 0.05."
)

opts_q2 <- c(
  "It is the probability that the null hypothesis is true after observing the data, computed via Bayesian updating from the prior probability set before the study.",
  "It is the probability that the observed effect size is reproducible in a future study with the same sample size and effect, assuming all assumptions of the test are met.",
  answer = "It is the long-run rate at which the procedure fails to reject H0 when H0 is actually false; the false-negative rate, conventionally written as beta and equal to 1 minus power.",
  "It is the probability of obtaining an exactly significant result with p = 0.05, given that the alternative hypothesis is correct in the population for the procedure used."
)

opts_q3 <- c(
  "Statistical power is a property of the test itself and does not depend on the alternative effect size or the sample size, only on the chosen alpha level.",
  "Statistical power is the rate at which a procedure replicates across independent samples drawn from the same population, regardless of the size of the true effect being tested.",
  answer = "Statistical power is the long-run rate at which the procedure correctly rejects H0 when H0 is actually false, **given a specific value of the true effect size**; without an anchored effect size, the word power is undefined.",
  "Statistical power is the probability that the procedure's reported p-value is below 0.05, conditional on the data being normally distributed and free of any influential observations."
)

opts_q4 <- c(
  "Alpha, power, effect size, and number of tails are independent quantities that can be chosen separately for any given test of a research hypothesis on a real dataset.",
  "Effect size determines the sample size required for any given alpha; power and the choice of test statistic are completely independent of these two quantities and add no further constraints.",
  "Alpha and beta are unrelated to sample size in the long run, and an experimenter can therefore freely choose alpha = .01 and beta = .05 at any sample size of n = 30 or larger.",
  answer = "Alpha, beta (or power), effect size, and sample size are linked by a fixed mathematical relationship; fixing any three of them determines the fourth, and software such as the pwr package performs the computation."
)

opts_q5 <- c(
  "The smallest effect size that the researcher's preferred theory has ever predicted in any of the previously published studies, regardless of the topic of the current investigation.",
  answer = "The smallest effect size below which the researcher would not change their behavior, conclusions, or recommendations, even if the effect were demonstrably real in the population; an anchor chosen in advance based on practical, theoretical, and decision-relevance considerations.",
  "The minimum effect size detectable at p < .05 given the sample size and noise level of the study, computed post hoc from the standard error of the regression coefficient of interest.",
  "The effect size that the convention of the research domain treats as small, typically a Cohen's d of 0.2 for two-sample comparisons or 0.1 for correlations between continuous variables."
)

opts_q6 <- c(
  "Without the SESOI the conventional rule of thumb at d = 0.5 still applies, so any claim of 80% power can be interpreted as referring to that medium effect size by default.",
  answer = "Without the SESOI the claim does not specify which effect size the power refers to, and two researchers can both say 80% power while one means d = 0.1 and the other means d = 1.0, two designs with completely different implications.",
  "Without the SESOI any computed power is purely speculative and cannot be used in any formal way to determine the sample size required for the planned hypothesis test.",
  "Without the SESOI the claim still has a precise meaning because the long-run false-positive rate alpha is fixed and the procedure remains calibrated regardless of the effect size."
)

opts_q7 <- c(
  "Statistical significance with a tiny effect is always strong evidence that the effect is real in the population, because a small effect with a small p-value indicates a precise estimate of a real difference.",
  "The effect being smaller than the SESOI is irrelevant for the interpretation of the result, since the SESOI plays no role once the test has already been run and the p-value has been computed and reported.",
  answer = "The effect is statistically detectable but practically negligible; statistical significance does not rescue practical irrelevance, and the appropriate interpretation is that there may be a real effect but one too small to change any decision based on the study.",
  "The result is uninterpretable because a small effect with a small p-value indicates a violation of the assumptions of the t-test or of the regression model used to compute the test statistic."
)

opts_q8 <- c(
  "Equivalence testing computes a Bayesian posterior probability that the null hypothesis is exactly true, given the observed data and a prior centered at zero with variance equal to the SESOI squared.",
  "Equivalence testing changes the alpha threshold of the standard hypothesis test to 0.5 so that the null hypothesis is rejected less often, producing claims of equivalence from non-significant results.",
  answer = "Equivalence testing reframes the question as 'is the effect smaller than the SESOI?' rather than 'is the effect different from zero?' and uses two one-sided tests against the SESOI bounds; rejecting both supports a positive claim of practical equivalence.",
  "Equivalence testing computes the same p-value as a standard hypothesis test but interprets a non-significant result as a positive demonstration that the population effect is exactly equal to zero."
)

opts_q9 <- c(
  "The pwr package replaces the need to specify effect size or sample size, because it computes the smallest detectable effect from the residual standard error of the model after the data are collected.",
  answer = "The pwr package takes three of the four quantities (alpha, power, effect size, sample size) and returns the fourth, with functions specialized for t-tests, correlations, ANOVA, chi-squared tests, and regression.",
  "The pwr package performs Bayesian power analysis by drawing many samples from the prior distribution of the effect size and computing the posterior probability that the test will reject.",
  "The pwr package adjusts the alpha level for multiple comparisons using Bonferroni and Benjamini-Hochberg corrections, ensuring that the family-wise error rate is controlled across all tests in a research program."
)

opts_q10 <- c(
  "The serious frequentist practice begins with the data and adapts the procedure to whatever the data look like, with the goal of maximizing the apparent statistical significance of the most interesting finding.",
  answer = "The serious frequentist practice pre-specifies the procedure (test, alpha, design, exclusion rules), anchors the sample size to the SESOI, runs the test once as specified, reports the result honestly, and accepts the calibrated long-run error rates as the basis for the inference.",
  "The serious frequentist practice never reports a p-value; instead it relies entirely on Bayes factors and posterior probabilities to communicate the evidence in favor of each hypothesis under consideration.",
  "The serious frequentist practice runs many candidate tests on the data and reports only the smallest p-value, while applying Bonferroni correction to keep the family-wise error rate at a conventional level."
)
```

**1. What is the alpha level (α) of a hypothesis test?**

`r longmcq(opts_q1)`

**2. What is the beta level (β)?**

`r longmcq(opts_q2)`

**3. What is statistical power, and why is the conventional one-sentence definition incomplete?**

`r longmcq(opts_q3)`

**4. What is the relationship between alpha, beta (or power), effect size, and sample size?**

`r longmcq(opts_q4)`

**5. What is the Smallest Effect Size of Interest (SESOI)?**

`r longmcq(opts_q5)`

**6. What is the problem with a claim like "my study has 80% power" when no effect size is named?**

`r longmcq(opts_q6)`

**7. A study designed with 80% power to detect a SESOI of 0.2 finds p = .002 with an observed effect of d = 0.05. How should this be interpreted?**

`r longmcq(opts_q7)`

**8. What does equivalence testing do, and why is it useful?**

`r longmcq(opts_q8)`

**9. How does the `pwr` package in R perform power analysis?**

`r longmcq(opts_q9)`

**10. What does "serious" frequentist practice (in the Neyman-Pearson sense, as developed by Lakens and others) look like?**

`r longmcq(opts_q10)`

## References

Anvari, F., & Lakens, D. (2021). Using anchor-based methods to determine the smallest effect size of interest. *Journal of Experimental Social Psychology*, 96, 104159. <https://doi.org/10.1016/j.jesp.2021.104159>

Lakens, D. (2017). Equivalence tests: A practical primer for t-tests, correlations, and meta-analyses. *Social Psychological and Personality Science*, 8(4), 355-362. <https://doi.org/10.1177/1948550617697177>

Lakens, D. (2022). *Improving your statistical inferences*. Chapter 2, "Error control." <https://lakens.github.io/statistical_inferences/02-errorcontrol.html>

Navarro, D. J. (2018). *Learning statistics with R: A tutorial for psychology students and other beginners* (Version 0.6), Chapter 11.8 ("Effect size, sample size and power"). <https://learningstatisticswithr.com>
