---
title: "Introduction to ggplot2: the grammar of graphics in practice"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
```

Section 5 of this book focuses on data visualization in practice and on communicating findings clearly. The principles of good data visualization were introduced in Chapter 10. The actual *code* for producing those visualizations was hidden in that chapter and in every chapter since, with the promise that we would come back to it. This is that chapter.

We will use `ggplot2`, the most widely used and most flexible data visualization package in R. It is built on a particular philosophy of how plots should be constructed, called the **grammar of graphics**. Understanding that philosophy is the key to using `ggplot2` well. Once the underlying logic clicks, building any plot becomes a matter of stacking the right pieces together rather than memorizing a long list of commands.

This is the first chapter in Section 5, and like the rest of the section, it is unapologetically code-heavy. The goal here is not to teach every option `ggplot2` offers (there are entire books for that, including [Wickham, Navarro, & Pedersen, 2024](https://ggplot2-book.org/)). The goal is to give you the core intuition for how `ggplot2` thinks about a plot, so you can read and write the code with confidence. The next chapters in Section 5 will build on this foundation to cover specific kinds of plots: visualizations for means and spread, count variables, distributions, coefficient and forest plots, and so on.

## Why `ggplot2`?

R has more than one way to make a plot. Base R has a `plot()` function that works fine for quick exploratory looks. There are other packages too. So why bother learning `ggplot2`?

Three reasons.

First, `ggplot2` is the de facto standard in modern R-based data analysis, especially in psychology and other empirical sciences. If you read methods sections of papers, look at code on GitHub, or take any modern R course, you will see `ggplot2` everywhere. Learning it gives you access to a huge body of community work.

Second, `ggplot2` is unusually principled. It is built on a small set of consistent ideas that, once learned, generalize across every type of plot you might want to make. Where base R's plotting commands tend to be ad hoc (a different command for each plot type, each with its own quirks), `ggplot2` uses the same logic to build a scatterplot, a histogram, a boxplot, a bar chart, a heatmap, or anything else. Learning the logic once pays off across every visualization you ever build.

Third, `ggplot2` produces publication-quality output by default. The plots look professional with very little effort, and the moment you need to customize them (which is often, in published research), the system is built to make customization principled rather than fiddly.

## The grammar of graphics, briefly

Chapter 10 introduced the **grammar of graphics** as a Venn diagram with three overlapping components: the *data*, the *aesthetics*, and the *geometry*. The idea behind `ggplot2`, which traces back to Wilkinson ([2005](https://link.springer.com/book/10.1007/0-387-28695-0)) and to Wickham ([2010](https://doi.org/10.1198/jcgs.2009.07098)), is that *every* statistical plot can be described as a combination of these three things plus a few decorations:

- **Data**: the underlying data frame. The rows and columns from which the plot is built.
- **Aesthetics**: how columns of the data are mapped to visual properties of the plot. The x-axis position, the y-axis position, the color of the points, the size of the bars, the shape of the markers, the transparency. Each of these is an *aesthetic*, and each one can be mapped to a column of the data.
- **Geometry** (the *geom*): the type of visual element used to represent the data. Points for a scatterplot, bars for a bar chart, lines for a line plot, a histogram for a histogram, and so on.

Build a plot by specifying these three things, and `ggplot2` figures out the rest.

A few extras then refine the result:

- **Scales**: how the values in the data map to the values on the axes (linear, log, categorical, etc.).
- **Facets**: how the plot is divided into small multiples (for example, one panel per experimental condition).
- **Themes**: the overall look (fonts, background, gridlines, title placement).
- **Labels**: axis titles, plot titles, captions, legends.

We will meet each of these in turn. The order matters less than the conceptual framing: data + aesthetics + geom is the heart of every plot.

## Your first plot: building it step by step

Let us build a plot piece by piece, using the `vgames_study` data we have used throughout the book. The goal will be a scatterplot of trait hostility against aggression, with violent and neutral conditions colored differently.

First, load the data.

```{r}
#| label: vgames-load
#| eval: false
vgames <- read.csv("data/vgames_study.csv")
```

```{r}
#| label: vgames-data
#| echo: false
vgames <- tryCatch(
  read.csv("data/vgames_study.csv"),
  error = function(e) {
    set.seed(1)
    n <- 60
    data.frame(
      participant_id = seq_len(2 * n),
      condition = rep(c("violent", "neutral"), each = n),
      trait_hostility = rnorm(2 * n, mean = 4, sd = 1.2),
      aggression = c(rnorm(n, 5.56, 1.6), rnorm(n, 4.21, 1.6))
    )
  }
)
```

The most minimal `ggplot2` call looks like this:

```{r}
#| label: fig-empty-plot
#| fig-width: 6
#| fig-height: 3.5
#| fig-align: center

ggplot(data = vgames, aes(x = trait_hostility, y = aggression))
```

Notice what just happened. We told `ggplot2`:

- the **data** is `vgames`,
- the **x aesthetic** is `trait_hostility` (so this column will determine horizontal position),
- the **y aesthetic** is `aggression` (this column will determine vertical position).

But we have not specified a **geom** yet. So `ggplot2` drew the canvas with the right axes, the right scale, and the right axis labels, but did not draw anything on it. There is nothing for the points to be points *of*. We need to add a geom.

Now add `geom_point()` to draw points at each row's (x, y) position:

```{r}
#| label: fig-first-scatter
#| fig-width: 6
#| fig-height: 3.5
#| fig-align: center

ggplot(data = vgames, aes(x = trait_hostility, y = aggression)) +
  geom_point()
```

That is your first scatterplot. The `+` symbol is how `ggplot2` adds layers to the plot. Read the code from top to bottom: "Start a plot using `vgames` data, map `trait_hostility` to x and `aggression` to y, then add a point for each row."

Now color the points by experimental condition. Inside `aes()`, we add `color = condition`:

```{r}
#| label: fig-scatter-colored
#| fig-width: 6.5
#| fig-height: 3.5
#| fig-align: center

ggplot(data = vgames, aes(x = trait_hostility, y = aggression,
                          color = condition)) +
  geom_point()
```

`ggplot2` has automatically assigned a different color to each level of `condition` and added a legend. We did not have to specify the colors manually or build the legend manually; that is the kind of thing the grammar takes care of when you map an aesthetic to a column.

Finally, add a regression line for each condition with `geom_smooth()`:

```{r}
#| label: fig-scatter-with-smooth
#| fig-width: 6.5
#| fig-height: 3.5
#| fig-align: center

ggplot(data = vgames, aes(x = trait_hostility, y = aggression,
                          color = condition)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x)
```

Notice how the new geom inherits the same x, y, and color mappings from the top-level `ggplot()` call. Because we mapped color to condition at the top, *both* the points and the lines are colored by condition. This is the power of the layered grammar: the data and aesthetic mappings flow downward through all the layers unless overridden.

Read the final code from top to bottom and you have the entire structure of the plot in front of you: start with this data, map these aesthetics, add points, add a smoothing line. Four lines of code, one publication-ready figure.

## Mapping vs setting: inside or outside `aes()`?

One of the most common sources of confusion when starting with `ggplot2` is the distinction between *mapping* an aesthetic to a column of data and *setting* an aesthetic to a constant value. The rule is simple: anything that depends on the data goes *inside* `aes()`; anything that is a constant goes *outside* `aes()`.

Compare these two calls:

```{r}
#| label: fig-mapping-vs-setting
#| fig-width: 8
#| fig-height: 3.2
#| fig-align: center

library(ggplot2)
plot_mapped <- ggplot(vgames, aes(x = trait_hostility, y = aggression)) +
  geom_point(aes(color = condition)) +
  labs(title = "color INSIDE aes(): mapped to data") +
  theme_minimal(base_size = 9)

plot_set <- ggplot(vgames, aes(x = trait_hostility, y = aggression)) +
  geom_point(color = "steelblue") +
  labs(title = "color OUTSIDE aes(): constant value") +
  theme_minimal(base_size = 9)

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(plot_mapped, plot_set, ncol = 2)
} else {
  gridExtra::grid.arrange(plot_mapped, plot_set, ncol = 2)
}
```

In the left panel, `color = condition` is *inside* `aes()`. We are mapping the color aesthetic to the values of the `condition` column. `ggplot2` picks two different colors automatically and adds a legend.

In the right panel, `color = "steelblue"` is *outside* `aes()`. We are setting the color of *all* points to a single fixed value, `"steelblue"`. There is no legend because there is no mapping from data to color; every point gets the same color regardless of the data.

The same rule applies to all aesthetics. `size = 3` outside `aes()` sets every point to the same size. `size = aggression` inside `aes()` would scale the point size by the aggression score. `alpha = 0.4` outside makes everything 40% transparent. `alpha = trait_hostility` inside would make each point's transparency depend on the hostility value.

When you see ggplot code that produces unexpected output, the first thing to check is usually whether an aesthetic ended up in the wrong place. A common mistake is writing `geom_point(aes(color = "steelblue"))`, which treats the literal string `"steelblue"` as if it were a column of data with a single value; `ggplot2` then assigns *some* color to that one-level column (often not steelblue at all) and adds a legend showing the literal text "steelblue". The fix is to move it outside: `geom_point(color = "steelblue")`.

## A short tour of common geoms

`ggplot2` has dozens of geoms. You will probably use eight or ten regularly. We illustrate the most common ones in turn, each on the `vgames_study` data.

**`geom_point()`** for scatterplots, as we have already seen.

**`geom_histogram()`** for the distribution of a single continuous variable. Only an x aesthetic is required:

```{r}
#| label: fig-histogram
#| fig-width: 6
#| fig-height: 3.0
#| fig-align: center

ggplot(vgames, aes(x = aggression)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue",
                 color = "white")
```

The `binwidth` argument controls how wide each bin is. The `fill` argument sets the bar color (note: `color` would set the *outline* color of the bars; for histograms and bars, the inside is controlled by `fill`).

**`geom_boxplot()`** for comparing the distribution of a continuous variable across groups:

```{r}
#| label: fig-boxplot
#| fig-width: 5.5
#| fig-height: 3.0
#| fig-align: center

ggplot(vgames, aes(x = condition, y = aggression)) +
  geom_boxplot(fill = "lightblue")
```

The box shows the middle 50% of the data (from the 25th to the 75th percentile), the line inside the box is the median, the whiskers extend out to the smallest and largest "non-outlying" values, and any points beyond the whiskers are shown individually as potential outliers.

**`geom_bar()`** and **`geom_col()`** for bar charts. `geom_bar()` counts the rows in each category for you; `geom_col()` plots a value you supply.

```{r}
#| label: fig-bar
#| fig-width: 5.5
#| fig-height: 3.0
#| fig-align: center

ggplot(vgames, aes(x = condition)) +
  geom_bar(fill = "steelblue")
```

Here, no y aesthetic is needed because `geom_bar()` is computing the row count for each level of `condition`. We will see in Chapter 31 why bar charts of *means* (as opposed to counts) are usually a bad idea and what to use instead.

**`geom_smooth()`** adds a smoothed regression line. We have already used it with `method = "lm"` for a linear fit. The default `method = "loess"` adds a non-parametric smooth that bends with the data.

**`geom_line()`** connects points with a line, useful for time series or for showing trends:

```{r}
#| label: fig-line-summary
#| fig-width: 5.5
#| fig-height: 3.0
#| fig-align: center

agg_summary <- vgames |>
  group_by(condition) |>
  summarize(
    mean_aggression = mean(aggression),
    .groups = "drop"
  )

ggplot(agg_summary, aes(x = condition, y = mean_aggression, group = 1)) +
  geom_line() +
  geom_point(size = 3)
```

The `group = 1` aesthetic tells `ggplot2` that the line should connect across categories (without it, ggplot would draw a line per group, which here means no line at all because each group has only one point).

**`geom_density()`** plots a smoothed estimate of the density of a continuous variable:

```{r}
#| label: fig-density
#| fig-width: 6
#| fig-height: 3.0
#| fig-align: center

ggplot(vgames, aes(x = aggression, fill = condition)) +
  geom_density(alpha = 0.4)
```

The `alpha = 0.4` makes the fills semi-transparent so we can see both densities even where they overlap. This kind of overlapping-density plot will return in Chapter 33 in the context of visualizing distributions.

These are the geoms you will reach for most often. Each follows the same pattern: a name like `geom_*()`, some required and optional arguments, and use of the inherited aesthetics from the top-level `ggplot()` call.

## Faceting: small multiples

A common need is to split a plot into multiple panels, one per category. This is called **faceting**. The function `facet_wrap()` lays the panels out in a wrapped grid, and `facet_grid()` lays them in a strict row-by-column grid.

```{r}
#| label: fig-facet-wrap
#| fig-width: 7
#| fig-height: 3.2
#| fig-align: center

ggplot(vgames, aes(x = trait_hostility, y = aggression)) +
  geom_point() +
  facet_wrap(~ condition)
```

The formula `~ condition` tells `ggplot2` to make one panel per unique value of `condition`. The same x and y axes are used in each panel by default (so the panels are directly comparable), but options exist to free the scales per panel if needed.

Faceting is one of `ggplot2`'s most powerful features. Whenever you find yourself thinking "I want to make the same plot for each subgroup of the data," facet it instead of writing a loop.

## Customizing the look: themes and labels

By default, `ggplot2` uses a grey background with white gridlines and a sensible default font. You can change all of this. The two most common customizations are themes (the overall visual style) and labels (the text on the plot).

The function `theme_minimal()` switches to a clean minimal look with a white background. Other built-in themes include `theme_classic()`, `theme_bw()`, `theme_light()`, and `theme_dark()`. Add them at the end of the plot:

```{r}
#| label: fig-theme-minimal
#| fig-width: 6.5
#| fig-height: 3.5
#| fig-align: center

ggplot(vgames, aes(x = trait_hostility, y = aggression,
                   color = condition)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  theme_minimal()
```

To set the title, axis labels, and other text, use `labs()`:

```{r}
#| label: fig-labeled
#| fig-width: 6.5
#| fig-height: 3.8
#| fig-align: center

ggplot(vgames, aes(x = trait_hostility, y = aggression,
                   color = condition)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(
    title    = "Aggression as a function of trait hostility",
    subtitle = "By experimental condition (violent vs neutral)",
    x        = "Trait hostility (1-7 scale)",
    y        = "Aggression (1-10 scale)",
    color    = "Condition",
    caption  = "Source: vgames_study (n = 120)"
  ) +
  theme_minimal()
```

Each entry of `labs()` is optional. Set as many or as few as you need. The argument names map to aesthetics, so `color = "Condition"` changes the legend title for the color mapping.

To fine-tune individual aspects of the theme (font size, legend position, gridline color), use `theme()` with specific arguments. For example, to move the legend to the bottom and increase the title size:

```{r}
#| label: fig-themed-fine
#| fig-width: 6.5
#| fig-height: 4
#| fig-align: center

ggplot(vgames, aes(x = trait_hostility, y = aggression,
                   color = condition)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(title = "Aggression as a function of trait hostility",
       x = "Trait hostility (1-7 scale)",
       y = "Aggression (1-10 scale)",
       color = "Condition") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title      = element_text(face = "bold", size = 14)
  )
```

`theme()` has a long list of options. Look up `?theme` when you need to control something specific.

## Saving a plot to a file

To save a plot to a file, use `ggsave()`. It picks the file format from the extension (`.png`, `.jpg`, `.pdf`, `.svg`, etc.) and writes the last plot you created to the chosen path.

```{r}
#| label: ggsave-example
#| eval: false

ggsave(
  filename = "hostility_aggression.png",
  width    = 6,
  height   = 4,
  units    = "in",
  dpi      = 300
)
```

By default, `ggsave()` saves the most recently displayed plot. To save a specific plot (one you have assigned to a variable), pass it as the `plot` argument:

```{r}
#| label: ggsave-assigned
#| eval: false

p <- ggplot(vgames, aes(x = trait_hostility, y = aggression)) +
  geom_point() +
  theme_minimal()

ggsave("my_plot.png", plot = p, width = 6, height = 4, dpi = 300)
```

For final figures in a paper, set `dpi = 300` or higher and check the visual result. For draft figures or exploratory output, defaults are usually fine.

## The coding intuition: how to read and write `ggplot2` code

The single most useful habit when reading or writing `ggplot2` code is to read it as a sentence. Every plot has the same structure:

> Start a plot using *this data*, map *these columns* to *these aesthetics*, then add *these layers* (points, lines, bars, etc.), then *facet* by *this column*, then change *the labels and the theme*.

Read the components in order. When you see `ggplot(vgames, aes(x = ..., y = ..., color = ...))`, that is the data and the global aesthetics. When you see `+ geom_point()`, that is a layer. When you see `+ facet_wrap(~ condition)`, that is a facet. When you see `+ labs(...) + theme_minimal()`, that is decoration. The plot is constructed in that order, and you should read it in that order.

When *writing* `ggplot2` code, follow the same logic. Start with the data and the essential aesthetics (typically x and y). Add the geom you want. If you need points colored or sized by some variable, add that aesthetic mapping (either at the top level for everything, or inside a single geom if just for one layer). If you need multiple panels, add a facet. If you need a particular look, add a theme and labels at the end.

The order of layers also matters visually: layers added later are drawn *on top* of earlier ones. If you want a regression line behind the points, write `geom_smooth()` first; if you want it in front, write `geom_smooth()` last. This is the same logic as drawing on a transparency stack.

Once you have internalized this pattern, you will be able to read any `ggplot2` code you encounter, and you will be able to construct any plot you can imagine by stacking the right components together.

The chapters in the rest of Section 5 apply this foundation to specific kinds of visualizations: visualizing means and spread (Chapter 31), count variables (Chapter 32), distributions (Chapter 33), and so on. Each chapter will introduce the geoms and customizations most useful for that specific purpose. The grammar stays the same throughout.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "The grammar of graphics is a strict set of rules in R that prescribes which plot type must be used for each kind of statistical analysis, with violations producing automatic warnings in the ggplot2 console.",
  answer = "The grammar of graphics is a conceptual framework in which every plot is built from a small set of consistent components (data, aesthetic mappings, geometric objects, scales, facets, themes), so that learning the components once generalizes to every plot type.",
  "The grammar of graphics is a coding convention specific to the ggplot2 package that requires plots to be written using nested function calls rather than the layered plus-sign syntax used in modern ggplot2 code.",
  "The grammar of graphics is a visual style imposed on ggplot2 output, governing default colors, fonts, and axis layouts, but having no influence on the underlying code or the choice of geometric objects."
)

opts_q2 <- c(
  "Aesthetics set inside aes() are constants that apply to all data points equally, while aesthetics set outside aes() are mappings that vary from row to row in the data, allowing for finer control over the visual encoding.",
  answer = "Aesthetics set inside aes() are mappings from columns of the data to visual properties, while aesthetics set outside aes() are constants that apply to all elements equally without using the data.",
  "Aesthetics set inside aes() are reserved for the x and y position only, while aesthetics set outside aes() must be used for color, shape, size, and all other visual properties of the plot.",
  "Aesthetics set inside aes() are evaluated only after the plot is drawn, while aesthetics set outside aes() are evaluated immediately when the ggplot call is constructed and stored in memory."
)

opts_q3 <- c(
  "geom_point() automatically computes the mean of the y variable for each unique value of the x variable, so the dots in a scatterplot represent group means rather than individual observations.",
  answer = "geom_point() places one point on the plot for each row of the data, with horizontal position determined by the x aesthetic and vertical position determined by the y aesthetic.",
  "geom_point() draws a point for each unique combination of x and y values, automatically summing duplicates so that a single point with high opacity represents many overlapping rows in the data.",
  "geom_point() randomly samples 100 points from the data to display on the plot, which keeps rendering fast for large datasets but means the displayed points do not exactly represent the underlying data."
)

opts_q4 <- c(
  "facet_wrap(~ condition) plots the entire dataset on a single panel, with the condition column controlling the color of each point but having no effect on the layout of the figure.",
  "facet_wrap(~ condition) divides the data by condition and computes a separate statistical summary for each level, displaying these summaries in a single panel as overlaid annotations.",
  answer = "facet_wrap(~ condition) divides the data by the values of the condition column and produces a separate panel of the same plot for each level of condition, with the panels arranged in a wrapped grid.",
  "facet_wrap(~ condition) splits the x-axis of the plot into separate panels by condition while sharing a single y-axis across all panels, allowing the reader to directly compare y values across conditions."
)

opts_q5 <- c(
  "Layers added later in the ggplot2 code are always drawn behind earlier layers, so the first geom added to the plot ends up on top of all subsequent layers in the final figure.",
  answer = "Layers added later in the ggplot2 code are drawn on top of earlier layers, so the order in which geoms are added determines which elements are visible in front when they overlap.",
  "Layers added later in the ggplot2 code are drawn in random order at render time, so the visual stacking of multiple geoms is not deterministic and may change between runs of the same code.",
  "Layers added later in the ggplot2 code are drawn only if the underlying data has changed since the previous layer was added, so adding multiple geoms to the same data results in only the first being displayed."
)

opts_q6 <- c(
  "ggsave() automatically uploads the saved plot to a cloud storage service for later retrieval, so the filename argument refers to a remote path rather than a location on the user's local disk.",
  "ggsave() saves only the underlying ggplot object as an R data file, so the resulting file must be loaded back into R and rendered with print() in order to view the image.",
  answer = "ggsave() writes the most recently displayed (or a specified) ggplot object to a file on disk, automatically choosing the image format from the filename extension and accepting arguments for size and resolution.",
  "ggsave() prints the plot to the active graphics device but does not write any file to disk; the user must subsequently use the file menu in RStudio to export the displayed plot manually."
)

opts_q7 <- c(
  "labs() controls how the data are filtered before plotting, so its arguments determine which subset of the dataset will be displayed in the final figure.",
  "labs() determines the statistical computations performed on each layer of the plot, with arguments such as title and subtitle controlling the type of summary calculated.",
  answer = "labs() sets the text labels that appear on the plot, with arguments mapping to titles, axis labels, legend titles, and captions; it does not change the underlying data or the geometric objects.",
  "labs() reorders the legend entries in alphabetical order and rotates the axis tick labels by ninety degrees, but it has no effect on the title, subtitle, or caption of the plot."
)
```

**1. Which of the following best describes the grammar of graphics as implemented in `ggplot2`?**

`r longmcq(opts_q1)`

**2. What is the difference between setting an aesthetic inside `aes()` and outside `aes()` in `ggplot2`?**

`r longmcq(opts_q2)`

**3. What does `geom_point()` do in a `ggplot2` call?**

`r longmcq(opts_q3)`

**4. What does `facet_wrap(~ condition)` do in a `ggplot2` call?**

`r longmcq(opts_q4)`

**5. How does the order in which layers are added affect the appearance of a `ggplot2` plot?**

`r longmcq(opts_q5)`

**6. What does the `ggsave()` function do in `ggplot2`?**

`r longmcq(opts_q6)`

**7. What does the `labs()` function control in a `ggplot2` plot?**

`r longmcq(opts_q7)`

## References

Wickham, H. (2010). A layered grammar of graphics. *Journal of Computational and Graphical Statistics, 19*(1), 3-28. <https://doi.org/10.1198/jcgs.2009.07098>

Wickham, H., Navarro, D., & Pedersen, T. L. (2024). *ggplot2: Elegant graphics for data analysis* (3rd ed.). Springer. <https://ggplot2-book.org/>

Wilkinson, L. (2005). *The grammar of graphics* (2nd ed.). Springer. <https://link.springer.com/book/10.1007/0-387-28695-0>
