---
title: "Visualizing averages, spread, and conditional means"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
library(tidyr)
```

This chapter is about the most common visualization task in psychology and the social-behavioral sciences: showing the average of a continuous variable, the spread around that average, and (often) how the average depends on another variable. The chapter has three messages.

First, the most common way researchers visualize a mean, the **bar chart with error bars**, is almost always a bad choice. We will see why, and we will see several alternatives that work much better.

Second, "the spread" is not one quantity. Standard deviation, standard error, and 95% confidence interval are three different things, each suited to a different question. Picking the wrong one is one of the most common mistakes in published figures.

Third, when the average depends on another variable, the visualization is no longer "the mean of Y" but something more interesting: "Y as a function of X." This shift in framing changes the entire question the figure is answering, and it is worth thinking carefully about which comparison the figure is asking the reader to make.

We will use the `vgames_study` data throughout. Load it.

```{r}
#| label: vgames-load
#| eval: false
vgames <- read.csv("data/vgames_study.csv")
```

```{r}
#| label: vgames-data
#| echo: false
vgames <- tryCatch(
  read.csv("data/vgames_study.csv"),
  error = function(e) {
    set.seed(1)
    n <- 60
    data.frame(
      participant_id  = seq_len(2 * n),
      condition       = rep(c("violent", "neutral"), each = n),
      hours_weekly    = c(rnorm(n, 8, 4), rnorm(n, 7, 4)),
      trait_hostility = rnorm(2 * n, mean = 4, sd = 1.2),
      empathy         = rnorm(2 * n, mean = 4.5, sd = 1.1),
      aggression      = c(rnorm(n, 5.56, 1.6), rnorm(n, 4.21, 1.6))
    )
  }
)
```

## The problem with bar charts of means

The bar chart with error bars on a mean is the default figure in many psychology papers. It looks like this:

```{r}
#| label: fig-barplot
#| fig-width: 5.5
#| fig-height: 3.4
#| fig-align: center

agg_summary <- vgames |>
  group_by(condition) |>
  summarize(
    mean_aggression = mean(aggression),
    sd_aggression   = sd(aggression),
    n               = n(),
    se              = sd_aggression / sqrt(n),
    .groups = "drop"
  )

ggplot(agg_summary, aes(x = condition, y = mean_aggression)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_errorbar(aes(ymin = mean_aggression - se,
                    ymax = mean_aggression + se),
                width = 0.15) +
  labs(x = "Condition", y = "Mean aggression") +
  theme_minimal()
```

The plot is clear: the violent condition has a higher mean aggression score than the neutral condition. But it is also concealing almost everything else about the data. The reader cannot see:

- how many participants are in each group,
- how spread out the individual scores are,
- whether the distributions are symmetric or skewed,
- whether there are outliers,
- how much the two distributions overlap.

These are not minor details. They are most of what we want to know about the data. The bar plus error bar reduces the entire dataset of 120 participants to four numbers (two means and two standard errors), and throws everything else away.

Worse, the bar chart actively misleads in a subtle way. The vertical bar starts at zero and rises to the mean. The visual impression is that the data go from zero to the mean. But the *data* are aggression scores from 1 to 10; very few participants score near zero. The bar's length is not a quantity that exists in the data; it is an artifact of the choice of plot.

The most direct demonstration of the problem is to show what completely different distributions look like when collapsed into the same bar plus error bar. The figure below shows three hypothetical datasets that all produce the same mean and standard error, but look completely different when the underlying distributions are exposed.

```{r}
#| label: fig-same-bar-different-distribution
#| fig-width: 9
#| fig-height: 4.2
#| fig-align: center

set.seed(42)
n_each <- 60

# Three datasets with the same mean (~5) and similar SE
df_normal   <- data.frame(score = rnorm(n_each, mean = 5, sd = 1.2),
                          dataset = "Normal")
df_bimodal  <- data.frame(score = c(rnorm(n_each / 2, mean = 3.5, sd = 0.4),
                                    rnorm(n_each / 2, mean = 6.5, sd = 0.4)),
                          dataset = "Bimodal")
df_skewed   <- data.frame(score = 10 - rexp(n_each, rate = 1/2) +
                                    rnorm(n_each, 0, 0.1),
                          dataset = "Skewed")
df_skewed$score <- df_skewed$score - mean(df_skewed$score) + 5

all_three <- rbind(df_normal, df_bimodal, df_skewed)
all_three$dataset <- factor(all_three$dataset,
                            levels = c("Normal", "Bimodal", "Skewed"))

summary_three <- all_three |>
  group_by(dataset) |>
  summarize(m = mean(score), se = sd(score) / sqrt(n()), .groups = "drop")

# Bar version
bar_three <- ggplot(summary_three, aes(x = dataset, y = m)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_errorbar(aes(ymin = m - se, ymax = m + se), width = 0.15) +
  labs(x = NULL, y = "Mean score",
       title = "Bar plot view (the misleading one)") +
  coord_cartesian(ylim = c(0, 8)) +
  theme_minimal(base_size = 10)

# Jitter version
jitter_three <- ggplot(all_three, aes(x = dataset, y = score)) +
  geom_jitter(width = 0.15, alpha = 0.5, color = "steelblue") +
  stat_summary(fun = mean, geom = "point", size = 3, color = "firebrick") +
  labs(x = NULL, y = "Score",
       title = "Individual data view (the honest one)") +
  coord_cartesian(ylim = c(0, 8)) +
  theme_minimal(base_size = 10)

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(bar_three, jitter_three, ncol = 2)
} else {
  gridExtra::grid.arrange(bar_three, jitter_three, ncol = 2)
}
```

The three bar plots look almost identical. The three jittered-point plots reveal that the underlying data are completely different: one is roughly normally distributed, one is bimodal (two clusters with nothing in the middle), and one is skewed. Anyone reading only the bar chart would draw the same conclusion about all three datasets, even though they are telling very different stories.

The general principle is **show your data when you can**. If the dataset is small or moderate (typical psychology experiments are under a few hundred participants), it costs almost nothing to show every individual point, and it conveys vastly more information than any summary plot.

## Alternatives that show the distribution

Several alternatives to the bar plot show the distribution of the data rather than hiding it. We work through the main ones in order of how much information they convey.

**Jittered points.** The simplest and often best option. Plot every individual observation as a point, with a small horizontal jitter to keep overlapping points visible.

```{r}
#| label: fig-jitter
#| fig-width: 5.5
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = condition, y = aggression)) +
  geom_jitter(width = 0.15, alpha = 0.5, color = "steelblue") +
  labs(x = "Condition", y = "Aggression") +
  theme_minimal()
```

Every participant is visible. The reader can see the sample size, the spread, the location of the central mass, and any outliers. There are no hidden summaries.

To make the mean visible, layer it on top with `stat_summary()`:

```{r}
#| label: fig-jitter-with-mean
#| fig-width: 5.5
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = condition, y = aggression)) +
  geom_jitter(width = 0.15, alpha = 0.5, color = "steelblue") +
  stat_summary(fun = mean, geom = "point",
               color = "firebrick", size = 3) +
  labs(x = "Condition", y = "Aggression") +
  theme_minimal()
```

The dark red points are the means, sitting on top of the cloud of individual observations. This figure shows everything the bar chart showed (the comparison of means) plus everything the bar chart hid (the underlying data).

**Boxplots.** A more compact summary that shows the median, the middle 50% of the data, the typical range, and outliers all at once.

```{r}
#| label: fig-boxplot
#| fig-width: 5.5
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = condition, y = aggression)) +
  geom_boxplot(fill = "lightblue", width = 0.5) +
  labs(x = "Condition", y = "Aggression") +
  theme_minimal()
```

Boxplots are useful when the dataset is large enough that individual points would overlap into mush. They show the basic shape (symmetric vs skewed), the spread (the box height), and any outliers (the points beyond the whiskers). They hide the actual sample size, however; a boxplot of *n* = 10 looks the same as a boxplot of *n* = 1000.

**Violin plots.** A smoothed density estimate, drawn symmetrically. Violin plots reveal multi-modal distributions that boxplots would hide.

```{r}
#| label: fig-violin
#| fig-width: 5.5
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = condition, y = aggression)) +
  geom_violin(fill = "lightblue", color = "steelblue4") +
  labs(x = "Condition", y = "Aggression") +
  theme_minimal()
```

The width of the violin at any height shows how dense the data are at that value. A "lumpy" violin with two bulges would reveal a bimodal distribution that a boxplot or bar chart would smooth over.

**Combined plots.** The most informative figures combine multiple geoms. A common pattern is violin (for the shape), boxplot (for the central summary), and jittered points (for the individual data):

```{r}
#| label: fig-combined
#| fig-width: 5.5
#| fig-height: 3.6
#| fig-align: center

ggplot(vgames, aes(x = condition, y = aggression)) +
  geom_violin(fill = "lightblue", color = NA, alpha = 0.5) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.35,
              color = "steelblue4", size = 0.9) +
  labs(x = "Condition", y = "Aggression") +
  theme_minimal()
```

The violin shows the smoothed distribution, the narrow boxplot inside shows the median and quartiles, and the jittered points show every observation. The figure does the work of three figures at once. Order matters: violins are drawn first (in the back), then the boxplot inside, then the points on top.

For most psychology applications, the combined plot or the jitter-with-mean plot is a strict improvement over the bar plot. The only situation in which the bar plot is preferred is when displaying *counts* (number of observations in each category), which we cover in Chapter 32.

## Showing the spread: SE, SD, CI

"The spread" is not one quantity. The three most common candidates each measure something different, and the choice between them depends on what question your figure is asking.

**Standard deviation (SD).** The average distance between an individual observation and the mean. SD describes the variability *of the data*. A larger SD means more variable individuals. SD is in the same units as the data (e.g., aggression-scale points). It does not depend on sample size; if you collect more participants from the same population, the SD stays roughly the same.

**Standard error (SE).** The standard deviation of the *sampling distribution* of the mean (Chapter 21). It describes the precision of *the mean as an estimate*. A smaller SE means a more precise estimate. SE is computed as SD divided by the square root of *n*, so it shrinks as the sample size grows. The same dataset collected at *n* = 30 has the same SD as at *n* = 300, but the SE at *n* = 300 is about 3.16 times smaller. SE answers a question about how well we know the mean, not about how variable the individuals are.

**95% confidence interval (CI).** A symmetric interval around the mean, of approximately ±1.96 × SE for large samples. The CI is interpreted as a long-run capture procedure (Chapter 22): 95% of intervals constructed this way will contain the true population mean. The CI is the most directly interpretable of the three for inferential purposes; if it excludes a hypothesized value (often zero), the two-sided test against that value rejects at α = .05.

The figure below shows the difference visually. Each panel shows the same `vgames_study` data, but with error bars representing SD, SE, or 95% CI respectively.

```{r}
#| label: fig-error-bars
#| fig-width: 9
#| fig-height: 3.6
#| fig-align: center

three_bars <- vgames |>
  group_by(condition) |>
  summarize(
    m  = mean(aggression),
    sd = sd(aggression),
    n  = n(),
    se = sd / sqrt(n),
    ci = qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )

# Helper function to make one panel
make_panel <- function(low_var, high_var, title) {
  ggplot(three_bars,
         aes(x = condition, y = m,
             ymin = m - .data[[low_var]],
             ymax = m + .data[[high_var]])) +
    geom_point(color = "firebrick", size = 3) +
    geom_errorbar(width = 0.15, color = "firebrick") +
    coord_cartesian(ylim = c(0, 9)) +
    labs(x = NULL, y = "Aggression", title = title) +
    theme_minimal(base_size = 10)
}

p_sd <- make_panel("sd", "sd", "Error bars = ±1 SD\n(spread of data)")
p_se <- make_panel("se", "se", "Error bars = ±1 SE\n(precision of mean)")
p_ci <- make_panel("ci", "ci", "Error bars = ±95% CI\n(inferential interval)")

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_sd, p_se, p_ci, ncol = 3)
} else {
  gridExtra::grid.arrange(p_sd, p_se, p_ci, ncol = 3)
}
```

The bars in the SD panel are much longer than in the SE panel, because SD does not shrink with sample size while SE does. The CI is roughly 2 × SE, so the CI panel is between the SE and the SD panel in width.

A few recommendations follow from this. Use **SD** when your point is about the variability of the individuals (for example, when discussing the substantive interpretation of effect sizes like Cohen's *d*). Use **SE** when your point is about the precision of the mean estimate (rarely useful on its own; usually superseded by the CI). Use **95% CI** when your point is about inference (the standard choice for most published figures showing group means). And whatever you choose, label it explicitly. A figure with error bars and no caption saying which kind of error bars they are is uninterpretable.

For an even more interpretable variant, plot the *difference* between conditions with its CI on a floating axis, as we did in Chapter 27 (section 27.10). The difference and its CI is what frequentist inference is really about; showing it directly removes the need for the reader to mentally subtract two means and combine their SEs.

## Easy customization

The plots above are clean but generic. A few small changes make them publication-ready.

**Reorder the categories.** R defaults to alphabetical order, which is rarely the order you want for a figure. The `forcats::fct_relevel()` function changes the order:

```{r}
#| label: fig-reordered
#| fig-width: 5.5
#| fig-height: 3.4
#| fig-align: center

vgames$condition_ordered <- forcats::fct_relevel(vgames$condition,
                                                 "neutral", "violent")

ggplot(vgames, aes(x = condition_ordered, y = aggression)) +
  geom_jitter(width = 0.15, alpha = 0.5, color = "steelblue") +
  stat_summary(fun = mean, geom = "point",
               color = "firebrick", size = 3) +
  labs(x = "Condition", y = "Aggression") +
  theme_minimal()
```

Now the neutral condition is on the left and the violent condition on the right, which matches the substantive "control, then treatment" ordering most readers expect.

**Choose colors.** The defaults are functional but not always good for accessibility or print. Use `scale_color_manual()` for points and lines, `scale_fill_manual()` for filled shapes:

```{r}
#| label: fig-recolored
#| fig-width: 5.8
#| fig-height: 3.5
#| fig-align: center

ggplot(vgames, aes(x = condition_ordered, y = aggression,
                   color = condition_ordered)) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  scale_color_manual(values = c("neutral" = "#377eb8",
                                "violent" = "#e41a1c"),
                     guide  = "none") +
  labs(x = "Condition", y = "Aggression") +
  theme_minimal()
```

The hex codes are from the ColorBrewer palette, which is widely used because the colors are designed to be distinguishable in print and by readers with most forms of color vision deficiency.

**Add an annotation.** A reference line, a label, or a small piece of text makes the figure self-explanatory.

```{r}
#| label: fig-annotated
#| fig-width: 5.8
#| fig-height: 3.5
#| fig-align: center

ggplot(vgames, aes(x = condition_ordered, y = aggression)) +
  geom_jitter(width = 0.15, alpha = 0.5, color = "steelblue") +
  stat_summary(fun = mean, geom = "point",
               color = "firebrick", size = 3) +
  geom_hline(yintercept = mean(vgames$aggression),
             linetype = "dashed", color = "grey50") +
  annotate("text", x = 0.6, y = mean(vgames$aggression) + 0.3,
           label = "Grand mean", hjust = 0, size = 3.2,
           color = "grey40") +
  labs(x = "Condition", y = "Aggression") +
  theme_minimal()
```

The dashed grey line is the grand mean across both conditions; it gives the reader a reference point to see how each condition's mean sits relative to the overall average. Annotations like this are usually quick to add and significantly improve the figure's standalone interpretability.

## Conditional means: lines and regressions

So far the figures have shown the mean of aggression *as a function of condition*, a categorical variable. Often we want to show the mean of one continuous variable as a function of another continuous variable. The figure then becomes a scatterplot with a fitted line on top.

```{r}
#| label: fig-scatter-lm
#| fig-width: 6
#| fig-height: 3.5
#| fig-align: center

ggplot(vgames, aes(x = trait_hostility, y = aggression)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE,
              color = "firebrick", formula = y ~ x) +
  labs(x = "Trait hostility (1-7)",
       y = "Aggression (1-10)") +
  theme_minimal()
```

The line is the predicted value of aggression at each level of trait hostility, from a linear regression. The shaded band around the line is the 95% confidence interval on the predicted line (computed automatically by `geom_smooth()`). At any value of trait hostility, the shaded band tells you the range of plausible values for the *mean* aggression at that hostility level, not for individual scores.

When there are multiple groups, you can fit one line per group by mapping a color aesthetic:

```{r}
#| label: fig-scatter-by-group
#| fig-width: 6.5
#| fig-height: 3.5
#| fig-align: center

ggplot(vgames, aes(x = trait_hostility, y = aggression,
                   color = condition_ordered)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x) +
  scale_color_manual(values = c("neutral" = "#377eb8",
                                "violent" = "#e41a1c"),
                     name = "Condition") +
  labs(x = "Trait hostility (1-7)",
       y = "Aggression (1-10)") +
  theme_minimal()
```

Now each condition gets its own regression line and its own confidence band. The figure shows two things at once: the marginal mean of aggression by condition (the height of each line at any given hostility value) and the relationship between hostility and aggression within each condition (the slope of each line). This is the visual fingerprint of a moderation effect (Chapter 18): the slope of the relationship depends on the value of another predictor.

If the relationship is non-linear, switch `method = "lm"` to `method = "loess"` for a smoother that bends with the data:

```{r}
#| label: fig-scatter-loess
#| fig-width: 6
#| fig-height: 3.5
#| fig-align: center

ggplot(vgames, aes(x = trait_hostility, y = aggression)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE,
              color = "firebrick",
              formula = y ~ x, span = 0.75) +
  labs(x = "Trait hostility (1-7)",
       y = "Aggression (1-10)") +
  theme_minimal()
```

The loess smoother is non-parametric: it does not assume a specific functional form. For exploratory plots, where you do not yet know what shape the relationship has, loess is often a better first look than the linear smoother.

## What is being compared, and to what reference?

Every figure showing means or conditional means is implicitly answering a question. The shape of the question determines what the figure should look like.

Three families of questions are particularly common in psychology, and they each call for a slightly different figure.

**"What is the average of *Y*, and how variable is it?"** The simplest question. Answer it with one of the distribution-showing plots from section 31.3 (jitter, boxplot, violin, or a combination), with the mean visible.

**"How does the average of *Y* compare between groups?"** The standard between-groups question. Use a plot that shows the distribution of *Y* in each group side by side (the combined plot is excellent for this). Add the mean and an appropriate measure of inferential precision (the 95% CI on each group's mean, *not* the SD or SE). A better alternative, when feasible, is to plot the *difference* between groups with its CI on a floating axis, because the difference and its CI are what an inferential test of the comparison directly delivers (Chapter 27, section 27.10). The visual question becomes "does the CI on the difference exclude zero?", which is exactly what a frequentist test answers.

**"How does the average of *Y* depend on *X*?"** The conditional-mean question. Use a scatterplot with a fitted line. The visual question becomes "what does the slope look like, and how precisely is it estimated?" (the shaded band around the line). If there are multiple groups, plotting one line per group lets the reader see whether the slope differs across groups (the moderation question of Chapter 18).

A fourth question is worth flagging because it is often answered with the wrong figure.

**"Is the effect of interest different from zero?"** Many researchers answer this question by plotting two group means side by side and asking the reader to mentally subtract them. This works, but it is indirect: the reader has to combine the two CIs in their head (and most readers, including most researchers, will get the eyeball calculation wrong; recall Chapter 27 section 27.6 on overlapping CIs). The direct answer is a single quantity (the difference) plotted with its own CI, with zero marked. If the CI does not contain zero, the effect is non-zero at the chosen confidence level; if it does, the data are consistent with no effect.

The general lesson is to think about which question your figure is asking the reader to answer, and then choose a plot that puts the answer directly in front of them. The bar-plus-error-bar template asks the reader to do extra mental work and hides most of the data. The combined plot, the difference-with-CI plot, and the scatter-with-fitted-line are all visualizations that put the answer to a specific substantive question directly on the page, with the data still showing.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "Bar charts of means are good visualizations because they make the differences between groups easy to see at a glance, and adding error bars provides all the information about variability that a reader needs.",
  answer = "Bar charts of means hide the underlying distribution of the data and the sample size, can be produced from very different distributions that happen to share the same mean and SE, and visually emphasize the area from zero to the mean rather than the data themselves.",
  "Bar charts of means are inappropriate only when the data are categorical, because the bar height incorrectly implies a continuous measurement of the underlying category; for continuous outcomes, bar charts are the appropriate default.",
  "Bar charts of means are misleading specifically because the error bars on top of each bar are always twice as long as the underlying standard error, which exaggerates the precision of the estimate by a factor of two."
)

opts_q2 <- c(
  "The standard deviation, standard error, and 95% confidence interval are three equivalent ways of expressing the same quantity, differing only in their scale: SD is on the raw scale, SE is divided by the square root of n, and CI is multiplied by 1.96.",
  answer = "The standard deviation describes the variability of the individual observations, the standard error describes the precision of the mean as an estimate, and the 95% confidence interval is an inferential interval that captures the population mean 95% of the time in the long run.",
  "The standard deviation describes the precision of the mean as an estimate, the standard error describes the variability of the individual observations, and the 95% confidence interval is a Bayesian credibility statement about where the true mean lies.",
  "The standard deviation, standard error, and 95% confidence interval describe different summaries of the data but should never be plotted as error bars; only the interquartile range from a boxplot is an appropriate visual representation of uncertainty."
)

opts_q3 <- c(
  "A jittered-points plot of group means is suitable only when the sample size is exactly equal across groups, and breaks down when groups have unequal numbers of participants because the visual density becomes incomparable across groups.",
  "A jittered-points plot of group means cannot show the central tendency of the data, so the researcher must add a separate bar chart of means alongside it to allow the reader to compare group averages.",
  answer = "A jittered-points plot shows every individual observation with a small horizontal random offset to keep overlapping points visible, and can be combined with a layered mean point or summary statistic to communicate both the distribution and the central tendency in a single figure.",
  "A jittered-points plot replaces each observation with a randomly generated synthetic data point at the same y-value, which preserves the marginal distribution but breaks any connection between the plotted points and the original observations."
)

opts_q4 <- c(
  "When fitting separate regression lines per group with geom_smooth(method = 'lm'), all groups must have the same number of observations, otherwise ggplot2 will throw an error about unbalanced data.",
  answer = "When fitting separate regression lines per group with geom_smooth(method = 'lm'), mapping a color aesthetic to the grouping variable causes ggplot2 to compute one fitted line per group, with its own confidence band, allowing the reader to see whether the slope of the relationship differs across groups.",
  "When fitting separate regression lines per group with geom_smooth(method = 'lm'), the lines are always forced to share a common intercept, with only the slopes allowed to differ between groups in a constrained parallel-lines model.",
  "When fitting separate regression lines per group with geom_smooth(method = 'lm'), the shaded confidence band around each line represents a 95% prediction interval for an individual observation in that group rather than a confidence interval on the predicted mean."
)

opts_q5 <- c(
  "Plotting the difference between two group means alongside its 95% confidence interval is statistically inappropriate because the difference is a derived quantity that introduces additional variance and should not be displayed in a published figure.",
  "Plotting the difference between two group means alongside its 95% confidence interval requires a Bayesian framework because frequentist confidence intervals are not defined for derived quantities like differences between means.",
  answer = "Plotting the difference between two group means alongside its 95% confidence interval directly answers the inferential question of whether the difference is non-zero, removing the need for the reader to mentally subtract the means and combine the two separate CIs.",
  "Plotting the difference between two group means alongside its 95% confidence interval is appropriate only when the two groups have equal variance, because the difference's CI is undefined when the variances differ across groups."
)

opts_q6 <- c(
  "The loess smoother fitted by geom_smooth(method = 'loess') is a parametric procedure that fits a global polynomial of degree 3 to the data, with the user able to change the polynomial degree but not the underlying functional form.",
  answer = "The loess smoother fitted by geom_smooth(method = 'loess') is a non-parametric procedure that does not assume a specific functional form, making it a useful exploratory choice when the shape of the relationship between X and Y is not yet known.",
  "The loess smoother fitted by geom_smooth(method = 'loess') is mathematically identical to the linear smoother fitted by geom_smooth(method = 'lm'), but with a different default color scheme and confidence-band style.",
  "The loess smoother fitted by geom_smooth(method = 'loess') is appropriate only for time-series data with autocorrelated errors, and produces misleading results when applied to independent cross-sectional observations of the kind found in vgames_study."
)

opts_q7 <- c(
  "When deciding how to visualize a comparison, the reader's eye is best served by always plotting the two group means with overlapping confidence intervals, because the overlap pattern is a reliable visual cue for statistical significance.",
  answer = "When deciding how to visualize a comparison, the figure should put the answer to the specific substantive question (the comparison, the conditional mean, the difference, etc.) directly on the page in a way that does not require the reader to perform mental arithmetic or combine separate uncertainty intervals in their head.",
  "When deciding how to visualize a comparison, the figure should aim to maximize the visual complexity of the display so that the reader can extract every possible feature of the data without having to consult the accompanying text.",
  "When deciding how to visualize a comparison, the figure must always include a bar chart as the primary visualization element, with other elements (points, lines, etc.) added only as supplementary annotations because bar charts are the most familiar template for readers in psychology."
)
```

**1. Why are bar charts of means commonly considered a poor choice for visualizing group differences?**

`r longmcq(opts_q1)`

**2. What is the difference between the standard deviation, the standard error, and the 95% confidence interval when shown as error bars on a group mean?**

`r longmcq(opts_q2)`

**3. What does a jittered-points plot of group means show, and how can the central tendency be added?**

`r longmcq(opts_q3)`

**4. When using `geom_smooth(method = "lm")` with a color aesthetic mapped to a grouping variable, what does the resulting figure show?**

`r longmcq(opts_q4)`

**5. What is the advantage of plotting the difference between two group means with its 95% confidence interval, rather than plotting the two means with their individual CIs?**

`r longmcq(opts_q5)`

**6. What does the loess smoother fitted by `geom_smooth(method = "loess")` do?**

`r longmcq(opts_q6)`

**7. Which of the following best describes the principle that should guide the choice of visualization for a comparison?**

`r longmcq(opts_q7)`

## References

Cumming, G. (2014). The new statistics: Why and how. *Psychological Science, 25*(1), 7-29. <https://doi.org/10.1177/0956797613504966>

Weissgerber, T. L., Milic, N. M., Winham, S. J., & Garovic, V. D. (2015). Beyond bar and line graphs: Time for a new data presentation paradigm. *PLOS Biology, 13*(4), e1002128. <https://doi.org/10.1371/journal.pbio.1002128>
