---
title: "Visualizing count variables"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
library(tidyr)
```

Chapter 31 argued that bar charts of *means* are usually a bad choice. This chapter is about the other use of bar charts: counts. **Count variables** are categorical or discrete: gender, condition, response choice, year, age band. A count visualization answers the question "how many?" for each category. For this question, the bar chart is not just acceptable but appropriate; it is exactly what the visual element was designed for. The chapter covers how to build clean count plots, how to handle two categorical variables together, when to switch from raw counts to proportions, how to order and customize, and a useful alternative for when a long list of categories starts to crowd a bar chart.

We use the `vgames_study` data throughout. Load it.

```{r}
#| label: vgames-load
#| eval: false
vgames <- read.csv("data/vgames_study.csv")
```

```{r}
#| label: vgames-data
#| echo: false
vgames <- tryCatch(
  read.csv("data/vgames_study.csv"),
  error = function(e) {
    set.seed(1)
    n <- 60
    data.frame(
      participant_id  = seq_len(2 * n),
      condition       = rep(c("violent", "neutral"), each = n),
      gender          = sample(c("woman", "man", "non-binary"),
                               2 * n, replace = TRUE,
                               prob = c(0.39, 0.52, 0.09)),
      age             = sample(18:35, 2 * n, replace = TRUE),
      aggression      = c(rnorm(n, 5.56, 1.6), rnorm(n, 4.21, 1.6))
    )
  }
)
```

## The bar chart, used well

The simplest count visualization is a single bar chart showing how many participants fall in each level of one categorical variable. Here is the count of participants in each experimental condition:

```{r}
#| label: fig-simple-bar
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

ggplot(vgames, aes(x = condition)) +
  geom_bar(fill = "steelblue", width = 0.6) +
  labs(x = "Condition", y = "Number of participants") +
  theme_minimal()
```

That is the basic pattern: one aesthetic (`x`) mapped to the categorical variable, and `geom_bar()` computes the count automatically. The bar height equals the number of rows in each category. There is no `y` aesthetic to specify because the y value is being computed for you. Bar charts of counts are honest in a way that bar charts of means are not, because the height of the bar corresponds to a real countable quantity (number of participants), not to a derived summary that throws away the distribution.

If you already have a summary table with the counts pre-computed, use `geom_col()` instead. `geom_col()` plots the value you supply as the y, rather than computing the count internally:

```{r}
#| label: fig-bar-from-summary
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

cond_counts <- vgames |>
  count(condition)

ggplot(cond_counts, aes(x = condition, y = n)) +
  geom_col(fill = "steelblue", width = 0.6) +
  labs(x = "Condition", y = "Number of participants") +
  theme_minimal()
```

`geom_bar()` and `geom_col()` produce visually identical output. The choice depends on whether your data are the raw observations (use `geom_bar()`) or a pre-computed summary table (use `geom_col()`).

## Two categorical variables at once

A common need is to show counts for the combinations of *two* categorical variables. For example, how many participants of each gender are in each experimental condition. There are three standard ways to display this: **stacked**, **dodged** (side-by-side), and **faceted**. Each emphasizes a different comparison.

**Stacked bars.** All counts within a condition are stacked on top of one another. The total bar height equals the total number of participants in the condition; the colored segments show how that total breaks down by gender.

```{r}
#| label: fig-stacked
#| fig-width: 5.5
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = condition, fill = gender)) +
  geom_bar() +
  labs(x = "Condition", y = "Number of participants",
       fill = "Gender") +
  theme_minimal()
```

Stacked bars emphasize the totals (the full bar heights) and the relative composition within each total. They are useful when the total is the primary quantity of interest and the breakdown is secondary.

**Dodged (side-by-side) bars.** Each gender gets its own bar within each condition, placed side by side.

```{r}
#| label: fig-dodged
#| fig-width: 6
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = condition, fill = gender)) +
  geom_bar(position = position_dodge(preserve = "single")) +
  labs(x = "Condition", y = "Number of participants",
       fill = "Gender") +
  theme_minimal()
```

Dodged bars emphasize the comparison of gender counts within each condition, because each gender's bar starts from the x-axis and its height is directly readable. They are usually a better choice than stacked bars when the comparison of interest is the individual counts (not the totals).

**Faceted bars.** Each condition gets its own panel, with the gender breakdown shown inside:

```{r}
#| label: fig-faceted-bar
#| fig-width: 7
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = gender)) +
  geom_bar(fill = "steelblue") +
  facet_wrap(~ condition) +
  labs(x = "Gender", y = "Number of participants") +
  theme_minimal()
```

Faceted bars work well when neither variable is clearly secondary and the reader will want to look at the within-panel distribution for each level. They use more space than stacked or dodged bars, but they are the easiest to read when each panel matters in its own right.

The choice between stacked, dodged, and faceted depends on which comparison is the message:

- Stacked: emphasizes the *total*, with composition as a secondary detail.
- Dodged: emphasizes the comparison of *categories within each x value*.
- Faceted: emphasizes each subset on its own, at the cost of more vertical space.

When in doubt, use dodged. It is the format that supports the most direct visual comparison and has the fewest perceptual pitfalls.

## Counts versus proportions

Raw counts are most useful when the total sample size is itself meaningful (for example, when you want to show that there were 63 women and 46 men in the study). Proportions are more useful when you want to compare *rates* across groups that have different totals (for example, when the violent condition has more total participants than the neutral condition, so a raw count of any subgroup is partially driven by the difference in totals).

To switch from counts to proportions in a stacked bar, set `position = "fill"` and the bars become normalized to 100%:

```{r}
#| label: fig-proportions
#| fig-width: 5.5
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = condition, fill = gender)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Condition", y = "Proportion",
       fill = "Gender") +
  theme_minimal()
```

Each bar now goes from 0% to 100%, and the segments show the share of each gender within the condition. This is the right choice when the comparison is "is the proportion of women in the violent condition different from the proportion of women in the neutral condition?" rather than "how many women are in each condition?".

The downside of the proportion plot is that the *total counts* are now invisible. A subgroup that contains 1 person out of 5 (20%) looks identical to a subgroup that contains 100 people out of 500 (20%). When the difference in totals matters, either keep the raw-count version or add the total counts as a text annotation.

## Ordering categories meaningfully

R's default for character (and unordered factor) variables is alphabetical order. This is rarely the order you want for a figure. Three common alternatives are worth knowing about.

**Order by count (most to least, or vice versa).** This is the right default for most exploratory figures, because the eye naturally lands on the largest category first.

```{r}
#| label: fig-ordered-count
#| fig-width: 5.5
#| fig-height: 3
#| fig-align: center

ggplot(vgames, aes(x = forcats::fct_infreq(gender))) +
  geom_bar(fill = "steelblue") +
  labs(x = "Gender", y = "Number of participants") +
  theme_minimal()
```

`forcats::fct_infreq()` reorders the factor by frequency, with the most common level first. Add `fct_rev()` around the call to reverse the order (least common first).

**Order by an explicit ordering.** When the categories have a natural order (Strongly Disagree → Strongly Agree, or "control" before "treatment"), specify the order explicitly with `forcats::fct_relevel()`.

```{r}
#| label: fig-ordered-explicit
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

vgames$condition_ordered <- forcats::fct_relevel(vgames$condition,
                                                 "neutral", "violent")

ggplot(vgames, aes(x = condition_ordered)) +
  geom_bar(fill = "steelblue", width = 0.6) +
  labs(x = "Condition", y = "Number of participants") +
  theme_minimal()
```

**Order by a different variable.** Sometimes you want categories ordered by a third quantity, such as the mean of the outcome. `forcats::fct_reorder()` does this:

```{r}
#| label: fig-ordered-by
#| fig-width: 5.5
#| fig-height: 3
#| fig-align: center

vgames$gender_by_mean <- forcats::fct_reorder(vgames$gender,
                                              vgames$aggression,
                                              .fun = mean)

ggplot(vgames, aes(x = gender_by_mean)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Gender (ordered by mean aggression)",
       y = "Number of participants") +
  theme_minimal()
```

Whichever ordering you use, name it explicitly in the axis label so the reader is not left guessing why the categories are in that order.

## Cleveland dot plots when there are many categories

When a bar chart has more than about six or seven categories, especially with long category names, it starts to feel cluttered. A useful alternative is the **Cleveland dot plot**: a horizontal layout in which each category is a row and a single point shows the count. The format is named after the statistician William Cleveland and is widely considered easier to read than a vertical bar chart for many categories ([Cleveland, 1985](https://doi.org/10.1080/01621459.1984.10478080)).

Suppose we bin participants by age into integer groups:

```{r}
#| label: fig-cleveland
#| fig-width: 6
#| fig-height: 4.5
#| fig-align: center

age_counts <- vgames |>
  count(age) |>
  arrange(n)

ggplot(age_counts, aes(x = n, y = forcats::fct_reorder(factor(age), n))) +
  geom_segment(aes(xend = 0, yend = forcats::fct_reorder(factor(age), n)),
               color = "grey70") +
  geom_point(color = "steelblue", size = 3) +
  labs(x = "Number of participants", y = "Age") +
  theme_minimal()
```

The dot is the count for each age. The horizontal segment from zero to the dot makes the count immediately visually evident. The category labels are on the y-axis, which means long labels do not overlap (a common problem in vertical bar charts when category names are long). The categories are sorted by count, which makes the figure easy to scan.

A Cleveland dot plot is essentially a horizontal bar chart with the bar replaced by a segment-plus-point. The information content is the same as a bar chart, but the layout scales better to many categories. For five or fewer categories, a vertical bar chart is fine. For more, switch.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "Bar charts are an appropriate visualization for both means and counts, because the bar height is just a numerical quantity and the choice between mean and count is a matter of preference rather than principle.",
  answer = "Bar charts are an appropriate visualization for counts because the bar height represents a real countable quantity (number of observations in each category), whereas for means the bar height represents a summary that throws away the distribution and is often misleading.",
  "Bar charts are an appropriate visualization for means but not for counts, because the count of observations in each category is not itself a continuous quantity and cannot be properly represented by the height of a bar.",
  "Bar charts are an appropriate visualization for counts only when the categorical variable has exactly two levels; for three or more levels, a pie chart should be used instead because the parts add up to a known whole."
)

opts_q2 <- c(
  "geom_bar() and geom_col() are interchangeable functions for the same visualization and the choice between them is purely stylistic, with geom_bar() being preferred in older code and geom_col() in newer code.",
  answer = "geom_bar() computes the count of observations in each category automatically from the raw data, while geom_col() plots a y-value that you supply, so the choice depends on whether you are starting from raw observations or from a pre-computed summary table.",
  "geom_bar() is restricted to vertical bar charts while geom_col() is restricted to horizontal bar charts, and the rest of the syntax (aesthetic mappings, fill colors, etc.) is otherwise identical between the two functions.",
  "geom_bar() requires a continuous variable on the x-axis while geom_col() requires a categorical variable, and they cannot be used interchangeably for the same kind of plot in ggplot2."
)

opts_q3 <- c(
  "Stacked bars are always preferable to dodged bars because they take up less horizontal space and make the total counts easier to read, with the breakdown by sub-category appearing automatically inside each bar.",
  "Dodged bars are always preferable to stacked bars because they place each sub-category on the x-axis baseline, which is the only way to make the heights of individual sub-categories directly comparable across groups.",
  answer = "Stacked bars emphasize the totals (the full bar height) with the breakdown as a secondary detail, while dodged bars emphasize the comparison of sub-categories within each x value because each sub-category starts from the axis; faceted bars give each panel its own subset at the cost of using more vertical space.",
  "Stacked, dodged, and faceted bars all show the same information in the same visual ranking of comparisons, and the choice between them is primarily a matter of personal preference rather than which comparison is being emphasized."
)

opts_q4 <- c(
  "Switching from raw counts to proportions with position = 'fill' always preserves the relative ranking of sub-categories within each group, so the choice between counts and proportions is a matter of axis labeling only.",
  answer = "Switching from raw counts to proportions with position = 'fill' normalizes each bar to 100%, which is appropriate when you want to compare rates across groups of different sizes, but it hides the underlying total counts, which can be misleading when the difference in totals between groups is itself an important part of the story.",
  "Switching from raw counts to proportions with position = 'fill' produces a visualization that is mathematically equivalent to a pie chart for each group, so there is no perceptual or interpretive advantage to using the bar format with this position setting.",
  "Switching from raw counts to proportions with position = 'fill' should be done only when the total sample size in each group is exactly equal, otherwise the resulting figure will misrepresent the relative sizes of the sub-categories."
)

opts_q5 <- c(
  "Categorical variables in R are always automatically ordered by the mean of an associated outcome variable when used in a ggplot2 figure, so the user does not need to specify an ordering manually.",
  "Categorical variables in R should always be ordered alphabetically in figures because alphabetical order is the most consistent across studies and journals and supports cross-study comparison.",
  answer = "Categorical variables in R default to alphabetical order, which is rarely the right order for a figure; forcats::fct_infreq() orders levels by frequency, fct_relevel() applies a user-specified order, and fct_reorder() orders levels by a third variable such as a group mean.",
  "Categorical variables in R should not be reordered for a figure because changing the order of the levels alters the underlying meaning of the categorical variable and invalidates any statistical analysis based on it."
)

opts_q6 <- c(
  "Cleveland dot plots are inappropriate visualizations of count data because they require a continuous x-axis, and the discrete count of observations in each category cannot be properly displayed in this format.",
  answer = "Cleveland dot plots show one row per category with a horizontal segment and dot representing the count, scaling better than vertical bar charts when there are many categories or long category labels and remaining easy to read across a sorted ordering.",
  "Cleveland dot plots are equivalent to box plots that show only the median value as a single point, and are used in situations where the spread of the data is not of interest to the reader.",
  "Cleveland dot plots are an outdated visualization style superseded by stacked and dodged bar charts in modern ggplot2 workflows, and should be avoided in favor of these more contemporary alternatives."
)
```

**1. Why are bar charts considered an appropriate visualization for counts but not for means?**

`r longmcq(opts_q1)`

**2. What is the difference between `geom_bar()` and `geom_col()` in `ggplot2`?**

`r longmcq(opts_q2)`

**3. When showing two categorical variables together, what is the difference between stacked, dodged, and faceted bar charts?**

`r longmcq(opts_q3)`

**4. What is the trade-off when switching from raw counts to proportions with `position = "fill"`?**

`r longmcq(opts_q4)`

**5. How should categorical variables be ordered in a `ggplot2` figure, and what tools are available for ordering them meaningfully?**

`r longmcq(opts_q5)`

**6. What is a Cleveland dot plot and when is it preferable to a vertical bar chart?**

`r longmcq(opts_q6)`

## References

Cleveland, W. S. (1985). *The elements of graphing data*. Wadsworth.

Cleveland, W. S., & McGill, R. (1984). Graphical perception: Theory, experimentation, and application to the development of graphical methods. *Journal of the American Statistical Association, 79*(387), 531-554. <https://doi.org/10.1080/01621459.1984.10478080>
