---
title: "Visualizing distributions"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
library(tidyr)
```

This chapter is about visualizing the *shape* of a single continuous variable. Where Chapter 31 focused on showing means and spread, this chapter focuses on showing the whole distribution: where the values cluster, where they thin out, whether they are symmetric or skewed, whether there are multiple modes, whether outliers sit far from the central mass.

Distribution plots are useful for several reasons. They reveal features that summary statistics like the mean and SD hide. They help diagnose whether a particular analysis is appropriate (a t-test on a heavily skewed outcome is suspect, for instance). They communicate the underlying data to the reader in a way that bar plots and means cannot. And, as we will see in the final section, the visual comparison of two overlapping distributions is a direct intuition pump for standardized effect sizes like Cohen's *d*.

We use the `vgames_study` data throughout.

```{r}
#| label: vgames-load
#| eval: false
vgames <- read.csv("data/vgames_study.csv")
```

```{r}
#| label: vgames-data
#| echo: false
vgames <- tryCatch(
  read.csv("data/vgames_study.csv"),
  error = function(e) {
    set.seed(1)
    n <- 60
    data.frame(
      participant_id  = seq_len(2 * n),
      condition       = rep(c("violent", "neutral"), each = n),
      hours_weekly    = c(rnorm(n, 8, 4), rnorm(n, 7, 4)),
      trait_hostility = rnorm(2 * n, mean = 4, sd = 1.2),
      empathy         = rnorm(2 * n, mean = 4.5, sd = 1.1),
      aggression      = c(rnorm(n, 5.56, 1.6), rnorm(n, 4.21, 1.6))
    )
  }
)
```

## Histograms

The most basic distribution plot is the **histogram**. It divides the range of the variable into bins of equal width, counts the observations in each bin, and shows the counts as adjacent bars.

```{r}
#| label: fig-histogram
#| fig-width: 6
#| fig-height: 3.2
#| fig-align: center

ggplot(vgames, aes(x = aggression)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue",
                 color = "white") +
  labs(x = "Aggression", y = "Number of participants") +
  theme_minimal()
```

The choice of bin width matters. Too few bins (a wide `binwidth`) smooths over real features of the distribution. Too many bins (a narrow `binwidth`) shows random noise as if it were structure. The figure below shows the same data with three different bin widths.

```{r}
#| label: fig-binwidth
#| fig-width: 9
#| fig-height: 3
#| fig-align: center

p_wide <- ggplot(vgames, aes(x = aggression)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 2 (too coarse)", x = NULL, y = NULL) +
  theme_minimal(base_size = 9)

p_good <- ggplot(vgames, aes(x = aggression)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 0.5 (reasonable)", x = NULL, y = NULL) +
  theme_minimal(base_size = 9)

p_narrow <- ggplot(vgames, aes(x = aggression)) +
  geom_histogram(binwidth = 0.1, fill = "steelblue", color = "white") +
  labs(title = "binwidth = 0.1 (too narrow)", x = NULL, y = NULL) +
  theme_minimal(base_size = 9)

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_wide, p_good, p_narrow, ncol = 3)
} else {
  gridExtra::grid.arrange(p_wide, p_good, p_narrow, ncol = 3)
}
```

A useful rule of thumb is to start with a `binwidth` that places about 20 to 30 bars across the range of the data, then adjust by eye. There is no objectively correct bin width; the right choice depends on what features of the distribution you want to communicate.

The y-axis of a histogram is "count" by default. Use `aes(y = after_stat(density))` to switch to a *density* y-axis, which scales the histogram so that the total area sums to 1. This is the right choice when you want to overlay theoretical distributions or compare histograms across datasets with different sample sizes.

```{r}
#| label: fig-histogram-density
#| fig-width: 6
#| fig-height: 3.2
#| fig-align: center

ggplot(vgames, aes(x = aggression, y = after_stat(density))) +
  geom_histogram(binwidth = 0.5, fill = "steelblue",
                 color = "white") +
  labs(x = "Aggression", y = "Density") +
  theme_minimal()
```

The shape is identical to the count histogram; only the y-axis scaling has changed.

## Density plots

A **density plot** is a smoothed version of a histogram. Instead of binning into discrete bars, it estimates a continuous curve that represents the underlying density of the data. The function `geom_density()` produces it.

```{r}
#| label: fig-density
#| fig-width: 6
#| fig-height: 3.2
#| fig-align: center

ggplot(vgames, aes(x = aggression)) +
  geom_density(fill = "steelblue", alpha = 0.4,
               color = "steelblue4", linewidth = 0.7) +
  labs(x = "Aggression", y = "Density") +
  theme_minimal()
```

Density plots are useful when the histogram's binning starts to feel arbitrary. They show the general shape (peaks, tails, skew) without the visual noise of discrete bars. The trade-off is that the smoothing involves a choice (the bandwidth, controlled by `bw` or `adjust`), and a poor bandwidth choice can either oversmooth (hiding real features) or undersmooth (showing noise as if it were structure). Same trade-off as the histogram, just with a different name for the parameter.

The clearest situation in which density beats histogram is when you want to overlay multiple groups on the same plot. Histograms get visually crowded when stacked; densities can sit on top of each other with transparency without losing clarity.

## Overlapping density plots

To compare the distributions of a variable across two or more groups, plot the density for each group on the same axes, with semi-transparent fills so both are visible where they overlap.

```{r}
#| label: fig-overlap-density
#| fig-width: 6
#| fig-height: 3.4
#| fig-align: center

ggplot(vgames, aes(x = aggression, fill = condition,
                   color = condition)) +
  geom_density(alpha = 0.35, linewidth = 0.7) +
  labs(x = "Aggression", y = "Density",
       fill = "Condition", color = "Condition") +
  theme_minimal()
```

The visual story is immediately readable: the violent-condition density (one color) is shifted to the right compared to the neutral-condition density (the other color). The shaded overlap shows where the two distributions cover the same values; the non-overlapping tails show where each distribution has data the other does not.

Overlapping density plots are one of the most informative visualizations available for comparing two groups. They show the means (where each curve peaks), the spread (how wide each curve is), the shape (symmetric, skewed, bimodal), and the overlap (how distinguishable the two groups are). Five summary statistics are visible at a glance in one figure.

The same approach works for more than two groups, with one important caveat: if there are too many groups, the layered densities become a mess. Three or four groups is usually the practical maximum for overlapping densities. For more, switch to ridgeline plots (next section).

## Ridgeline plots

A **ridgeline plot** (also called a "joyplot", in reference to the album cover for Joy Division's *Unknown Pleasures*) shows multiple distributions stacked vertically, with each distribution offset from the next so they can all be seen simultaneously without obscuring each other. It is the right choice when there are many groups to compare.

Ridgeline plots are produced by the `ggridges` package. Let us first construct an illustrative grouping: bin participants by age into four categories.

```{r}
#| label: fig-ridges
#| fig-width: 6.5
#| fig-height: 3.8
#| fig-align: center

vgames$age_bin <- cut(vgames$age,
                      breaks = c(17, 22, 26, 30, 36),
                      labels = c("18-22", "23-26",
                                 "27-30", "31-35"))

if (requireNamespace("ggridges", quietly = TRUE)) {
  library(ggridges)
  ggplot(vgames, aes(x = aggression, y = age_bin,
                     fill = age_bin)) +
    geom_density_ridges(alpha = 0.6, color = "white",
                        scale = 1.1, rel_min_height = 0.01) +
    labs(x = "Aggression", y = "Age band",
         fill = NULL) +
    theme_minimal() +
    theme(legend.position = "none")
} else {
  cat("(ggridges not installed; on your machine the figure would show\n")
  cat("four overlapping density curves, one per age band, stacked vertically.)\n")
}
```

Each band of the figure is a density estimate of aggression for participants of that age. Stacking them vertically lets the eye scan from top to bottom and quickly compare how the distribution changes across categories. A clean ridgeline plot can replace five or six small-multiples panels in a single visualization.

The `scale` argument controls how much the densities overlap vertically; values larger than 1 produce more overlap, smaller values less. The `rel_min_height` argument hides density tails below a threshold to clean up the plot.

Ridgeline plots are a relatively recent addition to the standard data-visualization toolkit, popularized by Allen, Poggiali, Whitaker, Marshall, van Langen, and Kievit ([2021](https://doi.org/10.12688/wellcomeopenres.15191.2)) as part of a broader "raincloud" plot family that combines density, boxplot, and individual-point elements.

## Visualizing standardized effect sizes with overlapping densities

Overlapping density plots are not just descriptive tools; they are the most intuitive visualization available for the *standardized* effect sizes we have used throughout Section 4. **Cohen's *d* is essentially a measure of how far apart two distributions are**, expressed in standard-deviation units. The visual fingerprint of *d* is exactly how separated two overlapping densities look.

The figure below shows the same two normal distributions, separated by Cohen's *d* values of 0.2, 0.5, 0.8, 1.5, and 2.0.

```{r}
#| label: fig-d-grid
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

d_vals <- c(0.2, 0.5, 0.8, 1.5, 2.0)

# Build six panels of two-distribution data
build_panel <- function(d) {
  x <- seq(-4, 4 + d, length.out = 400)
  rbind(
    data.frame(x = x, density = dnorm(x, mean = 0, sd = 1),
               group = "Group A"),
    data.frame(x = x, density = dnorm(x, mean = d, sd = 1),
               group = "Group B")
  ) |>
    mutate(facet = sprintf("d = %.1f", d))
}

grid_df <- do.call(rbind, lapply(d_vals, build_panel))
grid_df$facet <- factor(grid_df$facet,
                        levels = sprintf("d = %.1f", d_vals))

ggplot(grid_df, aes(x = x, y = density, fill = group, color = group)) +
  geom_area(alpha = 0.35, position = "identity") +
  geom_line(linewidth = 0.5) +
  facet_wrap(~ facet, ncol = 5) +
  scale_fill_manual(values = c("Group A" = "steelblue",
                               "Group B" = "darkorange")) +
  scale_color_manual(values = c("Group A" = "steelblue4",
                                "Group B" = "darkorange3")) +
  labs(x = NULL, y = NULL,
       fill = NULL, color = NULL) +
  theme_minimal(base_size = 9.5) +
  theme(legend.position = "top",
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

Reading the figure from left to right: at *d* = 0.2 the two distributions are almost identical; at *d* = 0.5 they are visibly distinct but heavily overlapping; at *d* = 0.8 the overlap is meaningful but the means are clearly separated; at *d* = 1.5 the overlap is much reduced; at *d* = 2.0 the distributions are essentially separated, with only a small overlap in the tails. Memorizing this grid is one of the most useful effect-size intuitions a researcher can develop. When a paper reports *d* = 0.6, you should be able to picture the corresponding amount of overlap without doing any calculation.

The arrow annotation below shows the same idea on a single panel: Cohen's *d* is the distance between the means, measured in standard-deviation units.

```{r}
#| label: fig-d-annotated
#| fig-width: 6.5
#| fig-height: 3.5
#| fig-align: center

x_d <- seq(-4, 5, length.out = 400)
df_d <- rbind(
  data.frame(x = x_d, density = dnorm(x_d, mean = 0, sd = 1),
             group = "Group A"),
  data.frame(x = x_d, density = dnorm(x_d, mean = 0.8, sd = 1),
             group = "Group B")
)

ggplot(df_d, aes(x = x, y = density,
                 fill = group, color = group)) +
  geom_area(alpha = 0.35, position = "identity") +
  geom_line(linewidth = 0.5) +
  geom_vline(xintercept = c(0, 0.8), linetype = "dashed",
             color = "grey40", linewidth = 0.4) +
  annotate("segment", x = 0, xend = 0.8,
           y = 0.45, yend = 0.45,
           arrow = grid::arrow(ends = "both",
                               length = unit(0.15, "cm")),
           color = "grey20") +
  annotate("text", x = 0.4, y = 0.49,
           label = "d = 0.8 SDs",
           size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("Group A" = "steelblue",
                               "Group B" = "darkorange")) +
  scale_color_manual(values = c("Group A" = "steelblue4",
                                "Group B" = "darkorange3")) +
  labs(x = "Outcome", y = "Density",
       fill = NULL, color = NULL) +
  coord_cartesian(ylim = c(0, 0.55)) +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid.minor = element_blank(),
        axis.text.y = element_blank())
```

The dashed vertical lines mark the means of the two distributions. The horizontal arrow between them, in standard-deviation units, *is* Cohen's *d*. The shaded overlap is the part of the two distributions that occupy the same range of outcome values. Pieces of intuition that fall directly out of this picture:

- A small Cohen's *d* means a small distance between the means relative to the SD. The means are close compared to the spread of either group.
- A large Cohen's *d* means the opposite. The means are well separated relative to the spread of either group.
- The amount of overlap is not the same as the amount of variance explained, but is closely related to it. Roughly, overlap shrinks as *d* grows.
- Even at *d* = 2.0, the distributions still overlap. Individual observations from the lower-mean group can exceed the mean of the higher-mean group, just less often. This is why standardized effect sizes do not predict individual-level differences with certainty (recall the discussion in Chapter 27, section 27.5).

When you see Cohen's *d* in a paper from now on, look at the grid in figure 33.5 first, find the closest value, and that picture is what the data look like. The figure does much of the interpretive work for you.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "The choice of binwidth in a histogram is purely cosmetic and has no effect on the visual interpretation of the underlying distribution, because the same data are being displayed regardless of how many bars are shown.",
  answer = "The choice of binwidth in a histogram affects whether real features of the distribution are visible or hidden: too-wide bins smooth over genuine structure, while too-narrow bins display random noise as if it were structure, and the right choice depends on what the figure is meant to communicate.",
  "The choice of binwidth in a histogram is determined automatically by ggplot2 using a default algorithm that always produces the optimal representation of the data, so users should never specify the binwidth manually.",
  "The choice of binwidth in a histogram is constrained by the sample size and cannot be adjusted independently, because the formula for binwidth selection includes the number of observations as a fixed input."
)

opts_q2 <- c(
  "A density plot is always a more accurate representation of the underlying data than a histogram because the smoothing procedure removes random sampling noise and reveals the true population distribution from which the sample was drawn.",
  answer = "A density plot replaces the discrete bars of a histogram with a continuous smoothed curve, which is useful for showing general shape without the visual noise of bars; the smoothing involves a bandwidth choice that has the same kind of trade-off (oversmoothing vs undersmoothing) as the binwidth choice in a histogram.",
  "A density plot and a histogram are mathematically equivalent representations of the same data, with the density plot using a continuous y-axis that always integrates to 1 and the histogram using a discrete y-axis of integer counts.",
  "A density plot can only be used for normally distributed data because the smoothing procedure used by geom_density() assumes a Gaussian kernel and produces misleading results for skewed or multimodal distributions."
)

opts_q3 <- c(
  "Overlapping density plots are appropriate only when the groups are perfectly normally distributed with equal variances, otherwise the overlap area is not interpretable as a measure of similarity between the groups.",
  answer = "Overlapping density plots show the means, spreads, shapes, and overlap of two or more groups in a single figure, making them one of the most informative comparisons available, with three or four groups being the practical maximum before the layered curves become visually unreadable.",
  "Overlapping density plots can only display two groups at a time, because adding a third density would require a separate axis and would invalidate the visual comparison of the first two distributions.",
  "Overlapping density plots are equivalent to side-by-side box plots and convey exactly the same information; the choice between them is a matter of personal preference rather than the kind of insight the figure provides."
)

opts_q4 <- c(
  "Ridgeline plots are an outdated form of distributional visualization that has been superseded by the more efficient overlapping density plot in modern ggplot2 workflows.",
  answer = "Ridgeline plots stack multiple density estimates vertically with each distribution offset from the next, allowing many groups to be compared simultaneously without the visual crowding that overlapping density plots suffer when more than three or four groups are shown.",
  "Ridgeline plots show one density per category but require all categories to have the same sample size, otherwise the relative heights of the densities cannot be compared across rows.",
  "Ridgeline plots display the cumulative distribution function rather than the density, so the interpretation of any individual ridge corresponds to the proportion of observations below each value rather than the density at that value."
)

opts_q5 <- c(
  "Cohen's d is computed from the area of overlap between two distributions, so a Cohen's d of 0.5 corresponds to two distributions whose total overlapping area is 50%.",
  answer = "Cohen's d is the distance between two means expressed in standard-deviation units, and the visual fingerprint of d is exactly how separated two overlapping distributions look; the grid of d values from 0.2 to 2.0 is a useful intuition pump for what each magnitude implies about overlap.",
  "Cohen's d measures the difference between two means in raw units of the outcome variable, so a Cohen's d of 0.5 always corresponds to a half-point difference on whatever scale the outcome is measured.",
  "Cohen's d is meaningful only for normally distributed outcomes with equal variances, and is undefined for the overlapping-density visualization when the two groups have different shapes or different spreads."
)

opts_q6 <- c(
  "Even when Cohen's d is as large as 2.0, the two distributions still show meaningful overlap in their tails, meaning that individuals from the lower-mean group can still exceed the mean of the higher-mean group, just less often than in the comparison group.",
  "Even when Cohen's d is as large as 2.0, the two distributions never overlap at all, because a d of 2.0 implies a separation that is too large for any individual observations to fall in the same range across groups.",
  answer = "Standardized effect sizes describe a population-level statistical difference between groups, and even substantial values such as Cohen's d = 2.0 leave meaningful overlap between the two distributions; individual observations from the lower-mean group can still exceed the mean of the higher-mean group at lower rates than the comparison group, so standardized effects do not predict individual differences with certainty.",
  "Standardized effect sizes such as Cohen's d are misleading when computed from samples and should be replaced by raw-unit mean differences in any visualization, because the standardization step destroys the connection between the figure and the underlying scale of the outcome."
)
```

**1. How does the choice of bin width affect a histogram, and what is a reasonable default?**

`r longmcq(opts_q1)`

**2. What is the relationship between a density plot and a histogram?**

`r longmcq(opts_q2)`

**3. What information does an overlapping density plot of two groups convey, and what is its practical limit on the number of groups?**

`r longmcq(opts_q3)`

**4. What does a ridgeline plot show and when is it preferable to an overlapping density plot?**

`r longmcq(opts_q4)`

**5. How does Cohen's *d* relate to the visual overlap of two density distributions?**

`r longmcq(opts_q5)`

**6. What does the persistent overlap between two distributions at large values of Cohen's *d* (such as *d* = 2.0) tell us about the interpretation of standardized effect sizes?**

`r longmcq(opts_q6)`

## References

Allen, M., Poggiali, D., Whitaker, K., Marshall, T. R., van Langen, J., & Kievit, R. A. (2021). Raincloud plots: A multi-platform tool for robust data visualization. *Wellcome Open Research, 4*, 63. <https://doi.org/10.12688/wellcomeopenres.15191.2>

Wilke, C. O. (2024). *ggridges: Ridgeline plots in ggplot2* [R package]. <https://CRAN.R-project.org/package=ggridges>
