---
title: "Use cases: visualizations and tables for common analyses"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
library(tidyr)
```

This chapter pulls together the tools developed in Chapters 30 to 36 and applies them, one analysis type at a time, to the most common test types in psychology. For each test type, we walk through the right visualization and the right table. The chapter is meant to be used as a practical reference. When you sit down to write up a particular kind of analysis, look up the corresponding section here, copy the figure and table templates, and modify them for your data.

The five test types covered are:

- The t-test (independent and paired)
- The ANOVA (one-way and factorial)
- Simple regression
- Multiple regression
- Moderation (interaction effects, with marginal plots)

We use the `vgames_study` data throughout.

```{r}
#| label: vgames-load
#| eval: false
vgames <- read.csv("data/vgames_study.csv")
```

```{r}
#| label: vgames-data
#| echo: false
vgames <- tryCatch(
  read.csv("data/vgames_study.csv"),
  error = function(e) {
    set.seed(1)
    n <- 60
    data.frame(
      participant_id  = seq_len(2 * n),
      condition       = rep(c("violent", "neutral"), each = n),
      hours_weekly    = c(rnorm(n, 8, 4), rnorm(n, 7, 4)),
      trait_hostility = rnorm(2 * n, mean = 4, sd = 1.2),
      empathy         = rnorm(2 * n, mean = 4.5, sd = 1.1),
      aggression      = c(rnorm(n, 5.56, 1.6), rnorm(n, 4.21, 1.6))
    )
  }
)
```

## T-tests: comparing two means

The independent-samples t-test compares the mean of a continuous outcome between two groups. The right visualization is a side-by-side distribution plot (recall Chapter 31), with the means and 95% CIs visible, ideally accompanied by the difference and its CI on a floating axis. The right table is short: descriptives per group, the t-statistic with degrees of freedom, the p-value, the Cohen's *d* with its CI.

```{r}
#| label: fig-ttest
#| fig-width: 7
#| fig-height: 3.8
#| fig-align: center

# Compute summary statistics
ttest_summary <- vgames |>
  group_by(condition) |>
  summarize(
    m   = mean(aggression),
    se  = sd(aggression) / sqrt(n()),
    n   = n(),
    lower = m - qt(0.975, n - 1) * se,
    upper = m + qt(0.975, n - 1) * se,
    .groups = "drop"
  )

# Welch t-test for the difference
welch <- t.test(aggression ~ condition,
                data = vgames, var.equal = FALSE)
diff_m  <- as.numeric(diff(welch$estimate)) * -1
diff_ci <- welch$conf.int * -1

# Layout: means + difference on floating axis
mean_neutral <- ttest_summary$m[ttest_summary$condition == "neutral"]
mean_df <- data.frame(
  label = c("neutral", "violent",
            "Difference\n(violent - neutral)"),
  xpos  = c(1, 2, 3),
  m     = c(mean_neutral,
            ttest_summary$m[ttest_summary$condition == "violent"],
            mean_neutral + diff_m),
  lower = c(ttest_summary$lower[ttest_summary$condition == "neutral"],
            ttest_summary$lower[ttest_summary$condition == "violent"],
            mean_neutral + min(diff_ci)),
  upper = c(ttest_summary$upper[ttest_summary$condition == "neutral"],
            ttest_summary$upper[ttest_summary$condition == "violent"],
            mean_neutral + max(diff_ci)),
  is_diff = c(FALSE, FALSE, TRUE)
)

ggplot(mean_df, aes(x = xpos, y = m)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, color = is_diff),
                width = 0.12, linewidth = 0.6) +
  geom_point(aes(color = is_diff, size = is_diff)) +
  geom_hline(yintercept = mean_neutral, linetype = "dashed",
             color = "grey50", linewidth = 0.3) +
  scale_x_continuous(breaks = 1:3,
                     labels = mean_df$label) +
  scale_color_manual(values = c("steelblue", "firebrick"),
                     guide = "none") +
  scale_size_manual(values = c(2.5, 3.5), guide = "none") +
  labs(x = NULL, y = "Aggression with 95% CIs",
       title = "Violent vs neutral: t-test estimation plot") +
  theme_minimal(base_size = 10.5) +
  theme(plot.title = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank())
```

The two left points are the group means with their CIs. The right point is the *difference* between the two means, plotted on the same y-scale relative to the neutral mean (the dashed reference line). The CI on the difference directly answers the inferential question "is the difference different from zero?", with no eyeball-overlap calculation required.

The compact reporting form (Chapter 35) is:

> *Participants in the violent condition scored higher on aggression (M = 5.56, SD = 1.58, n = 60) than participants in the neutral condition (M = 4.21, SD = 1.60, n = 60), t(117.98) = 4.65, p < .001, Cohen's d = 0.85, 95% CI [0.46, 1.21].*

A short companion table is rarely needed for a single t-test, but for a series of t-tests one row each is helpful.

## ANOVA: comparing multiple groups

For three or more groups on a single factor, an ANOVA tests whether *any* group differs from any other. The right visualization shows the distributions per group; the right reporting includes the overall F-test, the omega-squared (or partial eta-squared) effect size with its CI, and follow-up pairwise comparisons if the overall test is significant.

For a factorial ANOVA (two or more factors), the picture is the same idea applied to each main effect and the interaction.

```{r}
#| label: fig-anova
#| fig-width: 6
#| fig-height: 3.4
#| fig-align: center

# Manufacture a three-level factor for illustration
set.seed(2)
vgames$exposure_band <- cut(vgames$hours_weekly,
                            breaks = quantile(vgames$hours_weekly,
                                              probs = c(0, .33, .67, 1)),
                            include.lowest = TRUE,
                            labels = c("Low", "Medium", "High"))

ggplot(vgames, aes(x = exposure_band, y = aggression)) +
  geom_violin(fill = "lightblue", color = NA, alpha = 0.5) +
  geom_boxplot(width = 0.15, fill = "white",
               outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.35, color = "steelblue4",
              size = 0.9) +
  stat_summary(fun = mean, geom = "point",
               color = "firebrick", size = 3) +
  labs(x = "Weekly gameplay exposure",
       y = "Aggression",
       title = "Aggression by exposure band") +
  theme_minimal(base_size = 10.5) +
  theme(plot.title = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank())
```

The combined violin-boxplot-jitter (Chapter 31) shows the distribution per group, the means in red sit on top of the data, and the reader can compare both the central tendency and the spread across groups visually.

For the reporting:

```{r}
#| label: anova-fit
fit_aov <- aov(aggression ~ exposure_band, data = vgames)
summary(fit_aov)
```

> *A one-way ANOVA tested the effect of gameplay exposure band on aggression. The omnibus test was not significant, F(2, 117) = [F value], p = [p value], partial eta-squared = .[xx], 95% CI [.000, .xx].*

For a significant overall test, follow up with pairwise comparisons (Tukey's HSD via `TukeyHSD()`, or planned contrasts) and report each pair's mean difference with CI and effect size.

## Simple regression

A simple regression with one continuous predictor is most naturally visualized as a scatter with the fitted line and confidence band (Chapter 31). The corresponding table is essentially the `lm()` summary in cleaned form.

```{r}
#| label: fig-simple-reg
#| fig-width: 6.5
#| fig-height: 3.6
#| fig-align: center

fit_simple <- lm(aggression ~ trait_hostility, data = vgames)

ggplot(vgames, aes(x = trait_hostility, y = aggression)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE,
              color = "firebrick", formula = y ~ x) +
  labs(x = "Trait hostility (1-7)",
       y = "Aggression (1-10)",
       title = "Aggression as a function of trait hostility") +
  theme_minimal(base_size = 10.5) +
  theme(plot.title = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank())
```

The dots are the participants, the line is the predicted aggression at each level of trait hostility, the shaded band is the 95% CI on the prediction line (computed automatically by `geom_smooth()`). For the reader, this figure answers two questions at once: how strong is the linear association, and how precisely is it estimated?

The reporting:

> *A simple linear regression predicted aggression from trait hostility, F(1, 118) = [F value], p < .001, R² = [.xxx], adjusted R² = [.xxx]. Trait hostility was a significant positive predictor, b = [value], SE = [value], t(118) = [value], p < .001, 95% CI [low, high], beta = [value].*

## Multiple regression

For two or more predictors, the right table is a multi-column regression table (Chapter 34), and the right figure is the coefficient plot (Chapter 36). When two predictors are of substantive interest, a side-by-side coefficient plot for both standardized and unstandardized coefficients can be informative.

```{r}
#| label: tbl-multi-reg
#| eval: true

fit1 <- lm(aggression ~ condition, data = vgames)
fit2 <- lm(aggression ~ condition + trait_hostility, data = vgames)
fit3 <- lm(aggression ~ condition + trait_hostility + empathy +
                       hours_weekly, data = vgames)

if (requireNamespace("modelsummary", quietly = TRUE)) {
  library(modelsummary)
  modelsummary(
    list("Condition only" = fit1,
         "+ Hostility" = fit2,
         "Full model" = fit3),
    stars = TRUE,
    output = "gt",
    fmt = 3,
    coef_rename = c(
      "conditionviolent" = "Condition (violent)",
      "trait_hostility"  = "Trait hostility",
      "empathy"          = "Empathy",
      "hours_weekly"     = "Hours weekly"
    ),
    notes = "*Note*. Standard errors in parentheses. *p* < .05 (*), *p* < .01 (**), *p* < .001 (***)."
  )
} else {
  cat("(modelsummary not installed; on your machine the table would show\n")
  cat("the three nested models side by side with coefficients, SEs, and stars.)\n")
}
```

For the visualization, a coefficient plot from the full model:

```{r}
#| label: fig-multi-reg-coefplot
#| fig-width: 6.5
#| fig-height: 3.4
#| fig-align: center

if (requireNamespace("broom", quietly = TRUE)) {
  library(broom)
  coef_df <- tidy(fit3, conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
    mutate(label = c("Condition (violent)", "Trait hostility",
                     "Empathy", "Hours weekly"))
  coef_df$label <- factor(coef_df$label, levels = rev(coef_df$label))

  ggplot(coef_df, aes(x = estimate, y = label)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0, linewidth = 0.6,
                   color = "steelblue4") +
    geom_point(size = 3, color = "steelblue") +
    labs(x = "Unstandardized coefficient (95% CI)",
         y = NULL,
         title = "Predictors of aggression: full model coefficients") +
    theme_minimal(base_size = 10.5) +
    theme(plot.title = element_text(face = "bold", size = 11),
          panel.grid.minor = element_blank())
}
```

The table shows everything; the plot makes the substantive pattern visible at a glance. Used together, they reinforce each other.

## Moderation and marginal plots

A moderation analysis tests whether the slope of one predictor depends on the value of another. When the moderator is categorical (as in `condition × trait_hostility`), the figure is two regression lines, one per moderator level. When the moderator is continuous, the figure shows the predicted relationship at several representative values of the moderator (typically the mean and ±1 SD).

```{r}
#| label: fig-moderation
#| fig-width: 6.5
#| fig-height: 3.6
#| fig-align: center

ggplot(vgames, aes(x = trait_hostility, y = aggression,
                   color = condition)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE,
              formula = y ~ x) +
  scale_color_manual(values = c("neutral" = "#377eb8",
                                "violent" = "#e41a1c"),
                     name = "Condition") +
  labs(x = "Trait hostility (1-7)",
       y = "Aggression (1-10)",
       title = "Moderation: hostility-aggression slope by condition") +
  theme_minimal(base_size = 10.5) +
  theme(plot.title = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank())
```

If the two slopes differ noticeably, the visual fingerprint of moderation (Chapter 18) is present: the lines are not parallel. If they were parallel, the relationship would be the same in both conditions and there would be no interaction.

For the reporting:

> *The interaction between trait hostility and condition was significant, b = [value], SE = [value], t(116) = [value], p = [value], 95% CI [low, high]. Simple-slopes analysis indicated that trait hostility predicted aggression more strongly in the violent condition (b = [value], SE = [value], p < .001) than in the neutral condition (b = [value], SE = [value], p = [value]).*

For a continuous moderator, the equivalent figure shows the predicted line at several moderator values:

```{r}
#| label: fig-marginal-cont
#| fig-width: 6.5
#| fig-height: 3.6
#| fig-align: center

# Predict at low, mean, high empathy
mean_emp <- mean(vgames$empathy)
sd_emp   <- sd(vgames$empathy)

fit_mod <- lm(aggression ~ trait_hostility * empathy,
              data = vgames)

newdata <- expand.grid(
  trait_hostility = seq(1, 7, by = 0.1),
  empathy = c(mean_emp - sd_emp, mean_emp, mean_emp + sd_emp)
)
preds <- predict(fit_mod, newdata, interval = "confidence")
newdata$predicted <- preds[, "fit"]
newdata$lower     <- preds[, "lwr"]
newdata$upper     <- preds[, "upr"]
newdata$empathy_level <- factor(
  rep(c("Low (-1 SD)", "Mean", "High (+1 SD)"),
      each = nrow(newdata) / 3),
  levels = c("Low (-1 SD)", "Mean", "High (+1 SD)")
)

ggplot(newdata, aes(x = trait_hostility, y = predicted,
                    color = empathy_level, fill = empathy_level)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2,
              color = NA) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("steelblue", "grey40", "firebrick")) +
  scale_fill_manual(values = c("steelblue", "grey40", "firebrick")) +
  labs(x = "Trait hostility (1-7)",
       y = "Predicted aggression (1-10)",
       color = "Empathy",
       fill  = "Empathy",
       title = "Marginal effect: hostility-aggression slope at three empathy levels") +
  theme_minimal(base_size = 10.5) +
  theme(plot.title = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank())
```

The three lines show the predicted relationship between trait hostility and aggression at three representative levels of empathy. If the lines are parallel, there is no interaction; if they fan out or converge, there is a moderation effect, and the visual pattern shows in which direction the moderator changes the relationship. This kind of plot is sometimes called a **marginal effects plot** or a **conditional effects plot** and is the standard visualization for continuous moderation.

## The general principle

Each analysis type has a natural visualization that puts the substantive answer directly in front of the reader. The natural visualization is rarely the bar chart with error bars. It is, depending on the analysis: an estimation plot with a floating difference axis (t-test), a combined violin-jitter-mean by group (ANOVA), a scatter with fitted line (simple regression), a coefficient plot (multiple regression), or a multi-line marginal effects plot (moderation). Each of these is built with the same `ggplot2` grammar (Chapter 30) and stacks the same kinds of layers; the difference is which layers, in which order, are most appropriate to the analysis.

When you reach for a visualization, ask first what the substantive question is, then choose the figure that answers it most directly. The corresponding table (Chapter 34) and reporting style (Chapter 35) follow from there.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "The right visualization for a t-test is a standard bar chart with error bars showing the means of the two groups, because this is the most familiar format for psychology readers.",
  answer = "The right visualization for a t-test puts both group means with their CIs alongside the difference between means with its own CI on a floating axis, so the inferential question 'does the difference exclude zero?' is visible directly without requiring the reader to combine separate CIs.",
  "The right visualization for a t-test is the underlying t-distribution with the observed t-statistic marked, because this makes the long-run frequency interpretation of the p-value visually apparent.",
  "The right visualization for a t-test is a single point estimate of Cohen's d with its 95% CI, with no display of the underlying data because the standardized effect size carries all relevant information."
)

opts_q2 <- c(
  "The right visualization for a one-way ANOVA is a single bar chart showing the grand mean across all groups, because the omnibus F-test is asking about the average aggregation across groups rather than individual group means.",
  answer = "The right visualization for a one-way ANOVA shows the distribution of the outcome variable per group (using violin, boxplot, jitter, or a combination), with group means and 95% CIs overlaid, supporting visual comparison of both central tendency and spread across groups.",
  "The right visualization for a one-way ANOVA is the F-distribution with the observed F-statistic marked, because this directly conveys the strength of evidence against the null hypothesis of no group differences.",
  "The right visualization for a one-way ANOVA is the residual-versus-fitted-value diagnostic plot, because ANOVA assumes equal variances and the diagnostic plot is the only way to evaluate this assumption."
)

opts_q3 <- c(
  "The right visualization for a simple regression is a histogram of the residuals, which shows whether the assumption of normally distributed residuals is met by the data.",
  answer = "The right visualization for a simple regression is a scatterplot of the outcome against the predictor with the fitted regression line and its 95% CI band overlaid, conveying both the strength of the linear association and the precision of the estimate.",
  "The right visualization for a simple regression is a bar chart of the coefficient values, because the height of each bar makes the relative magnitudes of the predictors easy to compare.",
  "The right visualization for a simple regression is a Q-Q plot of the standardized residuals against theoretical quantiles, because this shows whether the data meet the assumptions of the regression."
)

opts_q4 <- c(
  "The right visualization for a multiple regression is a single scatter plot showing the outcome variable as a function of all predictors simultaneously, with each predictor encoded as a different color or shape of point in the same scatter.",
  answer = "The right visualization for a multiple regression is a coefficient plot, showing one row per predictor with the estimate as a dot and the 95% CI as a horizontal line, with standardized coefficients used when predictors are on different scales to allow a fair comparison of effect magnitudes.",
  "The right visualization for a multiple regression is a heatmap of the correlation matrix among the predictors, because multicollinearity is the central concern in interpreting a multiple regression and visualizing it is the most important diagnostic.",
  "The right visualization for a multiple regression is a forest plot of the predictors with each predictor's confidence interval shown in a different color, since each predictor's effect is contingent on the values of the other predictors."
)

opts_q5 <- c(
  "The right visualization for a moderation effect with a categorical moderator is a side-by-side bar chart showing the mean of the outcome at each combination of the focal predictor and the moderator level.",
  answer = "The right visualization for a moderation effect with a categorical moderator is one regression line per level of the moderator on the same axes, so non-parallel lines reveal the interaction; for a continuous moderator, the equivalent is the predicted line at several representative values (often mean and ±1 SD) of the moderator.",
  "The right visualization for a moderation effect is the F-statistic distribution for the interaction term, because the interaction's significance is the key inferential question and the distribution makes the long-run frequency interpretation visual.",
  "The right visualization for a moderation effect is a single regression line averaged across moderator levels, since the moderation effect represents a deviation from the average relationship that should be inspected separately as a residual diagnostic."
)
```

**1. What is the appropriate visualization for an independent-samples t-test?**

`r longmcq(opts_q1)`

**2. What is the appropriate visualization for a one-way ANOVA?**

`r longmcq(opts_q2)`

**3. What is the appropriate visualization for a simple regression with one continuous predictor?**

`r longmcq(opts_q3)`

**4. What is the appropriate visualization for a multiple regression with several predictors of similar substantive interest?**

`r longmcq(opts_q4)`

**5. What is the appropriate visualization for a moderation (interaction) effect?**

`r longmcq(opts_q5)`

## References

Cumming, G. (2014). The new statistics: Why and how. *Psychological Science, 25*(1), 7-29. <https://doi.org/10.1177/0956797613504966>
