---
title: "DAGs: drawing your causal assumptions"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r}
#| label: dag-helper
#| include: false

# Helper for drawing clean DAGs throughout Section 6 chapters.
# nodes: data.frame with columns name, label, x, y
# edges: data.frame with columns from, to (referring to node names)
# node_pad: line-shortening at each end so the arrowhead lands outside
#           the label, not inside it. Larger labels need larger node_pad.
draw_dag <- function(nodes, edges, title = "",
                     node_pad = 0.46,
                     col_node = "steelblue4", col_edge = "grey25",
                     highlight_edges = NULL,
                     highlight_color = "firebrick",
                     xlim = NULL, ylim = NULL,
                     arrow_length = 0.45, arrow_angle = 20,
                     label_size = 3.4) {
  edges_df <- edges |>
    dplyr::left_join(
      nodes |> dplyr::rename(from = name,
                              x_from = x, y_from = y),
      by = "from") |>
    dplyr::left_join(
      nodes |> dplyr::rename(to = name,
                              x_to = x, y_to = y),
      by = "to") |>
    dplyr::mutate(
      ang = atan2(y_to - y_from, x_to - x_from),
      x_from_s = x_from + node_pad * cos(ang),
      y_from_s = y_from + node_pad * sin(ang),
      x_to_s   = x_to   - node_pad * cos(ang),
      y_to_s   = y_to   - node_pad * sin(ang),
      hl = if (!is.null(highlight_edges))
        paste(from, to) %in% highlight_edges else FALSE
    )
  p <- ggplot() +
    geom_segment(data = edges_df,
                 aes(x = x_from_s, y = y_from_s,
                     xend = x_to_s, yend = y_to_s,
                     color = hl, linewidth = hl),
                 arrow = arrow(length = unit(arrow_length, "cm"),
                               type = "closed",
                               angle = arrow_angle)) +
    scale_color_manual(values = c("FALSE" = col_edge,
                                  "TRUE" = highlight_color),
                       guide = "none") +
    scale_linewidth_manual(values = c("FALSE" = 0.75,
                                      "TRUE" = 1.4),
                           guide = "none") +
    geom_label(data = nodes, aes(x = x, y = y, label = label),
               size = label_size, fontface = "bold",
               fill = "white", color = col_node,
               label.padding = unit(0.30, "lines"),
               label.r = unit(0.45, "lines")) +
    labs(title = title) +
    theme_void(base_size = 10) +
    theme(plot.title = element_text(face = "bold",
                                    hjust = 0.5, size = 10.5),
          plot.margin = margin(8, 8, 8, 8))
  if (!is.null(xlim) && !is.null(ylim))
    p <- p + coord_equal(xlim = xlim, ylim = ylim, clip = "off")
  else
    p <- p + coord_equal(clip = "off")
  p
}
```

Chapter 41 closed by acknowledging that many research questions cannot be answered by random assignment. When the treatment is something we cannot manipulate (a personality trait, a life event, a long-term lifestyle pattern), we are stuck with **observational data**, and the observed group mean difference may or may not be the average treatment effect. We need a way to reason carefully about which differences between groups reflect real causal effects and which reflect pre-existing differences in the groups themselves.

The conceptual tool for this reasoning is the **directed acyclic graph**, or **DAG**. A DAG is a picture of your causal assumptions. It is not derived from the data and it cannot be tested directly against the data. It is a diagram of what you believe causes what, drawn before you run the regression, and it determines what set of variables you should put on the right-hand side of `lm()` to recover a causal estimate.

This chapter introduces DAGs in the simplest possible way. We will see what a DAG is (just nodes and arrows), how to read one (by tracing paths), and what kind of statement a DAG actually makes (about your beliefs, not about the data). Chapter 43 will then introduce the three classic third-variable structures that DAGs let us see clearly. Chapter 44 will tie them together into the procedure for choosing the right set of control variables.

DAGs in their modern form for psychology come from Pearl ([1995](https://doi.org/10.2307/2337329)) and were brought to applied psychological research by Rohrer ([2018](https://doi.org/10.1177/2515245917745629)). The chapter follows Rohrer's framing closely.

## What a DAG is

A DAG has two ingredients:

- **Nodes**: the variables. Each variable in your causal model is one node. Draw them as labels or as circles with labels inside.
- **Arrows**: causal claims. An arrow from variable *A* to variable *B* means "*A* causes *B*." The arrow has a direction; *A* → *B* is different from *B* → *A*.

That is the entire content of a DAG. Two ingredients, no more.

The "directed" in "directed acyclic graph" refers to the fact that the arrows have a direction (one-headed, never two-headed). The "acyclic" means that the arrows do not form a loop; you cannot follow arrows from one variable back to itself, directly or indirectly. A variable does not cause itself.

The simplest DAG we can draw for `vgames_study` has just two nodes (the violent-game variable and the aggression variable) and one arrow:

```{r}
#| label: fig-minimal-dag
#| echo: false
#| fig-width: 4
#| fig-height: 2
#| fig-align: center

nodes_min <- data.frame(
  name  = c("G", "A"),
  label = c("games", "aggression"),
  x = c(-0.5, 1.0), y = c(0, 0)
)
edges_min <- data.frame(from = "G", to = "A")
draw_dag(nodes_min, edges_min,
         title = "The minimal DAG",
         xlim = c(-1.0, 1.6), ylim = c(-0.4, 0.4))
```

This DAG says: the violent game causes aggression. That is the substantive claim we want our study to support. The arrow is the causal effect we want to estimate. The DAG does not say what *magnitude* the effect is, or whether the relationship is linear or curved; it just says "the violent game causally affects aggression in some way."

For the rest of the chapter, when we draw a DAG, we adopt the convention that the arrow we want to estimate (the causal effect of *X* on *Y*) is the one we have written explicitly. Other arrows we add represent additional causal beliefs about other variables in the system.

## Building up the DAG

The minimal DAG is rarely the right DAG. In reality, there are usually other variables involved that affect both the predictor and the outcome. To do justice to the situation, we need to extend the DAG to include them.

Suppose we believe that trait hostility also matters: hostile people tend to choose violent games (when given the choice) *and* tend to score higher on aggression measures. Both arrows go out of trait hostility, one into games and one into aggression. We add trait hostility as a node:

```{r}
#| label: fig-dag-hostility
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

nodes_h <- data.frame(
  name  = c("G", "A", "H"),
  label = c("games", "aggression", "trait\nhostility"),
  x = c(-0.5, 1.0, 0.25), y = c(0, 0, 1.2)
)
edges_h <- data.frame(
  from = c("G", "H", "H"),
  to   = c("A", "G", "A")
)
draw_dag(nodes_h, edges_h,
         title = "A DAG with trait hostility added",
         xlim = c(-1.0, 1.6), ylim = c(-0.4, 1.7))
```

This is a much more substantively realistic DAG, and we will revisit it in Chapter 43. For now, notice what it shows. There are now multiple ways the variables games and aggression can be statistically connected:

1. The direct causal arrow: games → aggression. This is the real causal effect we want.
2. A "back way around": games ← trait hostility → aggression. Trait hostility causes both, so it produces a statistical connection between the two even if games has no causal effect on aggression at all.

The second of these is a **backdoor path**: it goes "out the back" of games (against an arrow that points into games) and reaches aggression via some other route. The path carries a *spurious* association between games and aggression, induced by the common cause. If we want a clean causal estimate, we need to figure out how to suppress the backdoor path while keeping the direct causal path open.

This is the central project of the rest of Section 6.

## Paths in a DAG

A **path** between two variables is any sequence of arrows that connects them. The arrows can go in either direction along the path; we trace the path by walking from node to node regardless of which way each arrow points.

In the DAG above, the path between games and aggression can be either of these:

- *Direct causal path:* games → aggression. One arrow, forward.
- *Backdoor path:* games ← trait hostility → aggression. Two arrows, the first one points *into* games (which is what makes it a backdoor) and the second one points *out of* trait hostility into aggression.

We need to distinguish two qualitatively different kinds of paths.

**Causal paths.** A causal path is one that follows arrows in the forward direction throughout. It starts at *X* and reaches *Y* by always going with the arrows, never against them. A causal path transmits a *causal* association: if *X* changes, *Y* changes in the direction the path implies. We *want* to leave causal paths open, because they carry the real effect we are trying to measure.

**Backdoor paths.** A backdoor path is one that starts with an arrow pointing *into* *X* (not out of it) and reaches *Y* somehow, possibly via additional variables. The first segment of a backdoor path goes against the direction of the arrow on *X*. Backdoor paths transmit *spurious* associations: changes in the common cause produce co-movement in *X* and *Y*, even when *X* has no causal effect on *Y*. We want to *close* backdoor paths, because they are sources of bias.

The figure below highlights the two kinds of paths in the trait-hostility DAG.

```{r}
#| label: fig-paths-highlight
#| echo: false
#| fig-width: 9
#| fig-height: 3
#| fig-align: center

p_causal <- draw_dag(
  nodes_h, edges_h,
  title = "Causal path (highlighted)",
  highlight_edges = c("G A"),
  xlim = c(-1.0, 1.6), ylim = c(-0.4, 1.7))
p_back <- draw_dag(
  nodes_h, edges_h,
  title = "Backdoor path (highlighted)",
  highlight_edges = c("H G", "H A"),
  xlim = c(-1.0, 1.6), ylim = c(-0.4, 1.7))

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_causal, p_back, ncol = 2)
} else {
  gridExtra::grid.arrange(p_causal, p_back, ncol = 2)
}
```

The left panel shows the causal path: the single arrow going from games to aggression. The right panel shows the backdoor path: two arrows, both emanating from trait hostility. Both paths transmit a statistical association between games and aggression. Only the first is a causal effect.

The whole game of causal inference from observational data is now stated as follows. *Keep the causal paths open; close the backdoor paths.* If we can do that, the association we measure equals the causal effect.

## How "controlling for" relates to paths

Here is the link to the regression machinery you have been using. When you include a variable in a regression by adding it to the right-hand side of `lm()` (`lm(aggression ~ condition + trait_hostility)`), you are **statistically controlling for** that variable. The technical operation is to compute the slope on the condition variable *holding trait hostility constant*. In DAG terms, controlling for a variable is said to **block** the paths that pass through it.

In the trait-hostility DAG above, controlling for trait hostility blocks the backdoor path games ← trait hostility → aggression. The path passes through the trait hostility node, and conditioning on (holding constant) that node closes the path. The path no longer transmits any association.

This is the basic idea of why "adding controls" can fix a confounding problem. The control variable closes the backdoor path that was creating the spurious association, leaving only the direct causal path open. The coefficient on the focal predictor then reflects the causal effect alone.

But, and this is the warning that Chapter 43 will develop carefully, blocking is *not* always good. Sometimes the path passing through a variable is the *causal* path we want to keep open (which is the case for **mediators**), and blocking it kills part of the effect we are trying to measure. Sometimes blocking a variable does not close a path but *opens* one that was previously closed (which is the case for **colliders**), creating a fake association out of nothing.

This is the central message: whether to control for a variable depends on its causal role in the DAG. The role is determined by which way the arrows point. The data alone cannot tell us this; it has to come from our background knowledge. The DAG makes that knowledge explicit and checkable.

## The three elementary causal structures

Most of the DAG patterns you will ever encounter are built from three simple sub-structures. Rohrer ([2018](https://doi.org/10.1177/2515245917745629)) calls them **chains**, **forks**, and **inverted forks**. We will give each one its own DAG and describe what kind of association it transmits.

```{r}
#| label: fig-three-structures
#| echo: false
#| fig-width: 9.5
#| fig-height: 3.6
#| fig-align: center

nodes_chain <- data.frame(
  name = c("A", "B", "C"),
  label = c("A", "B", "C"),
  x = c(-1.2, 0.0, 1.2), y = c(0, 0, 0))
edges_chain <- data.frame(from = c("A", "B"),
                          to   = c("B", "C"))
p_chain <- draw_dag(nodes_chain, edges_chain,
                    title = "Chain: A -> B -> C",
                    node_pad = 0.22,
                    xlim = c(-1.6, 1.6),
                    ylim = c(-0.6, 0.6))

nodes_fork <- data.frame(
  name = c("A", "B", "C"),
  label = c("A", "B", "C"),
  x = c(-1.2, 0.0, 1.2), y = c(-0.4, 0.8, -0.4))
edges_fork <- data.frame(from = c("B", "B"),
                         to   = c("A", "C"))
p_fork <- draw_dag(nodes_fork, edges_fork,
                   title = "Fork: A <- B -> C",
                   node_pad = 0.22,
                   xlim = c(-1.6, 1.6),
                   ylim = c(-0.8, 1.2))

nodes_inv <- data.frame(
  name = c("A", "B", "C"),
  label = c("A", "B", "C"),
  x = c(-1.2, 0.0, 1.2), y = c(0.8, -0.4, 0.8))
edges_inv <- data.frame(from = c("A", "C"),
                        to   = c("B", "B"))
p_inv <- draw_dag(nodes_inv, edges_inv,
                  title = "Inverted fork: A -> B <- C",
                  node_pad = 0.22,
                  xlim = c(-1.6, 1.6),
                  ylim = c(-0.8, 1.2))

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_chain, p_fork, p_inv, ncol = 3)
} else {
  gridExtra::grid.arrange(p_chain, p_fork, p_inv, ncol = 3)
}
```

**Chain** (*A* → *B* → *C*). One arrow flows from *A* to *B*, then another from *B* to *C*. A chain transmits a *causal* association between the endpoints (*A* causally affects *C* via *B*). If you change *A*, *B* changes (caused by *A*), and *C* changes (caused by the new *B*). When *B* sits on a chain from *X* to *Y* and we want the total causal effect of *X* on *Y*, we *do not* want to block *B*, because that would kill part of the effect.

**Fork** (*A* ← *B* → *C*). *B* has arrows pointing out to both *A* and *C*. A fork transmits an association between *A* and *C*, but the association is *not* causal: it reflects the common cause *B*, not any causal relationship between *A* and *C* themselves. Forks are the structure of confounding. When *B* sits on a fork between *X* and *Y* in a DAG, we *do* want to block *B*, because that closes the backdoor path.

**Inverted fork** (*A* → *B* ← *C*). Both *A* and *C* have arrows pointing into *B*. *B* is the common effect of *A* and *C*. The inverted fork *does not* transmit any association between *A* and *C*; they are causally unrelated and statistically independent if *B* is left alone. The surprising fact is that *controlling for* *B* opens an association between *A* and *C* that did not exist before. Inverted forks are the structure of **collider bias**, and they are the reason why "control for everything" is dangerous. Chapter 43 develops this in detail.

These three structures are exhaustive. Any path between two variables, however long, breaks down into chains, forks, and inverted forks. Once you know how each structure transmits association, you can analyze the whole DAG by analyzing each path piece by piece.

## What a DAG is a statement about

A common misunderstanding when first encountering DAGs is to think of them as something derived from the data. They are not. A DAG is a statement about *causal beliefs*. It encodes the structure of causes you bring to the analysis, based on theory, prior research, and substantive knowledge of the problem.

Two DAGs can be consistent with the same observed data:

- *Story 1*: games → aggression. The violent game causally raises aggression.
- *Story 2*: aggression → games. People feeling more aggressive choose the violent game.

Both stories produce a positive correlation between game choice and aggression in observational data. The data cannot distinguish them. To choose one over the other, you have to bring in causal knowledge from outside the data: temporal ordering (in `vgames_study`, the game came before the aggression measurement, so games → aggression is the temporally plausible direction), substantive theory, prior experiments. The DAG is the place where you make those external commitments explicit.

This is sometimes worrying to people who want statistics to be "purely data-driven." The reality is that causal claims cannot be purely data-driven. You always need to bring causal knowledge to the data; the question is whether you bring it explicitly (as in a DAG) or implicitly (hidden in the choice of which variables to control for in a regression). The DAG makes the assumptions visible, so that you, your collaborators, and your readers can examine and challenge them. Explicit assumptions are *better* than hidden assumptions, even when the assumptions themselves are debatable.

This is what Rohrer ([2018](https://doi.org/10.1177/2515245917745629), p. 28) emphasizes: drawing a DAG forces you to commit to specific causal claims and to make them visible. The DAG can be wrong, but its wrongness can be discussed and refined. A hidden assumption, by contrast, cannot.

## A practical workflow

The procedure for using DAGs in an applied analysis is straightforward. We will develop it across the next two chapters, but the overall shape is:

1. **Draw the DAG.** Before any analysis, sit down with what you know about the variables in your study and draw a DAG showing what you believe causes what. Include any variables that might be related to both your focal *X* and your *Y*.

2. **Identify the focal causal arrow.** Mark the arrow whose magnitude you want to estimate (typically *X* → *Y*).

3. **Identify the backdoor paths.** List all the paths from *X* to *Y* that start with an arrow into *X*. These are the spurious paths you need to close.

4. **Choose an adjustment set.** Find a set of variables that, when controlled for in the regression, blocks every backdoor path *without* blocking the focal causal arrow and *without* opening any collider paths. Chapter 44 develops the precise rule for this (the backdoor criterion).

5. **Run the regression.** Fit `lm(Y ~ X + <adjustment set>)`. Under the assumptions encoded in the DAG, the coefficient on *X* is now an estimate of the causal effect.

6. **Honestly report the assumptions.** State the DAG you assumed, the adjustment set you used, and the threats to validity (variables that should have been measured but were not, plausible alternative DAGs that would change your conclusions). The DAG makes this reporting explicit.

The next two chapters work through steps 2 to 5 in detail. The point of this chapter is to establish that a DAG is the right starting point for thinking about any observational causal question.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "A directed acyclic graph is a statistical procedure for fitting a network of relationships among variables, with the strengths of the connections estimated from the data using maximum likelihood.",
  answer = "A directed acyclic graph is a diagram of causal assumptions: nodes represent variables and arrows represent direct causal effects, with the constraint that the arrows do not form loops (a variable cannot cause itself directly or indirectly).",
  "A directed acyclic graph is a graphical summary of the correlation matrix among variables in a dataset, with arrow direction determined by the temporal ordering of the variables' measurement.",
  "A directed acyclic graph is an alternative form of regression diagnostic plot used to assess linearity and independence assumptions, with arrows indicating the residuals at different fitted values."
)

opts_q2 <- c(
  "A causal path in a DAG is any sequence of arrows connecting two variables, regardless of the direction of the arrows along the path.",
  answer = "A causal path in a DAG is a path that follows arrows in the forward direction throughout (from cause to effect to further effect), transmitting a real causal association between the start and end nodes.",
  "A causal path in a DAG is a path that contains at least one fork in its structure, regardless of the directions of the other arrows along the path.",
  "A causal path in a DAG is a path that has been statistically tested and shown to have a coefficient greater than zero in absolute value at the conventional significance level."
)

opts_q3 <- c(
  "A backdoor path between X and Y is any path that contains more than two intermediate nodes, regardless of the direction of the arrows in the path.",
  answer = "A backdoor path between X and Y is a path that starts with an arrow pointing into X (going against the arrow) and reaches Y via some other route, transmitting a spurious association created by a common cause or some other non-causal structure.",
  "A backdoor path between X and Y is a hidden statistical association created by measurement error in the predictor or outcome, and disappears when the measurement error is correctly modeled.",
  "A backdoor path between X and Y is a regression coefficient that has the opposite sign from the focal effect, indicating a suppression effect in the multiple regression model."
)

opts_q4 <- c(
  "Controlling for a variable in a regression by adding it to the right-hand side of lm() removes that variable from the analysis entirely, effectively excluding it from the dataset for the purposes of the comparison.",
  answer = "Controlling for a variable in a regression by adding it to the right-hand side of lm() blocks the statistical paths that pass through that variable in the DAG; this can close backdoor paths (good, when the variable is a confounder) but can also block causal paths (bad, when the variable is a mediator) or open new paths (bad, when the variable is a collider).",
  "Controlling for a variable in a regression by adding it to the right-hand side of lm() is mathematically identical to dropping cases from the dataset based on the values of that variable.",
  "Controlling for a variable in a regression by adding it to the right-hand side of lm() always reduces the standard error of the coefficient on the focal predictor, regardless of the causal role of the variable being controlled."
)

opts_q5 <- c(
  "The three elementary causal structures (chains, forks, and inverted forks) are statistical patterns that can be detected automatically from the data using techniques such as partial correlation or conditional independence testing.",
  answer = "The three elementary causal structures (chains A -> B -> C, forks A <- B -> C, inverted forks A -> B <- C) are the building blocks of any DAG; chains transmit causal associations, forks transmit spurious associations from common causes, and inverted forks transmit no association unless their middle node is conditioned on, in which case a spurious association appears.",
  "The three elementary causal structures are equivalent to one another and the choice between them is a matter of graphical convention rather than substantive meaning.",
  "The three elementary causal structures are alternative ways of drawing the same causal relationship, with chains being used for short paths, forks for medium-length paths, and inverted forks for long or complex paths."
)

opts_q6 <- c(
  answer = "A DAG is a statement about causal beliefs based on theory, prior research, and substantive knowledge; it is not derived from the data and cannot be tested directly against the data, although different DAGs imply different testable conditional independence patterns.",
  "A DAG is a statistical model fitted to the data with parameters estimated by maximum likelihood, and its goodness of fit can be assessed using standard regression diagnostics such as the AIC or BIC.",
  "A DAG is determined entirely by the temporal ordering of the variables in the dataset: any variable measured earlier in time must be drawn as a cause of any variable measured later, regardless of substantive knowledge.",
  "A DAG is a representation of the conditional independence pattern observed in a particular dataset, with arrows added between every pair of variables that show a partial correlation above a chosen threshold."
)
```

**1. What is a directed acyclic graph (DAG)?**

`r longmcq(opts_q1)`

**2. What is a causal path in a DAG?**

`r longmcq(opts_q2)`

**3. What is a backdoor path between two variables in a DAG?**

`r longmcq(opts_q3)`

**4. What does it mean to "control for" a variable in the regression framework, in terms of paths in a DAG?**

`r longmcq(opts_q4)`

**5. What are the three elementary causal structures in a DAG, and how do they transmit (or block) statistical associations?**

`r longmcq(opts_q5)`

**6. What is the epistemic status of a DAG (where does it come from, and what is it a statement about)?**

`r longmcq(opts_q6)`

## References

Pearl, J. (1995). Causal diagrams for empirical research. *Biometrika, 82*(4), 669-688. <https://doi.org/10.2307/2337329>

Pearl, J., Glymour, M., & Jewell, N. P. (2016). *Causal inference in statistics: A primer*. Wiley.

Rohrer, J. M. (2018). Thinking clearly about correlations and causation: Graphical causal models for observational data. *Advances in Methods and Practices in Psychological Science, 1*(1), 27-42. <https://doi.org/10.1177/2515245917745629>
