---
title: "Confounders, mediators, colliders"
---

```{r}
#| label: setup
#| include: false
library(webexercises)
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r}
#| label: dag-helper
#| include: false

draw_dag <- function(nodes, edges, title = "",
                     node_pad = 0.46,
                     col_node = "steelblue4", col_edge = "grey25",
                     highlight_edges = NULL,
                     highlight_color = "firebrick",
                     xlim = NULL, ylim = NULL,
                     arrow_length = 0.45, arrow_angle = 20,
                     label_size = 3.2) {
  edges_df <- edges |>
    dplyr::left_join(
      nodes |> dplyr::rename(from = name,
                              x_from = x, y_from = y),
      by = "from") |>
    dplyr::left_join(
      nodes |> dplyr::rename(to = name,
                              x_to = x, y_to = y),
      by = "to") |>
    dplyr::mutate(
      ang = atan2(y_to - y_from, x_to - x_from),
      x_from_s = x_from + node_pad * cos(ang),
      y_from_s = y_from + node_pad * sin(ang),
      x_to_s   = x_to   - node_pad * cos(ang),
      y_to_s   = y_to   - node_pad * sin(ang),
      hl = if (!is.null(highlight_edges))
        paste(from, to) %in% highlight_edges else FALSE
    )
  p <- ggplot() +
    geom_segment(data = edges_df,
                 aes(x = x_from_s, y = y_from_s,
                     xend = x_to_s, yend = y_to_s,
                     color = hl, linewidth = hl),
                 arrow = arrow(length = unit(arrow_length, "cm"),
                               type = "closed",
                               angle = arrow_angle)) +
    scale_color_manual(values = c("FALSE" = col_edge,
                                  "TRUE" = highlight_color),
                       guide = "none") +
    scale_linewidth_manual(values = c("FALSE" = 0.75,
                                      "TRUE" = 1.4),
                           guide = "none") +
    geom_label(data = nodes, aes(x = x, y = y, label = label),
               size = label_size, fontface = "bold",
               fill = "white", color = col_node,
               label.padding = unit(0.30, "lines"),
               label.r = unit(0.45, "lines")) +
    labs(title = title) +
    theme_void(base_size = 10) +
    theme(plot.title = element_text(face = "bold",
                                    hjust = 0.5, size = 10.5),
          plot.margin = margin(8, 8, 8, 8))
  if (!is.null(xlim) && !is.null(ylim))
    p <- p + coord_equal(xlim = xlim, ylim = ylim, clip = "off")
  else
    p <- p + coord_equal(clip = "off")
  p
}
```

Chapter 42 introduced the DAG as a picture of causal assumptions and named the three elementary structures it can be built from (chains, forks, inverted forks). This chapter is about what those structures *do* when we use them in practice. We will look at the three classic kinds of third variables that researchers encounter when they want to control for something in a regression: **confounders**, **mediators**, and **colliders**. Each has a distinctive structure in a DAG, a distinctive rule for whether to include it as a control variable, and a distinctive consequence when you get the rule wrong.

This is the most important conceptual chapter of Section 6. Get the three types right, and the rest of causal inference becomes a matter of careful application. Get them wrong, and `lm()` will silently produce confidently biased answers.

We will work with `vgames_study` throughout, expanding the running example with a few new variables to give us realistic examples of each third-variable role. Throughout the chapter we will use the same regression machinery you have used since Chapter 16. The point is not to learn new R syntax. The point is to learn which variables to include and which to exclude, given a particular causal structure.

## Setup: the focal causal question

Throughout this chapter, the focal causal question is the same as the rest of Section 6: does playing the violent video game cause an increase in aggression? In a DAG, this is the arrow we want to estimate the magnitude of.

```{r}
#| label: fig-focal
#| echo: false
#| fig-width: 4
#| fig-height: 1.6
#| fig-align: center

nodes_focal <- data.frame(
  name = c("G", "A"),
  label = c("games", "aggression"),
  x = c(-0.5, 1.0), y = c(0, 0)
)
edges_focal <- data.frame(from = "G", to = "A")
draw_dag(nodes_focal, edges_focal,
         title = "The focal causal question",
         highlight_edges = "G A",
         xlim = c(-1.0, 1.6), ylim = c(-0.4, 0.4))
```

The red arrow is what we want to estimate. The question is: which other variables, if any, should we include in `lm(aggression ~ games + ...)` to recover an unbiased estimate of this arrow? The answer depends on the role each candidate variable plays in the larger causal structure.

## Confounders: control them

A **confounder** is a variable that causes *both* the predictor and the outcome. It is the fork structure from Chapter 42, applied to the situation where the fork sits between *X* and *Y* in the DAG.

In our running example, a plausible confounder is **trait hostility**. Hostile people are more likely to choose (or to enjoy, or to play more of) violent games, and they are also more likely to score higher on aggression measures, independently of whatever the game does to them. Trait hostility has arrows pointing to both games and aggression.

```{r}
#| label: fig-confounder
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

nodes_conf <- data.frame(
  name = c("G", "A", "H"),
  label = c("games", "aggression", "trait\nhostility"),
  x = c(-0.5, 1.0, 0.25), y = c(0, 0, 1.2)
)
edges_conf <- data.frame(
  from = c("G", "H", "H"),
  to   = c("A", "G", "A")
)
draw_dag(nodes_conf, edges_conf,
         title = "Confounder: trait hostility (C)",
         highlight_edges = c("H G", "H A"),
         xlim = c(-1.0, 1.6), ylim = c(-0.4, 1.7))
```

The red arrows in the figure are the backdoor path: games ← trait hostility → aggression. This path transmits a spurious association between games and aggression because the two share a common cause. Even if the game itself does nothing to aggression, hostile people will both play more violent games and score higher on aggression measures, and the regression will detect a positive coefficient.

**The rule: control for confounders.** Adding the confounder as a covariate in the regression blocks the backdoor path, leaving only the direct causal path open. In the language of `lm()`:

```{r}
#| label: vgames-load
#| eval: false
vgames <- read.csv("data/vgames_study.csv")
```

```{r}
#| label: vgames-data
#| echo: false
vgames <- tryCatch(
  read.csv("data/vgames_study.csv"),
  error = function(e) {
    set.seed(1)
    n <- 60
    data.frame(
      participant_id  = seq_len(2 * n),
      condition       = rep(c("violent", "neutral"), each = n),
      trait_hostility = rnorm(2 * n, mean = 4, sd = 1.2),
      aggression      = c(rnorm(n, 5.56, 1.6), rnorm(n, 4.21, 1.6))
    )
  }
)
```

```{r}
#| label: confounder-regression

fit_naive    <- lm(aggression ~ condition, data = vgames)
fit_adjusted <- lm(aggression ~ condition + trait_hostility,
                   data = vgames)

cat("Without controlling for trait hostility:\n")
print(coef(fit_naive))
cat("\nControlling for trait hostility:\n")
print(coef(fit_adjusted))
```

The first regression's coefficient on `conditionviolent` is the *raw* observed difference, which mixes the causal effect of the game with the confounding effect of trait hostility. The second regression's coefficient is the difference between groups *holding trait hostility constant*: it estimates the causal effect of the game alone, after closing the backdoor path through hostility.

The visual story is the same as in Chapter 41 (the randomization vs self-selection figure). When the confounder is controlled, the residual comparison is between two groups that are statistically equivalent on trait hostility. This is exactly what randomization would have given us for free; in observational data, we have to purchase it by careful adjustment.

A simulation makes the consequence concrete. We simulate a population where the violent game has a true causal effect of +1.0 on aggression, and where trait hostility is a confounder. We compare two estimates: the naive one (no adjustment) and the adjusted one.

```{r}
#| label: confounder-sim

set.seed(11)
n_sim <- 500
true_effect <- 1.0

# Trait hostility is the confounder.
sim <- data.frame(
  hostility = rnorm(n_sim, mean = 4, sd = 1.2)
)
# Probability of choosing violent goes up with hostility.
sim$games <- ifelse(
  plogis((sim$hostility - 4) * 1.0) > runif(n_sim),
  "violent", "neutral"
)
# Aggression depends on both the game and hostility.
sim$aggression <- 1.0 * sim$hostility +
  true_effect * (sim$games == "violent") +
  rnorm(n_sim, 0, 1)

est_naive    <- coef(lm(aggression ~ games, data = sim))["gamesviolent"]
est_adjusted <- coef(lm(aggression ~ games + hostility, data = sim))["gamesviolent"]

cat(sprintf("True causal effect             = %.2f\n", true_effect))
cat(sprintf("Naive estimate (no control)    = %.2f (biased)\n", est_naive))
cat(sprintf("Adjusted (control hostility)   = %.2f (unbiased)\n", est_adjusted))
```

The naive estimate is biased upward (around 1.6 or so). The adjusted estimate is close to the true effect of 1.0. Controlling for the confounder fixed the problem. This is the case where statistical control does exactly what we hope.

A quick note on randomization. In Chapter 41 we saw that under random assignment of `condition`, no arrows point *into* games (the assignment is determined by a coin), so there can be no backdoor path at all. Randomization makes confounders irrelevant *by construction*. Without randomization, we have to find and control the confounders by hand, and we can only do so if we have measured them. Unmeasured confounders are a problem no statistical method can fix.

## Mediators: leave them alone (for the total effect)

A **mediator** is a variable that sits *on* the causal path from *X* to *Y*. *X* causes *M*, and *M* causes *Y*. The mediator is part of *how* the cause produces the effect. This is the chain structure from Chapter 42, applied to our focal arrow.

In `vgames_study`, a plausible mediator is **physiological arousal**. The violent game increases arousal, and arousal increases aggression. The total causal effect of the game on aggression is partly carried *through* arousal.

```{r}
#| label: fig-mediator
#| echo: false
#| fig-width: 6
#| fig-height: 1.6
#| fig-align: center

nodes_med <- data.frame(
  name = c("G", "M", "A"),
  label = c("games", "arousal", "aggression"),
  x = c(-1.6, 0.0, 1.6), y = c(0, 0, 0)
)
edges_med <- data.frame(
  from = c("G", "M"),
  to   = c("M", "A")
)
draw_dag(nodes_med, edges_med,
         title = "Mediator: physiological arousal (M)",
         highlight_edges = c("G M", "M A"),
         xlim = c(-2.2, 2.2), ylim = c(-0.4, 0.4))
```

The red arrows are the *causal* path games → arousal → aggression. This is part of the effect we are trying to measure. The arousal-mediated piece of the effect is just as real and causal as any direct piece would be.

**The rule: do not control for mediators if you want the total effect.** Controlling for arousal blocks the path games → arousal → aggression. It removes the part of the games' effect that operates through arousal. You are left with only the part of the effect that bypasses arousal (the "direct" effect), which is a different quantity from the total effect.

To make this vivid: if the violent game raises aggression *entirely* through arousal (no direct effect at all), and you control for arousal in your regression, the coefficient on games will be approximately zero. You would conclude, incorrectly, that the game has no effect on aggression. In reality, the game has a real total causal effect; you have just subtracted off the entire mechanism by which it operates.

A simulation makes this concrete:

```{r}
#| label: mediator-sim

set.seed(22)
n_sim <- 500

sim_m <- data.frame(
  games = sample(c("neutral", "violent"), n_sim, replace = TRUE)
)
# Games cause arousal; arousal causes aggression.
sim_m$arousal <- ifelse(sim_m$games == "violent",
                         rnorm(n_sim, 1.5, 1),
                         rnorm(n_sim, 0, 1))
sim_m$aggression <- 1.0 * sim_m$arousal + rnorm(n_sim, 0, 1)

est_total  <- coef(lm(aggression ~ games, data = sim_m))["gamesviolent"]
est_direct <- coef(lm(aggression ~ games + arousal, data = sim_m))["gamesviolent"]

cat(sprintf("Total effect (no mediator in model)        = %.2f\n", est_total))
cat(sprintf("Direct effect only (mediator in model)     = %.2f\n", est_direct))
```

The total effect of the game on aggression is around 1.5 (the full effect operating through arousal). The direct effect, after controlling for arousal, is near zero, because in this simulation there is no direct path from games to aggression that bypasses arousal. Reporting the "direct" coefficient as "the effect of the violent game on aggression" would be a serious mischaracterization of the underlying causal structure.

A subtlety. Sometimes researchers genuinely want to know the *direct* effect (the effect that does not go through a particular mechanism), and in that case they correctly control for the mediator. This is the territory of **mediation analysis**, which requires its own additional assumptions and care (Rohrer, Hünermund, Arslan, & Elson, [2022](https://doi.org/10.1177/25152459221095827)). The rule of thumb for now is: if you want the *total* causal effect of *X* on *Y*, leave the mediators alone. If you want a different, more specialized causal quantity (the part of the effect that does not go through *M*), then mediation analysis is the right framework, but it is more delicate and is the topic of a later section, not Section 6.

## Colliders: do not control them

The last type of third variable is the **collider**. A collider is a variable that is *caused by* both *X* and *Y* (or by two variables on a path). Both arrows point *into* the collider. This is the inverted fork structure from Chapter 42.

In `vgames_study`, a plausible collider is **school disciplinary referral**: a student gets referred to a school counselor for any number of reasons, but two relevant ones are (a) playing a lot of violent video games (which might raise concern), and (b) showing high aggression in school (which might prompt teachers to refer). Both heavy violent-game playing and high aggression independently raise the probability of being referred. Referral is caused by both.

```{r}
#| label: fig-collider
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-align: center

nodes_col <- data.frame(
  name  = c("G", "A", "R"),
  label = c("games", "aggression", "disciplinary\nreferral"),
  x = c(-0.5, 1.0, 0.25), y = c(1.2, 1.2, 0.0)
)
edges_col <- data.frame(
  from = c("G", "G", "A"),
  to   = c("A", "R", "R")
)
draw_dag(nodes_col, edges_col,
         title = "Collider: school disciplinary referral (R)",
         highlight_edges = c("G R", "A R"),
         xlim = c(-1.0, 1.6), ylim = c(-0.4, 1.7))
```

The red arrows are the two arrows *into* the collider. Notice the structure: both games and aggression have arrows pointing at school disciplinary referral. Referral is a common *effect* of the two, not a common *cause*.

Here is the surprising fact about colliders that makes them the conceptual punchline of Section 6. **A collider sits on a path that is naturally closed.** The inverted fork games → referral ← aggression does *not* transmit any spurious association between games and aggression. As long as you leave the collider alone, it creates no problem.

But the moment you **control for** the collider (or select your sample on it), you **open** the path. Conditioning on a common effect creates a fake association between its causes, even when no real association exists.

The rule:

> **Do not control for colliders.** Controlling for a collider introduces bias rather than removing it.

This is the single most counterintuitive result in causal inference. It is also the strongest argument against the "control for everything" instinct. Let us see why it happens, with a simulation.

```{r}
#| label: collider-sim

set.seed(33)
n_sim <- 500

# Suppose, in the population, games and aggression are CAUSALLY UNRELATED
# (the true effect is zero). But they both cause school disciplinary referral.
sim_c <- data.frame(
  games      = rnorm(n_sim, 0, 1),
  aggression = rnorm(n_sim, 0, 1)
)
# Probability of referral depends on both games and aggression.
sim_c$referral <- as.numeric(
  sim_c$games + sim_c$aggression > 1
)

# Estimate 1: in the full population.
est_full      <- coef(lm(aggression ~ games, data = sim_c))["games"]
# Estimate 2: ONLY among referred students (= "controlling for" being referred).
est_referred  <- coef(lm(aggression ~ games,
                          data = sim_c[sim_c$referral == 1, ]))["games"]
# Estimate 3: controlling for referral in regression.
est_controlled <- coef(lm(aggression ~ games + referral,
                           data = sim_c))["games"]

cat(sprintf("True causal effect (games on aggression)             = 0.00\n"))
cat(sprintf("Estimate in full population (correct)                = %.2f\n",
            est_full))
cat(sprintf("Estimate among referred students only                = %.2f\n",
            est_referred))
cat(sprintf("Estimate controlling for referral in regression      = %.2f\n",
            est_controlled))
```

In the full population, games and aggression are uncorrelated; the regression returns a coefficient near zero, as it should. But the moment we restrict to referred students, or equivalently include referral as a control variable, a *negative* coefficient appears out of nowhere. The two variables, which were causally independent, now look correlated *because* we conditioned on a variable they both cause.

The intuition is this. Imagine you only look at students who were referred to the counselor. Among referred students, if someone plays very few violent games, they were probably referred because they are aggressive (something had to explain the referral). If someone is not very aggressive, they were probably referred because they play a lot of games. So within the referred group, gaming and aggression become *negatively* associated, even though they are completely independent in the full population. By conditioning on the common effect, you have manufactured a relationship out of nothing.

The figure below visualizes this. We plot games and aggression on two axes for the full population (left) and for only the referred sub-population (right). In the full population, the cloud of points has no tilt. In the referred sub-population, a clear tilt appears, even though the underlying causal structure has not changed.

```{r}
#| label: fig-collider-bias
#| echo: false
#| fig-width: 9
#| fig-height: 3.8
#| fig-align: center

set.seed(33)
n_sim_show <- 800
big_sim <- data.frame(
  games      = rnorm(n_sim_show, 0, 1),
  aggression = rnorm(n_sim_show, 0, 1)
)
big_sim$referral <- as.numeric(big_sim$games + big_sim$aggression > 1)
big_sim$referral_label <- ifelse(big_sim$referral == 1,
                                  "Referred", "Not referred")

p_full <- ggplot(big_sim, aes(x = games, y = aggression)) +
  geom_point(alpha = 0.35, color = "grey50", size = 1.0) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              color = "steelblue4", linewidth = 0.8) +
  labs(x = "Games", y = "Aggression",
       title = "Full population\n(games and aggression independent)") +
  coord_cartesian(xlim = c(-3.5, 3.5), ylim = c(-3.5, 3.5)) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 10),
        panel.grid.minor = element_blank())

p_ref <- ggplot(big_sim[big_sim$referral == 1, ],
                aes(x = games, y = aggression)) +
  geom_point(alpha = 0.55, color = "firebrick", size = 1.0) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              color = "firebrick4", linewidth = 0.8) +
  labs(x = "Games", y = "Aggression",
       title = "Restricted to referred students\n(spurious negative association)") +
  coord_cartesian(xlim = c(-3.5, 3.5), ylim = c(-3.5, 3.5)) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 10),
        panel.grid.minor = element_blank())

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_full, p_ref, ncol = 2)
} else {
  gridExtra::grid.arrange(p_full, p_ref, ncol = 2)
}
```

The left panel shows the full population: no relationship between games and aggression, just a circular cloud. The right panel restricts to the referred sub-population: a clear negative tilt has appeared. The tilt is not a real causal effect. It is collider bias, manufactured by conditioning on the common effect.

A practical generalization. "Controlling for a collider in regression" and "studying only a selected sample" are the same operation in DAG terms. Both condition on the collider; both open the spurious path. If you study only hospital patients (a sample selected on a variable that is plausibly a collider for many of the questions hospitals are used to study), you can get collider bias without ever putting the variable into a regression. This is why sampling matters for causal claims, not just statistical analysis: the sample you analyze is implicitly a conditional selection, and if you select on a collider, you introduce bias.

## The mirror image

Confounders and colliders look strikingly similar in a diagram: both involve a third variable connected to *X* and *Y*. But the arrow directions are opposite. For a confounder, the arrows go out of the third variable toward *X* and *Y*. For a collider, the arrows come into the third variable from *X* and *Y*.

```{r}
#| label: fig-mirror
#| echo: false
#| fig-width: 8
#| fig-height: 3.4
#| fig-align: center

nodes_conf2 <- data.frame(
  name = c("X", "Y", "C"),
  label = c("X", "Y", "C\n(confounder)"),
  x = c(-0.5, 1.0, 0.25), y = c(0, 0, 1.2)
)
edges_conf2 <- data.frame(
  from = c("C", "C"),
  to   = c("X", "Y")
)
p_left <- draw_dag(
  nodes_conf2, edges_conf2,
  title = "Confounder (X <- C -> Y):\npath open until you control C",
  xlim = c(-1.0, 1.6), ylim = c(-0.4, 1.7))

nodes_col2 <- data.frame(
  name = c("X", "Y", "C"),
  label = c("X", "Y", "C\n(collider)"),
  x = c(-0.5, 1.0, 0.25), y = c(0, 0, -1.2)
)
edges_col2 <- data.frame(
  from = c("X", "Y"),
  to   = c("C", "C")
)
p_right <- draw_dag(
  nodes_col2, edges_col2,
  title = "Collider (X -> C <- Y):\npath closed until you control C",
  xlim = c(-1.0, 1.6), ylim = c(-1.7, 0.4))

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_left, p_right, ncol = 2)
} else {
  gridExtra::grid.arrange(p_left, p_right, ncol = 2)
}
```

The two figures look like mirror images, and the actions they require are mirror images too:

- **Confounder.** Path is *open* by default. Controlling for C *closes* it. Therefore, control for confounders.
- **Collider.** Path is *closed* by default. Controlling for C *opens* it. Therefore, do not control for colliders.

The arrow directions determine which is which. And here is the punchline: *the data alone cannot tell us which way the arrows point.* The observed correlation between *X* and the third variable, or between the third variable and *Y*, is the same whether C causes *X* and *Y* or *X* and *Y* cause C. Only causal knowledge from outside the data can tell us the direction of the arrows. This is the precise sense in which Wysocki, Lawson, and Rhemtulla ([2022](https://doi.org/10.1177/25152459221095823)) say *statistical control requires causal justification*. Without the DAG, you cannot tell whether a particular control is fixing a problem or creating one.

## Two more cases worth knowing

The three classic cases above are the most important. But there are two additional patterns Rohrer (2018) and Wysocki, Lawson, and Rhemtulla (2022) note that are worth a brief mention, because they recover and connect to ideas from Sections 3 and 4.

**Causes of *Y* only (precision covariates).** Suppose a third variable *Z* affects the outcome but is unrelated to the treatment. For example, in a randomized experiment, a participant's baseline mood (before the experiment started) might affect their aggression scores at the end (Y) but, because the game (X) was randomly assigned, baseline mood is unrelated to which game they played.

```{r}
#| label: fig-precision
#| echo: false
#| fig-width: 5
#| fig-height: 1.8
#| fig-align: center

nodes_p <- data.frame(
  name = c("X", "Y", "Z"),
  label = c("X\n(games)", "Y\n(aggression)", "Z\n(baseline mood)"),
  x = c(-1.4, 0.4, 2.2), y = c(0, 0, 0)
)
edges_p <- data.frame(
  from = c("X", "Z"),
  to   = c("Y", "Y")
)
draw_dag(nodes_p, edges_p,
         title = "Cause of Y only: optional precision covariate",
         node_pad = 0.55,
         xlim = c(-2.0, 2.8), ylim = c(-0.4, 0.4))
```

This is not a confounder (because *Z* has no arrow to *X*), so leaving it out causes no bias. But controlling for *Z* "soaks up" some of the noise in *Y*, which shrinks the residual variance and makes your estimate of the X effect more precise. This is exactly the ANCOVA-style noise reduction we noted in Chapter 17 in connection with multiple regression. Adding such a covariate in a randomized experiment is harmless and often beneficial for precision. (In observational data, the picture is more complicated, but the principle that pre-treatment covariates can help precision still holds.)

**Causes of *X* only.** Suppose a third variable *Z* causes *X* but has no direct effect on *Y*.

```{r}
#| label: fig-instrument-style
#| echo: false
#| fig-width: 5
#| fig-height: 1.8
#| fig-align: center

nodes_ix <- data.frame(
  name = c("X", "Y", "Z"),
  label = c("X", "Y", "Z"),
  x = c(0, 1.2, -1.2), y = c(0, 0, 0)
)
edges_ix <- data.frame(
  from = c("Z", "X"),
  to   = c("X", "Y")
)
draw_dag(nodes_ix, edges_ix,
         title = "Cause of X only: usually not a control",
         node_pad = 0.22,
         xlim = c(-1.8, 1.8), ylim = c(-0.4, 0.4))
```

This is not a confounder (it has no path to *Y* except through *X*), so controlling it does not remove any bias. Controlling for it can actually hurt: it can reduce the variation in *X* available for the analysis (which lowers precision), and in some situations it can amplify any leftover bias from unmeasured confounders elsewhere ("bias amplification"). The simple rule: a variable that only causes *X* and does not affect *Y* is not a control variable you want to include.

## Pulling it together

The DAG-based rule for control variable selection now comes into focus:

| Type | Structure | What controlling does | Should you control? |
|:-----|:----------|:----------------------|:-------------------|
| Confounder | *X* ← C → *Y* | Closes a backdoor path | YES |
| Mediator | *X* → M → *Y* | Closes part of the causal path | NO (for the total effect) |
| Collider | *X* → C ← *Y* | Opens a non-causal path | NO |
| Cause of *Y* only | *X* → *Y* ← Z | Soaks up noise in *Y* | OPTIONAL (precision) |
| Cause of *X* only | Z → *X* → *Y* | Reduces variation in *X* | NO (usually) |

The big picture: every choice of which variables to put in a regression is a causal choice. The choice is determined by the role each candidate variable plays in the DAG, not by the data alone, not by "more is better," and not by p-values. The next chapter ties this into a single procedural rule (the **backdoor criterion**) and explains the practical consequences for how to read multiple regression output.

## Check your understanding

```{r}
#| echo: false
opts_q1 <- c(
  "A confounder is a variable that should not be controlled for in a regression, because including it changes the meaning of the focal coefficient and makes the estimated causal effect ambiguous.",
  answer = "A confounder is a variable that causally affects both the predictor and the outcome (the fork structure X <- C -> Y); controlling for it closes the backdoor path it opens and is required for an unbiased causal estimate.",
  "A confounder is a variable that is statistically associated with the predictor but has no causal effect on either the predictor or the outcome; controlling for it is harmless but unnecessary.",
  "A confounder is a variable whose effect on the outcome is itself confounded by other variables, requiring a chain of additional adjustments to be made before the focal causal effect can be estimated."
)

opts_q2 <- c(
  "A mediator is a variable that, like a confounder, sits between the predictor and the outcome and should always be controlled for in the regression to obtain an unbiased estimate of the focal causal effect.",
  answer = "A mediator is a variable that sits ON the causal path from the predictor to the outcome (the chain structure X -> M -> Y); it is part of the mechanism by which the cause produces the effect, and controlling for it removes part of the total causal effect, which is usually not what is wanted.",
  "A mediator is a variable that is causally affected by the outcome but has no effect on the predictor; controlling for it produces a biased estimate of the focal causal effect.",
  "A mediator is a variable that has been hypothesized in the literature but for which no direct empirical evidence is available; it is included in the regression as a placeholder to be tested in future work."
)

opts_q3 <- c(
  "A collider is a variable that, like a confounder, causally affects both the predictor and the outcome; the rule for whether to control it depends on the direction of the arrows along the causal path.",
  answer = "A collider is a variable that is causally affected by both the predictor and the outcome (the inverted fork structure X -> C <- Y); the path through a collider is closed by default, but controlling for the collider (or selecting the sample on it) opens the path and creates a spurious association between X and Y that did not exist in the population.",
  "A collider is a variable that has no causal effect on either the predictor or the outcome but is statistically correlated with both; controlling for it has no effect on the estimated causal effect.",
  "A collider is a variable that has been measured but is missing for some participants in the sample; the rule for whether to control it depends on the proportion of missing values in the dataset."
)

opts_q4 <- c(
  "The data alone can tell us whether a third variable is a confounder, mediator, or collider, by examining the partial correlations of the variable with the predictor and the outcome.",
  answer = "The data alone cannot tell us whether a third variable is a confounder, mediator, or collider, because the observed associations are the same regardless of which way the causal arrows point; only causal knowledge from outside the data (theory, prior research, substantive understanding) can settle the question.",
  "The data alone can tell us whether a third variable is a confounder, mediator, or collider, by checking the temporal ordering of the variables in the dataset and applying the rule that earlier variables are causes of later ones.",
  "The data alone can tell us whether a third variable is a confounder, mediator, or collider, by computing the partial regression coefficient and applying the rule that a positive coefficient indicates a confounder, a negative one a collider, and a near-zero one a mediator."
)

opts_q5 <- c(
  "Confounders and colliders are mirror images of each other and should always be controlled together in the same regression, because the action required for one cancels out the action required for the other.",
  answer = "Confounders and colliders are mirror images of each other in a DAG (a confounder has arrows pointing out of the third variable, a collider has arrows pointing in), and the correct action is opposite: a confounder should be controlled to close its open backdoor path, while a collider should be left alone because controlling it opens a path that was closed.",
  "Confounders and colliders are equivalent under the assumptions of the linear model, and the choice between calling a variable one or the other is a matter of statistical convention rather than substantive meaning.",
  "Confounders and colliders are the same kind of variable but with different statistical signatures: confounders tend to be measured before the predictor and the outcome, while colliders tend to be measured after."
)

opts_q6 <- c(
  "Selecting the sample of participants based on a variable that is a collider for the focal causal question has no effect on the estimated causal effect, because the regression uses only the participants who are in the sample and so the selection is irrelevant.",
  answer = "Selecting the sample of participants based on a variable that is a collider for the focal causal question is equivalent in DAG terms to controlling for the collider in the regression, and produces the same kind of spurious association: the focal causal effect can appear distorted within the selected sample even when it is correctly zero or differently signed in the full population.",
  "Selecting the sample of participants based on a variable that is a collider for the focal causal question is acceptable as long as the selection is performed before the data are collected, because pre-data selection is part of the study design rather than the analysis.",
  "Selecting the sample of participants based on a variable that is a collider for the focal causal question is acceptable as long as the researcher reports the selection criteria in the methods section of the paper, because transparency makes any bias visible to the reader."
)
```

**1. What is a confounder, and how should it be handled in a regression intended to estimate a causal effect?**

`r longmcq(opts_q1)`

**2. What is a mediator, and how should it be handled in a regression intended to estimate the total causal effect of X on Y?**

`r longmcq(opts_q2)`

**3. What is a collider, and how should it be handled in a regression intended to estimate a causal effect?**

`r longmcq(opts_q3)`

**4. Can the data alone tell us whether a third variable is a confounder, a mediator, or a collider?**

`r longmcq(opts_q4)`

**5. Why are confounders and colliders described as "mirror images" of each other?**

`r longmcq(opts_q5)`

**6. What is the relationship between "controlling for a collider in a regression" and "selecting the sample on a collider"?**

`r longmcq(opts_q6)`

## References

Elwert, F., & Winship, C. (2014). Endogenous selection bias: The problem of conditioning on a collider variable. *Annual Review of Sociology, 40*, 31-53. <https://doi.org/10.1146/annurev-soc-071913-043455>

Rohrer, J. M. (2018). Thinking clearly about correlations and causation: Graphical causal models for observational data. *Advances in Methods and Practices in Psychological Science, 1*(1), 27-42. <https://doi.org/10.1177/2515245917745629>

Rohrer, J. M., Hünermund, P., Arslan, R. C., & Elson, M. (2022). That's a lot to process! Pitfalls of popular path models. *Advances in Methods and Practices in Psychological Science, 5*(2). <https://doi.org/10.1177/25152459221095827>

Wysocki, A. C., Lawson, K. M., & Rhemtulla, M. (2022). Statistical control requires causal justification. *Advances in Methods and Practices in Psychological Science, 5*(2). <https://doi.org/10.1177/25152459221095823>
